---
title: "010_pr_labor_historical"
author: "Jorge R. Soldevila Irizarry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 999)
selected_years <- c(2005, 2010, 2015, 2020)

# Source central setup file
source("00_project_setup.R")
```

# Puerto Ricans in the US Labor Force 1950-2010

```{r, Calculate estimates for Puerto Ricans and Total US Pop, message=FALSE, warning=FALSE}
pr_labfor_50_10 <- calculate_lfpr(data_hist, HISPAN == 2 |
  HISPAND == 200 |
  ANCESTR1 == 261)

us_labfor_50_10 <- calculate_lfpr(data_hist)
```

## We can visualize how the Puerto Rican population that can work, that is, non-institutionalized, 16 years and over population, has changed over time.

```{r Visualize Total Puerto Ricans working age, message=FALSE,warning=FALSE}
# We can now visualize the working age Puerto Rican population since 1950
ggplot() +
  geom_line(
    data = pr_labfor_50_10,
    aes(x = YEAR, y = total_pop, color = "PR in US"),
    size = 1
  ) +
  geom_point(
    data = pr_labfor_50_10,
    aes(x = YEAR, y = total_pop, color = "PR in US"),
    size = 2
  ) +
  geom_text(
    data = pr_labfor_50_10,
    aes(
      x = YEAR,
      y = total_pop,
      label = label_number(
        scale = 1e-6,
        suffix = "M",
        accuracy = 0.1
      )(total_pop)
    ),
    vjust = -1.5,
    show.legend = FALSE
  ) +
  # important_years_lines +
  my_theme +
  labs(title = "Working age non-institutionalized Puerto Ricans, 1950-2024", x = "Year", y = "Estimated Population") +
  scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

## We can visualize now the population that was in the labor force

```{r Visualize Total Puerto Ricans in labor force, message=FALSE,warning=FALSE}
# We can now visualize the Puerto Rican population in the labor force since 1950.

ggplot() +
  geom_line(
    data = pr_labfor_50_10,
    aes(x = YEAR, y = in_labfor, color = "PR in US"),
    size = 1
  ) +
  geom_point(
    data = pr_labfor_50_10,
    aes(x = YEAR, y = in_labfor, color = "PR in US"),
    size = 2
  ) +
  geom_text(
    data = pr_labfor_50_10,
    aes(
      x = YEAR,
      y = in_labfor,
      label = label_number(
        scale = 1e-6,
        suffix = "M",
        accuracy = 0.1
      )(in_labfor)
    ),
    vjust = -1.5,
    show.legend = FALSE
  ) +
  my_theme +
  labs(title = "Puerto Ricans in the Labor Force, 1950-2024", x = "Year", y = "Estimated Population") +
  scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

## We can now visualize the Labor Force Participation Rate for Puerto Ricans

```{r Visualize Total Puerto Ricans LFPR, message=FALSE,warning=FALSE}
# We can now visualize the Puerto Rican Labor FOrce Participation Rate since 1950.

ggplot() +
  geom_line(
    data = pr_labfor_50_10,
    aes(x = YEAR, y = labfor_rate, color = "PR"),
    size = 1
  ) +
  geom_point(
    data = pr_labfor_50_10,
    aes(x = YEAR, y = labfor_rate, color = "PR"),
    size = 2
  ) +
  geom_line(
    data = us_labfor_50_10,
    aes(x = YEAR, y = labfor_rate, color = "US"),
    size = 1
  ) +
  geom_point(
    data = us_labfor_50_10,
    aes(x = YEAR, y = labfor_rate, color = "US"),
    size = 2
  ) +
  my_theme +
  scale_color_manual(values = c("PR" = "orange", "US" = "blue")) +
  labs(title = "Labor Force Participaiton Rate for Puerto Ricans and Total Population in US, 1950-2024", x = "Year", y = "Rate") +
  scale_x_continuous(breaks = seq(1950, 2024, by = 10))
```

## Unemployment Rate for Puerto Ricans and Total Population

```{r Unemployment Rate Visualization, message=FALSE, warning=FALSE}
ggplot() +
  geom_line(
    data = pr_labfor_50_10,
    aes(x = YEAR, y = unemp_rate, color = "PR"),
    size = 1
  ) +
  geom_point(
    data = pr_labfor_50_10,
    aes(x = YEAR, y = unemp_rate, color = "PR"),
    size = 2
  ) +
  geom_line(
    data = us_labfor_50_10,
    aes(x = YEAR, y = unemp_rate, color = "US"),
    size = 1
  ) +
  geom_point(
    data = us_labfor_50_10,
    aes(x = YEAR, y = unemp_rate, color = "US"),
    size = 2
  ) +
  my_theme +
  scale_color_manual(values = c("PR" = "orange", "US" = "blue")) +
  labs(title = "Unemployment Rate for Puerto Ricans and Total Population in US, 1950-2024", x = "Year", y = "Rate") +
  scale_x_continuous(breaks = seq(1950, 2024, by = 10))
```

## Population Not in Labor Force for Puerto Ricans and Total Population

```{r Not in Labor Force Visualization, message=FALSE, warning=FALSE}
ggplot() +
  geom_line(data = pr_labfor_50_10, aes(
    x = YEAR, y = out_labfor_pct,
    color = "PR"
  ), size = 1) +
  geom_point(data = pr_labfor_50_10, aes(
    x = YEAR, y = out_labfor_pct,
    color = "PR"
  ), size = 2) +
  geom_line(data = us_labfor_50_10, aes(
    x = YEAR, y = out_labfor_pct,
    color = "US"
  ), size = 1) +
  geom_point(data = us_labfor_50_10, aes(
    x = YEAR, y = out_labfor_pct,
    color = "US"
  ), size = 2) +
  my_theme +
  scale_color_manual(values = c(
    "PR" = "orange",
    "US" = "blue"
  )) +
  labs(
    title = "Percent of Populaiton Out of Workforce for Puerto Ricans and Total Population in US, 1950-2024",
    x = "Year",
    y = "Rate"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

## Not surprising, LFPR varies between male and females.

```{r Calculations for LFPR for PR and US pop by sex, message=FALSE,warning=FALSE}
pr_labfor_sex <- data_hist %>%
  filter(
    HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261,
    AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3,
    SEX != 9
  ) %>%
  group_by(YEAR, SEX) %>%
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    sex = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female"
    ),
    labfor_rate = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor - unemp) / total_pop * 100
  ) %>%
  collect()

us_labfor_sex <- data_hist %>%
  filter(
    AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3,
    SEX != 9
  ) %>%
  group_by(YEAR, SEX) %>%
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    sex = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female"
    ),
    labfor_rate = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor - unemp) / total_pop * 100
  ) %>%
  collect()
```

## Visualize trends over years by sex

### Distribution by sex

```{r Distribution by sex, message=FALSE, warning=FALSE}
# Distribution of Puerto Ricans by sex
pr_labfor_sex <- pr_labfor_sex |>
  group_by(YEAR) |>
  mutate(pct = (total_pop / sum(total_pop) * 100)) %>%
  ungroup()

ggplot(pr_labfor_sex, aes(x = YEAR, y = pct, fill = sex)) +
  geom_col(position = "dodge") +
  labs(
    title = "PR Population Percentage by Sex Over Time",
    x = "Year",
    y = "Percentage (%)",
    fill = "Sex"
  ) +
  my_theme +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))

# Distribution of US Population by sex
us_labfor_sex <- us_labfor_sex |>
  group_by(YEAR) |>
  mutate(pct = (total_pop / sum(total_pop) * 100)) %>%
  ungroup()

ggplot(us_labfor_sex, aes(x = YEAR, y = pct, fill = sex)) +
  geom_col(position = "dodge") +
  labs(
    title = "US Population Percentage by Sex Over Time",
    x = "Year",
    y = "Percentage (%)",
    fill = "Sex"
  ) +
  my_theme +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

### Labor Force Participation Rate by Sex

```{r LFPR by sex, message=FALSE, warning=FALSE}
# We can now visualize the Puerto Rican population in the labor force by sex since 1950.
ggplot() +
  geom_line(data = pr_labfor_sex, aes(
    x = YEAR, y = labfor_rate,
    color = sex
  )) +
  geom_point(data = pr_labfor_sex, aes(
    x = YEAR, y = labfor_rate,
    color = sex
  )) +
  my_theme +
  labs(
    title = "Trends in Labor Force Participation for Puerto Ricans by Sex 1950-2010",
    x = "Year",
    y = "Estimated Population"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))

# We can now visualize the total US population in the labor force by sex since 1950.
ggplot() +
  geom_line(data = us_labfor_sex, aes(
    x = YEAR, y = labfor_rate,
    color = sex
  )) +
  geom_point(data = us_labfor_sex, aes(
    x = YEAR, y = labfor_rate,
    color = sex
  )) +
  my_theme +
  labs(
    title = "Trends in Labor Force Participation for Total Population by Sex 1950-2024",
    x = "Year",
    y = "Estimated Population"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

## We can also analyze change by place of birth

```{r Puerto Ricans in Labor Force by place of birth, message=FALSE,warning=FALSE}
pr_labfor_bpl <- data_hist %>%
  filter(
    HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261,
    AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3,
  ) %>%
  mutate(
    plc_bth = case_when(
      BPL <= 56 ~ "US",
      BPL == 110 ~ "Puerto Rico",
      TRUE ~ NA
    )
  ) %>%
  filter(!is.na(plc_bth)) %>%
  group_by(YEAR, plc_bth) %>%
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    labfor_rate = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor - unemp) / total_pop * 100
  ) %>%
  collect()
```

# We can now visualize the Puerto Rican population in the labor force by Place of Birth.

```{r LFPR for Puerto Ricans by Place of Birth, message=FALSE,warning=FALSE}
ggplot() +
  geom_line(data = pr_labfor_bpl, aes(
    x = YEAR, y = labfor_rate,
    color = plc_bth
  )) +
  geom_point(data = pr_labfor_bpl, aes(
    x = YEAR, y = labfor_rate,
    color = plc_bth
  )) +
  my_theme +
  labs(
    title = "Trends in Labor Force Participation by Place of Birth 1950-2024",
    x = "Year",
    y = "Estimated Population"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

## LFPR by Age Groups

```{r Age Goup Table, message=FALSE, warning=FALSE}
# Create a base table with the virtual grouping
acs_with_age <- data_hist %>%
  mutate(
    age_grp = case_when(
      AGE < 16 ~ "minor",
      AGE >= 16 & AGE <= 24 ~ "16 to 24",
      AGE >= 25 & AGE <= 54 ~ "25 to 54",
      AGE >= 55 ~ "55+"
    )
  )
```

```{r LFPR Age Group calculation, message=FALSE,warning=FALSE}
# Calculate for PR in US by Age Group
pr_age_lf_rate_us <- acs_with_age %>%
  filter(
    HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261, AGE >= 16,
    EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3
  ) %>%
  group_by(YEAR, age_grp) %>%
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    labfor_rate = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor - unemp) / total_pop * 100
  ) %>%
  collect()

age_lf_rate_us <- acs_with_age %>%
  filter(AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3) %>%
  group_by(YEAR, age_grp) %>%
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    labfor_rate = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor - unemp) / total_pop * 100
  ) %>%
  collect()
```

## LFPR by Age Visualization

#### Puerto Ricans

```{r LFPR Age Group Visualization, message=FALSE,warning=FALSE}
ggplot(pr_age_lf_rate_us, aes(x = YEAR, y = labfor_rate, color = age_grp)) +
  geom_line(size = 1) +
  geom_point() +
  # important_years_lines +
  my_theme +
  labs(
    title = "Labor Force Participation Rate for Puerto Ricans in the US by Age Group",
    x = "Year",
    y = "Participation Rate (%)"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

#### Total US Population

```{r LFPR Age Group Visualization message=FALSE,warning=FALSE}
ggplot(age_lf_rate_us, aes(x = YEAR, y = labfor_rate, color = age_grp)) +
  geom_line(size = 1) +
  geom_point() +
  # important_years_lines +
  my_theme +
  labs(
    title = "Labor Force Participation Rate for US Population by Age Group",
    x = "Year",
    y = "Participation Rate (%)"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

## LFPR by Age Updated

```{r Age Goup Table updated, message=FALSE, warning=FALSE}
# Create a base table with the virtual grouping
acs_with_age_updated <- data_hist %>%
  mutate(
    age_grp = case_when(
      AGE >= 16 & AGE <= 21 ~ "16-21",
      AGE >= 22 & AGE <= 27 ~ "22-27",
      AGE >= 28 & AGE <= 34 ~ "28-34",
      AGE >= 35 & AGE <= 44 ~ "35-44",
      AGE >= 45 & AGE <= 54 ~ "45-54",
      AGE >= 55 & AGE <= 64 ~ "55-64",
      AGE >= 65 ~ "65+"
    )
  )
```

```{r LFPR Age Group calculation, message=FALSE,warning=FALSE}
# Calculate for PR in US by Age Group
pr_age_lf_rate_updated <- acs_with_age_updated %>%
  filter(
    HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261, AGE >= 16,
    EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3
  ) %>%
  group_by(YEAR, age_grp) %>%
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    labfor_rate = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor - unemp) / total_pop * 100
  ) %>%
  collect()

age_lf_rate_updated <- acs_with_age_updated %>%
  filter(AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3) %>%
  group_by(YEAR, age_grp) %>%
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    labfor_rate = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor - unemp) / total_pop * 100
  ) %>%
  collect()
```

## LFPR by Age Visualization Updated

#### Puerto Ricans Updated

```{r LFPR Age Group Visualization, message=FALSE,warning=FALSE}
# 1. Define your highlight colors and sizes
# Give "22-27" a bright color (like red or blue) and the rest a muted grey
my_colors <- c(
  "16-21" = "#a6cee3",
  "22-27" = "#E3120B", # Bright Red for highlight
  "28-34" = "#1f78b4",
  "35-44" = "#b2df8a",
  "45-54" = "#33a02c",
  "55-64" = "#fb9a99",
  "65+"   = "orange"
)

# Make the "22-27" line thicker than the rest
my_lines <- c(
  "16-21" = 0.75,
  "22-27" = 1.5, # Thicker line for highlight
  "28-34" = 0.75,
  "35-44" = 0.75,
  "45-54" = 0.75,
  "55-64" = 0.75,
  "65+"   = 0.75
)

# 2. Update your ggplot code
# Note: we added `linewidth = age_grp` inside the aes() function
ggplot(pr_age_lf_rate_updated, aes(x = YEAR, y = labfor_rate, color = age_grp, linewidth = age_grp)) +
  geom_line() +
  geom_point(aes(size = age_grp), show.legend = FALSE) + # Scale points to match
  scale_color_manual(values = my_colors) +
  scale_linewidth_manual(values = my_lines) +
  scale_size_manual(values = my_lines * 2) + # Makes points slightly larger than lines
  # important_years_lines +
  my_theme +
  labs(
    title = "Labor Force Participation Rate for Puerto Ricans in the US by Age Group",
    x = "Year",
    y = "Participation Rate (%)",
    color = "Age Group", # Cleans up legend titles
    linewidth = "Age Group" # Ensures color and line width share the same legend
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

#### Total US Population Updated

```{r LFPR Age Group Visualization message=FALSE,warning=FALSE}
ggplot(age_lf_rate_updated, aes(x = YEAR, y = labfor_rate, color = age_grp, linewidth = age_grp)) +
  geom_line() +
  geom_point(aes(size = age_grp), show.legend = FALSE) + # Scale points to match
  scale_color_manual(values = my_colors) +
  scale_linewidth_manual(values = my_lines) +
  scale_size_manual(values = my_lines * 2) + # Makes points slightly larger than lines
  # important_years_lines +
  my_theme +
  labs(
    title = "Labor Force Participation Rate for Total Population in the US by Age Group",
    x = "Year",
    y = "Participation Rate (%)",
    color = "Age Group", # Cleans up legend titles
    linewidth = "Age Group" # Ensures color and line width share the same legend
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

## Visualize percent of population out of the labor force by age

```{r Percent Out of the Labor Force for Puerto Ricans, warning=FALSE,message=FALSE}
# 2. Update your ggplot code
# Note: we added `linewidth = age_grp` inside the aes() function
ggplot(pr_age_lf_rate_updated, aes(x = YEAR, y = out_labfor_pct, color = age_grp, linewidth = age_grp)) +
  geom_line() +
  geom_point(aes(size = age_grp), show.legend = FALSE) + # Scale points to match
  scale_color_manual(values = my_colors) +
  scale_linewidth_manual(values = my_lines) +
  scale_size_manual(values = my_lines * 2) + # Makes points slightly larger than lines
  # important_years_lines +
  my_theme +
  labs(
    title = "Percent of Puerto Ricans Out of the Labor Force in the US by Age Group",
    x = "Year",
    y = "Participation Rate (%)",
    color = "Age Group", # Cleans up legend titles
    linewidth = "Age Group" # Ensures color and line width share the same legend
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

```{r Percent Out of the Labor Force for total pop, warning=FALSE,message=FALSE}
# 2. Update your ggplot code
# Note: we added `linewidth = age_grp` inside the aes() function
ggplot(age_lf_rate_updated, aes(x = YEAR, y = out_labfor_pct, color = age_grp, linewidth = age_grp)) +
  geom_line() +
  geom_point(aes(size = age_grp), show.legend = FALSE) + # Scale points to match
  scale_color_manual(values = my_colors) +
  scale_linewidth_manual(values = my_lines) +
  scale_size_manual(values = my_lines * 2) + # Makes points slightly larger than lines
  # important_years_lines +
  my_theme +
  labs(
    title = "Percent of PPopulation Out of the Labor Force in the US by Age Group",
    x = "Year",
    y = "Participation Rate (%)",
    color = "Age Group", # Cleans up legend titles
    linewidth = "Age Group" # Ensures color and line width share the same legend
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

## Unemployment Rate for Puerto Ricans by Age Group

```{r Unemployment Rate for Puerto Ricans, warning=FALSE,message=FALSE}
ggplot(pr_age_lf_rate_updated, aes(x = YEAR, y = unemp_rate, color = age_grp, linewidth = age_grp)) +
  geom_line() +
  geom_point(aes(size = age_grp), show.legend = FALSE) + # Scale points to match
  scale_color_manual(values = my_colors) +
  scale_linewidth_manual(values = my_lines) +
  scale_size_manual(values = my_lines * 2) + # Makes points slightly larger than lines
  # important_years_lines +
  my_theme +
  labs(
    title = "Unemployment Rate for Puerto Ricans in the US by Age Group",
    x = "Year",
    y = "Participation Rate (%)",
    color = "Age Group", # Cleans up legend titles
    linewidth = "Age Group" # Ensures color and line width share the same legend
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

```{r Unemployment Rate for total pop, warning=FALSE,message=FALSE}
ggplot(age_lf_rate_updated, aes(x = YEAR, y = unemp_rate, color = age_grp, linewidth = age_grp)) +
  geom_line() +
  geom_point(aes(size = age_grp), show.legend = FALSE) + # Scale points to match
  scale_color_manual(values = my_colors) +
  scale_linewidth_manual(values = my_lines) +
  scale_size_manual(values = my_lines * 2) + # Makes points slightly larger than lines
  # important_years_lines +
  my_theme +
  labs(
    title = "Unemployment Rate for Total Population in the US by Age Group",
    x = "Year",
    y = "Participation Rate (%)",
    color = "Age Group", # Cleans up legend titles
    linewidth = "Age Group" # Ensures color and line width share the same legend
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

## Education Attainment for Puerto Ricans

```{r Education Attainment Group, message=FALSE,warning=FALSE}
# Create a base table with the virtual grouping
acs_with_educ <- data_hist %>%
  mutate(
    educ_grp = case_when(
      EDUC <= 5 ~ "Less than High School",
      EDUC == 6 ~ "High School Diploma/GED",
      EDUC %in% c(7, 8, 9) ~ "Some College / Associate's",
      EDUC == 10 ~ "Bachelor's Degree",
      EDUC >= 11 ~ "Master's Degree or Higher",
      TRUE ~ "Unknown/NA"
    )
  ) %>%
  filter(educ_grp != "Unknown/NA")
```

```{r Calculate Education Attainment, message=FALSE,warning=FALSE}
# Calculate for PR in US by Education Group
pr_educ_lf_rate_us <- acs_with_educ %>%
  filter(
    HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261,
    AGE >= 25,
    EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3
  ) %>%
  group_by(YEAR, educ_grp) %>%
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  mutate(
    labfor_rate = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor - unemp) / total_pop * 100,
    educ_attainment_pct = (total_pop / sum(total_pop)) * 100
  ) %>%
  ungroup() %>%
  collect()

# Calculate for Total US Population by Education Group
# Applied AGE >= 25 constraint here as well
educ_lf_rate_us <- acs_with_educ %>%
  filter(AGE >= 25, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3, ) %>%
  group_by(YEAR, educ_grp) %>%
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  mutate(
    labfor_rate = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor - unemp) / total_pop * 100,
    educ_attainment_pct = (total_pop / sum(total_pop)) * 100
  ) %>%
  ungroup() %>%
  collect()
```

### Visualize Educational Attainment per year

```{r PR Educational Attainment Viz, message=FALSE, warning=FALSE}
ggplot(pr_educ_lf_rate_us, aes(x = YEAR, y = educ_attainment_pct, color = educ_grp)) +
  geom_line(linewidth = 1) +
  geom_point() +
  # important_years_lines +
  my_theme +
  labs(
    title = "Educational Attainment for Puerto Ricans in the US",
    x = "Year",
    y = "Percentage of Population (%)"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

```{r Total Pop Educaitonal Attainment Viz, message=FALSE, warning=FALSE}
ggplot(educ_lf_rate_us, aes(x = YEAR, y = educ_attainment_pct, color = educ_grp)) +
  geom_line(linewidth = 1) +
  geom_point() +
  # important_years_lines +
  my_theme +
  labs(
    title = "Educational Attainment for Total Population in the US",
    x = "Year",
    y = "Percentage of Population (%)"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

### Puerto Ricans

```{r LFPR by Educaitonal Attainment PR, message=FALSE, warning=FALSE}
ggplot(pr_educ_lf_rate_us, aes(x = YEAR, y = labfor_rate, color = educ_grp)) +
  geom_line(linewidth = 1) + # Updated size to linewidth
  geom_point() +
  # important_years_lines +
  my_theme +
  labs(
    title = "Labor Force Participation Rate for Puerto Ricans in the US by Education",
    x = "Year",
    y = "Participation Rate (%)"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

### Total US Population

```{r LFPR by Educaitonal Attainment Total Pop, message=FALSE, warning=FALSE}
ggplot(educ_lf_rate_us, aes(x = YEAR, y = labfor_rate, color = educ_grp)) +
  geom_line(linewidth = 1) + # Updated size to linewidth
  geom_point() +
  # important_years_lines +
  my_theme +
  labs(
    title = "Labor Force Participation Rate for US Population by Education",
    x = "Year",
    y = "Participation Rate (%)"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

### Employment Populaiton Ration by Educational Attainment

```{r EPR by Educaitonal Attainment PR, message=FALSE, warning=FALSE}
ggplot(pr_educ_lf_rate_us, aes(x = YEAR, y = etop_ratio, color = educ_grp)) +
  geom_line(linewidth = 1) + # Updated size to linewidth
  geom_point() +
  # important_years_lines +
  my_theme +
  labs(
    title = "Employment to Population Ratio for Puerto Ricans in the US by Education",
    x = "Year",
    y = "Participation Rate (%)"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

```{r EPR by Educaitonal Attainment Total Pop, message=FALSE, warning=FALSE}
ggplot(educ_lf_rate_us, aes(x = YEAR, y = etop_ratio, color = educ_grp)) +
  geom_line(linewidth = 1) + # Updated size to linewidth
  geom_point() +
  # important_years_lines +
  my_theme +
  labs(
    title = "Labor Force Participation Rate for US Population by Education",
    x = "Year",
    y = "Participation Rate (%)"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

### Unemployment Rate by Educational Attainment

```{r Unemployment by Educaitonal Attainment PR, message=FALSE, warning=FALSE}
ggplot(pr_educ_lf_rate_us, aes(x = YEAR, y = unemp_rate, color = educ_grp)) +
  geom_line(linewidth = 1) + # Updated size to linewidth
  geom_point() +
  # important_years_lines +
  my_theme +
  labs(
    title = "Employment to Population Ratio for Puerto Ricans in the US by Education",
    x = "Year",
    y = "Participation Rate (%)"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

```{r EPR by Educaitonal Attainment Total Pop, message=FALSE, warning=FALSE}
ggplot(educ_lf_rate_us, aes(x = YEAR, y = unemp_rate, color = educ_grp)) +
  geom_line(linewidth = 1) + # Updated size to linewidth
  geom_point() +
  # important_years_lines +
  my_theme +
  labs(
    title = "Labor Force Participation Rate for US Population by Education",
    x = "Year",
    y = "Participation Rate (%)"
  ) +
  scale_x_continuous(breaks = seq(1950, 2010, by = 10))
```

## LFPR by Education by Sex

```{r LFPR by EDUC by SEX, message=FALSE, warning=FALSE}
# Calculate for PR in US by Education Group by Sex
# Fixed: Changed AGE >= 16 to AGE >= 25
pr_educ_lf_rate_sex <- acs_with_educ %>%
  filter(
    HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261,
    AGE >= 25,
    EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3,
    SEX != 9
  ) %>%
  group_by(YEAR, SEX, educ_grp) %>%
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  mutate(
    sex = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female"
    ),
    labfor_rate = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor - unemp) / total_pop * 100,
    educ_attainment_pct = (total_pop / sum(total_pop)) * 100
  ) %>%
  ungroup() %>%
  collect()

# Calculate for Total US Population by Education Group by Sex
# Fixed: Added missing AGE >= 25 filter
educ_lf_rate_sex <- acs_with_educ %>%
  filter(AGE >= 25, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3, SEX != 9) %>%
  group_by(YEAR, SEX, educ_grp) %>%
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  mutate(
    sex = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female"
    ),
    labfor_rate = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor - unemp) / total_pop * 100,
    educ_attainment_pct = (total_pop / sum(total_pop)) * 100
  ) %>%
  ungroup() %>%
  collect()
```

```{r Education Attainment by sex for PR, message=FALSE, warning=FALSE}
ggplot(pr_educ_lf_rate_sex, aes(
  x = YEAR, y = educ_attainment_pct,
  color = educ_grp
)) +
  geom_line(linewidth = 1) +
  geom_point() +
  facet_wrap(~sex) +
  my_theme +
  labs(
    title = "Educational Attainment for Puerto Ricans in the US",
    x = "Year",
    y = "Percentage of Population (%)"
  )
```

```{r Education Attainment by sex for total pop, message=FALSE, warning=FALSE}
ggplot(educ_lf_rate_sex, aes(x = YEAR, y = educ_attainment_pct, color = educ_grp)) +
  geom_line(linewidth = 1) +
  geom_point() +
  facet_wrap(~sex) +
  my_theme +
  labs(
    title = "Educational Attainment for Total Populaiton in the US",
    x = "Year",
    y = "Percentage of Population (%)"
  )
```

### LFPR by Education Visualization by Sex

#### Puerto Ricans by Sex

```{r LFPR by Sex by EDUC for PR viz, message=FALSE, warning=FALSE}
ggplot(pr_educ_lf_rate_sex, aes(x = YEAR, y = labfor_rate, color = educ_grp)) +
  geom_line(linewidth = 1) + # Updated size to linewidth
  geom_point() +
  facet_wrap(~sex) +
  my_theme +
  labs(
    title = "Labor Force Participation Rate for Puerto Ricans by Education & Sex",
    x = "Year",
    y = "Participation Rate (%)"
  )
```

#### Total US Population by Sex

```{r LFPR by Sex by EDUC for total pop viz, message=FALSE, warning=FALSE}
ggplot(educ_lf_rate_sex, aes(x = YEAR, y = labfor_rate, color = educ_grp)) +
  geom_line(linewidth = 1) + # Updated size to linewidth
  geom_point() +
  facet_wrap(~sex) +
  my_theme +
  labs(
    title = "Labor Force Participation Rate for US Population by Education & Sex",
    x = "Year",
    y = "Participation Rate (%)"
  )
```

## Language Proficiency and LFPR

```{r Create Language Group, message=FALSE,warning=FALSE}
# Create a base table with the virtual grouping
acs_with_eng <- data_hist %>%
  mutate(
    eng_grp = case_when(
      SPEAKENG %in% c(1, 2) ~ "Limited English (Not well / Not at all)",
      SPEAKENG == 3 ~ "Speaks English Well",
      SPEAKENG == 4 ~ "Speaks English Very Well",
      SPEAKENG == 5 ~ "Speaks Only English",
      TRUE ~ "Unknown/NA"
    )
  ) %>%
  filter(eng_grp != "Unknown/NA")
```

```{r Calculate English Proficiency, message=FALSE, warning=FALSE}
# Calculate for PR in US by English Proficiency
pr_eng_lf_rate_us <- acs_with_eng %>%
  filter(
    HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261, AGE >= 16,
    EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3
  ) %>%
  group_by(YEAR, eng_grp) %>%
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  mutate(
    labfor_rate = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor - unemp) / total_pop * 100,
    eng_pct = total_pop / sum(total_pop) * 100,
  ) %>%
  collect()
```

### Visualize English Proficiency

```{r English Proficiency Viz, message=FALSE, warning=FALSE}
ggplot(pr_eng_lf_rate_us, aes(x = YEAR, y = labfor_rate, color = eng_grp)) +
  geom_line(linewidth = 1) +
  geom_point() +
  my_theme +
  labs(
    title = "Labor Force Participation Rate for Puerto Ricans in the US by English Proficiency",
    x = "Year",
    y = "Participation Rate (%)"
  )
```

## Industry Analysis

### Initial data prep

```{r Industry Data preparation, message=FALSE, warning=FALSE}
ind_sector_data <- data_hist %>%
  filter(
    HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261,
    AGE >= 16,
    EMPSTAT == 1,
  ) %>%
  mutate(ind_sector = case_when(
    IND1990 >= 10 & IND1990 <= 32 ~ "01. Agriculture, Forestry, Fishing & Hunting",
    IND1990 >= 40 & IND1990 <= 50 ~ "02. Mining, Quarrying, & Oil/Gas Extraction",
    IND1990 >= 450 & IND1990 <= 472 ~ "03. Utilities",
    IND1990 == 60 ~ "04. Construction",
    IND1990 >= 100 & IND1990 <= 392 ~ "05. Manufacturing",
    IND1990 >= 500 & IND1990 <= 571 ~ "06. Wholesale Trade",

    # Excludes 641 (Eating/Drinking Places) to align with modern NAICS grouping
    (IND1990 >= 580 & IND1990 <= 691) & IND1990 != 641 ~ "07. Retail Trade",
    IND1990 >= 400 & IND1990 <= 432 ~ "08. Transportation & Warehousing",
    IND1990 >= 440 & IND1990 <= 442 ~ "09. Information", # Maps 1990 Communications here
    IND1990 >= 700 & IND1990 <= 711 ~ "10. Finance & Insurance",
    IND1990 == 712 ~ "11. Real Estate & Rental/Leasing",

    # Merges 1990 Engineering/Architecture/Accounting with related professional codes
    IND1990 == 841 | (IND1990 >= 881 & IND1990 <= 893) ~ "12. Professional, Scientific, & Tech Services",
    IND1990 >= 721 & IND1990 <= 741 ~ "13. Administrative, Support, & Waste Mgmt",
    IND1990 >= 842 & IND1990 <= 860 ~ "14. Educational Services",
    (IND1990 >= 812 & IND1990 <= 840) | (IND1990 >= 861 & IND1990 <= 873) ~ "15. Health Care & Social Assistance",
    IND1990 >= 800 & IND1990 <= 810 ~ "16. Arts, Entertainment, & Recreation",

    # Combines 1990 Eating/Drinking places (641) with Hotels/Lodging (762, 770)
    IND1990 %in% c(641, 762, 770) ~ "17. Accommodation & Food Services",
    IND1990 %in% c(750, 751, 752, 761, 771, 772, 781, 791) ~ "18. Other Services (Except Public Admin)",
    IND1990 >= 900 & IND1990 <= 932 ~ "19. Public Administration",
    IND1990 >= 940 & IND1990 <= 960 ~ "20. Military",
    TRUE ~ "21. Unclassified / Not in Labor Force"
  )) %>%
  group_by(YEAR, ind_sector) %>%
  summarise(
    workers = sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # 2. Collect into local memory
  collect() %>%
  # 3. Local Calculation & Cleaning
  group_by(YEAR) %>%
  mutate(pct_workers = workers / sum(workers) * 100) %>%
  ungroup() %>%
  # Catch any missing sectors and label them
  mutate(ind_sector = replace_na(ind_sector, "21. Unclassified / Not in Labor Force"))

ind_sector_data_us <- data_hist %>%
  filter(
    AGE >= 16,
    EMPSTAT == 1,
  ) %>%
  mutate(ind_sector = case_when(
    IND1990 >= 10 & IND1990 <= 32 ~ "01. Agriculture, Forestry, Fishing & Hunting",
    IND1990 >= 40 & IND1990 <= 50 ~ "02. Mining, Quarrying, & Oil/Gas Extraction",
    IND1990 >= 450 & IND1990 <= 472 ~ "03. Utilities",
    IND1990 == 60 ~ "04. Construction",
    IND1990 >= 100 & IND1990 <= 392 ~ "05. Manufacturing",
    IND1990 >= 500 & IND1990 <= 571 ~ "06. Wholesale Trade",

    # Excludes 641 (Eating/Drinking Places) to align with modern NAICS grouping
    (IND1990 >= 580 & IND1990 <= 691) & IND1990 != 641 ~ "07. Retail Trade",
    IND1990 >= 400 & IND1990 <= 432 ~ "08. Transportation & Warehousing",
    IND1990 >= 440 & IND1990 <= 442 ~ "09. Information", # Maps 1990 Communications here
    IND1990 >= 700 & IND1990 <= 711 ~ "10. Finance & Insurance",
    IND1990 == 712 ~ "11. Real Estate & Rental/Leasing",

    # Merges 1990 Engineering/Architecture/Accounting with related professional codes
    IND1990 == 841 | (IND1990 >= 881 & IND1990 <= 893) ~ "12. Professional, Scientific, & Tech Services",
    IND1990 >= 721 & IND1990 <= 741 ~ "13. Administrative, Support, & Waste Mgmt",
    IND1990 >= 842 & IND1990 <= 860 ~ "14. Educational Services",
    (IND1990 >= 812 & IND1990 <= 840) | (IND1990 >= 861 & IND1990 <= 873) ~ "15. Health Care & Social Assistance",
    IND1990 >= 800 & IND1990 <= 810 ~ "16. Arts, Entertainment, & Recreation",

    # Combines 1990 Eating/Drinking places (641) with Hotels/Lodging (762, 770)
    IND1990 %in% c(641, 762, 770) ~ "17. Accommodation & Food Services",
    IND1990 %in% c(750, 751, 752, 761, 771, 772, 781, 791) ~ "18. Other Services (Except Public Admin)",
    IND1990 >= 900 & IND1990 <= 932 ~ "19. Public Administration",
    IND1990 >= 940 & IND1990 <= 960 ~ "20. Military",
    TRUE ~ "21. Unclassified / Not in Labor Force"
  )) %>%
  group_by(YEAR, ind_sector) %>%
  summarise(
    workers = sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # 2. Collect into local memory
  collect() %>%
  # 3. Local Calculation & Cleaning
  group_by(YEAR) %>%
  mutate(pct_workers = workers / sum(workers) * 100) %>%
  ungroup() %>%
  # Catch any missing sectors and label them
  mutate(ind_sector = replace_na(ind_sector, "21. Unclassified / Not in Labor Force"))
```

### Visualize Puerto Rican workers by industry

```{r PR workers by Industry, message=FALSE, warning=FALSE}
# Using facet_wrap to clearly view trends across 20+ sectors without overlap
ggplot(ind_sector_data, aes(x = YEAR, y = workers, color = ind_sector)) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~ind_sector, ncol = 4) +
  my_theme +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 8, face = "bold") # Shrink facet titles slightly to fit
  ) +
  labs(
    title = "Total PR Workers by Industry Sector",
    subtitle = "Estimated workers over time. Note: Y-axis scales vary by panel to show specific trends.",
    x = "Year",
    y = "Total Workers"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

### Visualize Total workers by Industry
```{r US workers by Industry, message=FALSE, warning=FALSE}
# Using facet_wrap to clearly view trends across 20+ sectors without overlap
ggplot(ind_sector_data_us, aes(x = YEAR, y = workers, color = ind_sector)) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~ind_sector, ncol = 4) +
  my_theme +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 8, face = "bold") # Shrink facet titles slightly to fit
  ) +
  labs(
    title = "Total US Workers by Industry Sector",
    subtitle = "Estimated workers over time. Note: Y-axis scales vary by panel to show specific trends.",
    x = "Year",
    y = "Total Workers"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

## Percentage point difference

```{r PR percent poing change calc, message=FALSE, warning=FALSE}
# Calculate the percentage point shift from the earliest to latest available year
ind_change <- ind_sector_data %>%
  group_by(ind_sector) %>%
  arrange(YEAR) %>%
  summarise(
    first_year = first(YEAR),
    last_year = last(YEAR),
    first_share = first(pct_workers),
    last_share = last(pct_workers),
    pp_change = last_share - first_share,
    .groups = "drop"
  ) %>%
  arrange(desc(pp_change))

# Display top growing and declining sectors
print(head(ind_change, 5))
print(tail(ind_change, 5))
```

```{r Total Pop percent poing change calc, message=FALSE, warning=FALSE}
# Calculate the percentage point shift from the earliest to latest available year
ind_change_us <- ind_sector_data_us %>%
  group_by(ind_sector) %>%
  arrange(YEAR) %>%
  summarise(
    first_year = first(YEAR),
    last_year = last(YEAR),
    first_share = first(pct_workers),
    last_share = last(pct_workers),
    pp_change = last_share - first_share,
    .groups = "drop"
  ) %>%
  arrange(desc(pp_change))

# Display top growing and declining sectors
print(head(ind_change_us, 5))
print(tail(ind_change_us, 5))
```

### Workforce Share by Industry in 1950

```{r PR Workforce Share by Industry viz, message=FALSE, warning=FALSE}
ggplot(ind_change, aes(x = first_share, y = reorder(ind_sector, first_share))) +
  # Using a single professional color instead of a distracting rainbow
  geom_col(fill = "#2c3e50", alpha = 0.9) +
  my_theme +
  # Remove the redundant legend
  theme(legend.position = "none") +
  # Format the x-axis to show percentage signs (assuming first_share is 0-100)
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Industry Share among Puerto Rican Workers, 1950",
    subtitle = "Workforce Share by Industry",
    x = "Share of Workforce",
    y = NULL # Removes the unnecessary "ind_sector" axis title
  )
```

### Workforce Share by Industry 1950 to 2010

```{r Puerto Rican workforce share by industry, message=FALSE, warning=FALSE}
ggplot(ind_sector_data, aes(x = pct_workers, y = ind_sector, fill = ind_sector)) +
  # Changed from line/point to a bar chart
  geom_col(fill = "#2c3e50", alpha = 0.9) +

  # Facet by YEAR instead of industry
  facet_wrap(~YEAR, ncol = 3) +
  my_theme +
  theme(
    # Keep legend hidden since the Y-axis already labels the industries
    legend.position = "none",
    strip.text = element_text(size = 10, face = "bold")
  ) +

  # X-axis is now the percentage since we flipped the orientation
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Puerto Rican Workforce Share by Industry",
    subtitle = "Percentage of total workforce across decennial years",
    x = "Share of Workforce (%)",
    y = NULL # Hides the redundant "ind_sector" axis title
  )
```

```{r PR percent poing change viz, message=FALSE, warning=FALSE}
ggplot(ind_change, aes(x = reorder(ind_sector, pp_change), y = pp_change, fill = pp_change > 0)) +
  geom_col() +
  coord_flip() +
  my_theme +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("FALSE" = "#d95f02", "TRUE" = "#1b9e77")) +
  labs(
    title = "Change in Industry Composition among Puerto Rican Workers",
    subtitle = "Percentage-point change in workforce share between first and last year",
    x = NULL,
    y = "Percentage Point Change"
  )
```

```{r Total pop PP viz, message=FALSE, warning=FALSE}
ggplot(ind_change_us, aes(x = reorder(ind_sector, pp_change), y = pp_change, fill = pp_change > 0)) +
  geom_col() +
  coord_flip() +
  my_theme +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("FALSE" = "#d95f02", "TRUE" = "#1b9e77")) +
  labs(
    title = "Change in Industry Composition among Total Workers",
    subtitle = "Percentage-point change in workforce share between first and last year",
    x = NULL,
    y = "Percentage Point Change"
  )
```

## Calculate index

```{r PR growth index calc, message=FALSE, warning=FALSE}
# Calculate an index to see relative growth within each sector (Baseline Year = 100)
ind_index <- ind_sector_data %>%
  filter(ind_sector != "20. Military", ind_sector != "21. Unclassified / Not in Labor Force") %>%
  group_by(ind_sector) %>%
  arrange(YEAR) %>%
  mutate(
    index = workers / first(workers) * 100
  ) %>%
  ungroup()
```

```{r US Pop index calc, message=FALSE,warning=FALSE}
# Calculate an index to see relative growth within each sector (Baseline Year = 100)
ind_index_us <- ind_sector_data_us %>%
  filter(ind_sector != "20. Military", ind_sector != "21. Unclassified / Not in Labor Force") %>%
  group_by(ind_sector) %>%
  arrange(YEAR) %>%
  mutate(
    index = workers / first(workers) * 100
  ) %>%
  ungroup()
```

```{r PR growth index plot, message=FALSE, warning=FALSE}
ggplot(ind_index, aes(x = YEAR, y = index, color = ind_sector)) +
  geom_line(size = 1) +
  geom_point(size = 1) +
  facet_wrap(~ind_sector, ncol = 4) +
  my_theme +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 8, face = "bold")
  ) +
  # Adding a horizontal baseline reference line at index = 100
  geom_hline(yintercept = 100, linetype = "dashed", color = "black", alpha = 0.5) +
  labs(
    title = "Industry Employment Growth Index",
    subtitle = "First year = 100. Dashed line represents baseline.",
    x = "Year",
    y = "Index"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

```{r Total growth index pop, message=FALSE, warning=FALSE}
ggplot(ind_index_us, aes(x = YEAR, y = index, color = ind_sector)) +
  geom_line(size = 1) +
  geom_point(size = 1) +
  facet_wrap(~ind_sector, ncol = 4) +
  my_theme +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 8, face = "bold")
  ) +
  # Adding a horizontal baseline reference line at index = 100
  geom_hline(yintercept = 100, linetype = "dashed", color = "black", alpha = 0.5) +
  labs(
    title = "Industry Employment Growth Index",
    subtitle = "First year = 100. Dashed line represents baseline.",
    x = "Year",
    y = "Index"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

### Over/Under Representation Index

```{r Over/under representation Index, message=FALSE, warning=FALSE}
# Calculate the Over/Under Representation Index
# > 1.0 means PR workers are overrepresented in this industry compared to the US
# < 1.0 means PR workers are underrepresented

representation_data <- ind_sector_data %>%
  # Select only the columns we need and rename the percentage column for clarity
  select(YEAR, ind_sector, pr_pct = pct_workers) %>%
  inner_join(
    # Join with the US data, also isolating and renaming its percentage column
    ind_sector_data_us %>% select(YEAR, ind_sector, us_pct = pct_workers),
    by = c("YEAR", "ind_sector")
  ) %>%
  mutate(
    rep_index = pr_pct / us_pct
  ) %>%
  # Filter out unclassified and military (if desired) for cleaner plotting
  filter(
    ind_sector != "21. Unclassified / Not in Labor Force",
    ind_sector != "20. Military"
  )
```

```{r Over/under representation Index plot, message=FALSE, warning=FALSE}
ggplot(representation_data, aes(x = YEAR, y = rep_index, color = ind_sector)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  # Use free_y so small index variations in smaller industries are visible
  facet_wrap(~ind_sector, ncol = 4, scales = "free_y") +

  # The crucial baseline: 1.0 means PR participation exactly matches US participation
  geom_hline(yintercept = 1, linetype = "dashed", color = "black", linewidth = 0.8) +
  my_theme +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 8, face = "bold")
  ) +
  labs(
    title = "Puerto Rican Worker Representation Index by Industry",
    subtitle = "Index > 1.0 = Overrepresented vs. US Average | Index < 1.0 = Underrepresented",
    x = "Year",
    y = "Representation Index (PR Share / US Share)"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

### Over/Under Representation Index Adjusted for Small Denominator

```{r Over/under representation Index Adjusted, message=FALSE, warning=FALSE}
# Calculate the Controlled Over/Under Representation Index
# Using the total US workforce as the baseline
representation_data_national_controlled <- ind_sector_data %>%
  select(YEAR, ind_sector, pr_pct = pct_workers) %>%
  inner_join(
    # Reverting back to the national baseline data
    ind_sector_data_us %>% select(YEAR, ind_sector, us_pct = pct_workers),
    by = c("YEAR", "ind_sector")
  ) %>%
  # <--- Small Denominator Control: US Baseline must be at least 1% of the workforce
  filter(us_pct >= 1.0) %>%
  mutate(
    rep_index = pr_pct / us_pct
  ) %>%
  # Filter out unclassified and military sectors
  filter(
    ind_sector != "21. Unclassified / Not in Labor Force",
    ind_sector != "20. Military"
  )
```

```{r Over/under representation Index Adjusted Viz, message=FALSE, warning=FALSE}
ggplot(representation_data_national_controlled, aes(x = YEAR, y = rep_index, color = ind_sector)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~ind_sector, ncol = 4, scales = "free_y") +

  # The 1.0 baseline indicates exact parity with the total US workforce
  geom_hline(yintercept = 1, linetype = "dashed", color = "black", linewidth = 0.8) +
  my_theme +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 8, face = "bold")
  ) +
  labs(
    title = "Puerto Rican Worker Representation Index by Industry",
    subtitle = "Compared to National Baseline | Excludes industries < 1% of total US workforce",
    x = "Year",
    y = "Representation Index (PR Share / Total US Share)"
  )
```

## Industry Analysis by Sex

### Calculate Industry metrics by Sex

```{r Calculate Industry by Sex, message=FALSE,warning=FALSE}
# Calculate Puerto Rican workers by Industry and Sex
pr_ind_sex <- data_hist %>%
  filter(
    HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261,
    AGE >= 16,
    EMPSTAT == 1,
    SEX != 9 # Remove unknown sex
  ) %>%
  mutate(
    sex_label = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female"
    ),
    ind_sector = case_when(
      IND1990 >= 10 & IND1990 <= 32 ~ "01. Agriculture, Forestry, Fishing & Hunting",
      IND1990 >= 40 & IND1990 <= 50 ~ "02. Mining, Quarrying, & Oil/Gas Extraction",
      IND1990 >= 450 & IND1990 <= 472 ~ "03. Utilities",
      IND1990 == 60 ~ "04. Construction",
      IND1990 >= 100 & IND1990 <= 392 ~ "05. Manufacturing",
      IND1990 >= 500 & IND1990 <= 571 ~ "06. Wholesale Trade",
      (IND1990 >= 580 & IND1990 <= 691) & IND1990 != 641 ~ "07. Retail Trade",
      IND1990 >= 400 & IND1990 <= 432 ~ "08. Transportation & Warehousing",
      IND1990 >= 440 & IND1990 <= 442 ~ "09. Information",
      IND1990 >= 700 & IND1990 <= 711 ~ "10. Finance & Insurance",
      IND1990 == 712 ~ "11. Real Estate & Rental/Leasing",
      IND1990 == 841 | (IND1990 >= 881 & IND1990 <= 893) ~ "12. Professional, Scientific, & Tech Services",
      IND1990 >= 721 & IND1990 <= 741 ~ "13. Administrative, Support, & Waste Mgmt",
      IND1990 >= 842 & IND1990 <= 860 ~ "14. Educational Services",
      (IND1990 >= 812 & IND1990 <= 840) | (IND1990 >= 861 & IND1990 <= 873) ~ "15. Health Care & Social Assistance",
      IND1990 >= 800 & IND1990 <= 810 ~ "16. Arts, Entertainment, & Recreation",
      IND1990 %in% c(641, 762, 770) ~ "17. Accommodation & Food Services",
      IND1990 %in% c(750, 751, 752, 761, 771, 772, 781, 791) ~ "18. Other Services (Except Public Admin)",
      IND1990 >= 900 & IND1990 <= 932 ~ "19. Public Administration",
      IND1990 >= 940 & IND1990 <= 960 ~ "20. Military",
      TRUE ~ "21. Unclassified / Not in Labor Force"
    )
  ) %>%
  group_by(YEAR, sex_label, ind_sector) %>%
  summarise(
    workers = sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  collect() %>%
  # Calculate percentages within each sex group per year
  group_by(YEAR, sex_label) %>%
  mutate(
    pct_workers = (workers / sum(workers)) * 100
  ) %>%
  ungroup() %>%
  mutate(ind_sector = replace_na(ind_sector, "21. Unclassified / Not in Labor Force"))

# Calculate Puerto Rican workers by Industry and Sex
us_ind_sex <- data_hist %>%
  filter(
    AGE >= 16,
    EMPSTAT == 1,
    SEX != 9 # Remove unknown sex
  ) %>%
  mutate(
    sex_label = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female"
    ),
    ind_sector = case_when(
      IND1990 >= 10 & IND1990 <= 32 ~ "01. Agriculture, Forestry, Fishing & Hunting",
      IND1990 >= 40 & IND1990 <= 50 ~ "02. Mining, Quarrying, & Oil/Gas Extraction",
      IND1990 >= 450 & IND1990 <= 472 ~ "03. Utilities",
      IND1990 == 60 ~ "04. Construction",
      IND1990 >= 100 & IND1990 <= 392 ~ "05. Manufacturing",
      IND1990 >= 500 & IND1990 <= 571 ~ "06. Wholesale Trade",
      (IND1990 >= 580 & IND1990 <= 691) & IND1990 != 641 ~ "07. Retail Trade",
      IND1990 >= 400 & IND1990 <= 432 ~ "08. Transportation & Warehousing",
      IND1990 >= 440 & IND1990 <= 442 ~ "09. Information",
      IND1990 >= 700 & IND1990 <= 711 ~ "10. Finance & Insurance",
      IND1990 == 712 ~ "11. Real Estate & Rental/Leasing",
      IND1990 == 841 | (IND1990 >= 881 & IND1990 <= 893) ~ "12. Professional, Scientific, & Tech Services",
      IND1990 >= 721 & IND1990 <= 741 ~ "13. Administrative, Support, & Waste Mgmt",
      IND1990 >= 842 & IND1990 <= 860 ~ "14. Educational Services",
      (IND1990 >= 812 & IND1990 <= 840) | (IND1990 >= 861 & IND1990 <= 873) ~ "15. Health Care & Social Assistance",
      IND1990 >= 800 & IND1990 <= 810 ~ "16. Arts, Entertainment, & Recreation",
      IND1990 %in% c(641, 762, 770) ~ "17. Accommodation & Food Services",
      IND1990 %in% c(750, 751, 752, 761, 771, 772, 781, 791) ~ "18. Other Services (Except Public Admin)",
      IND1990 >= 900 & IND1990 <= 932 ~ "19. Public Administration",
      IND1990 >= 940 & IND1990 <= 960 ~ "20. Military",
      TRUE ~ "21. Unclassified / Not in Labor Force"
    )
  ) %>%
  group_by(YEAR, sex_label, ind_sector) %>%
  summarise(
    workers = sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  collect() %>%
  # Calculate percentages within each sex group per year
  group_by(YEAR, sex_label) %>%
  mutate(
    pct_workers = (workers / sum(workers)) * 100
  ) %>%
  ungroup() %>%
  mutate(ind_sector = replace_na(ind_sector, "21. Unclassified / Not in Labor Force"))
```

### Visualize Puerto Rican workers by Industry by Sex

```{r PR worker share by Ind by Sex, message=FALSE, warning=FALSE}
ggplot(pr_ind_sex, aes(x = YEAR, y = pct_workers, color = sex_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~ind_sector, ncol = 4) +

  # Use distinct colors for contrast
  scale_color_manual(values = c("Male" = "#1f78b4", "Female" = "#e31a1c")) +
  my_theme +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 8, face = "bold")
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Puerto Rican Workforce Share by Industry and Sex (1950-2010)",
    subtitle = "Percentage of male vs. female workforce in each sector.",
    x = "Year",
    y = "Share of Workforce (%)",
    color = "Sex"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

### Visualize Total workers by Industry by Sex

```{r US worker share by Ind by Sex, message=FALSE, warning=FALSE}
ggplot(us_ind_sex, aes(x = YEAR, y = pct_workers, color = sex_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~ind_sector, ncol = 4) +

  # Use distinct colors for contrast
  scale_color_manual(values = c("Male" = "#1f78b4", "Female" = "#e31a1c")) +
  my_theme +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 8, face = "bold")
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Total Workforce Share by Industry and Sex (1950-2010)",
    subtitle = "Percentage of male vs. female workforce in each sector.",
    x = "Year",
    y = "Share of Workforce (%)",
    color = "Sex"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

## Industry analysis by Income

### Calculate Industry Income

```{r Calculate Industry Income, message=FALSE,warning=FALSE}
# Calculate average wage by industry for Puerto Rican workers
pr_income_ind <- data_hist %>%
  filter(
    HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261,
    AGE >= 16,
    EMPSTAT == 1,
    INCWAGE > 0, INCWAGE < 999999 # Filter out N/A and zero-income rows
  ) %>%
  mutate(ind_sector = case_when(
    IND1990 >= 10 & IND1990 <= 32 ~ "01. Agriculture, Forestry, Fishing & Hunting",
    IND1990 >= 40 & IND1990 <= 50 ~ "02. Mining, Quarrying, & Oil/Gas Extraction",
    IND1990 >= 450 & IND1990 <= 472 ~ "03. Utilities",
    IND1990 == 60 ~ "04. Construction",
    IND1990 >= 100 & IND1990 <= 392 ~ "05. Manufacturing",
    IND1990 >= 500 & IND1990 <= 571 ~ "06. Wholesale Trade",
    (IND1990 >= 580 & IND1990 <= 691) & IND1990 != 641 ~ "07. Retail Trade",
    IND1990 >= 400 & IND1990 <= 432 ~ "08. Transportation & Warehousing",
    IND1990 >= 440 & IND1990 <= 442 ~ "09. Information",
    IND1990 >= 700 & IND1990 <= 711 ~ "10. Finance & Insurance",
    IND1990 == 712 ~ "11. Real Estate & Rental/Leasing",
    IND1990 == 841 | (IND1990 >= 881 & IND1990 <= 893) ~ "12. Professional, Scientific, & Tech Services",
    IND1990 >= 721 & IND1990 <= 741 ~ "13. Administrative, Support, & Waste Mgmt",
    IND1990 >= 842 & IND1990 <= 860 ~ "14. Educational Services",
    (IND1990 >= 812 & IND1990 <= 840) | (IND1990 >= 861 & IND1990 <= 873) ~ "15. Health Care & Social Assistance",
    IND1990 >= 800 & IND1990 <= 810 ~ "16. Arts, Entertainment, & Recreation",
    IND1990 %in% c(641, 762, 770) ~ "17. Accommodation & Food Services",
    IND1990 %in% c(750, 751, 752, 761, 771, 772, 781, 791) ~ "18. Other Services (Except Public Admin)",
    IND1990 >= 900 & IND1990 <= 932 ~ "19. Public Administration",
    IND1990 >= 940 & IND1990 <= 960 ~ "20. Military",
    TRUE ~ "21. Unclassified / Not in Labor Force"
  )) %>%
  group_by(YEAR, ind_sector) %>%
  summarise(
    avg_wage = sum(INCWAGE * CPI99 * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE),
    # avg_wage = sum(INCWAGE * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE),
    total_workers = sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  collect() %>%
  mutate(ind_sector = replace_na(ind_sector, "21. Unclassified / Not in Labor Force"))

us_income_ind <- data_hist %>%
  filter(
    AGE >= 16,
    EMPSTAT == 1,
    INCWAGE > 0, INCWAGE < 999999 # Filter out N/A and zero-income rows
  ) %>%
  mutate(ind_sector = case_when(
    IND1990 >= 10 & IND1990 <= 32 ~ "01. Agriculture, Forestry, Fishing & Hunting",
    IND1990 >= 40 & IND1990 <= 50 ~ "02. Mining, Quarrying, & Oil/Gas Extraction",
    IND1990 >= 450 & IND1990 <= 472 ~ "03. Utilities",
    IND1990 == 60 ~ "04. Construction",
    IND1990 >= 100 & IND1990 <= 392 ~ "05. Manufacturing",
    IND1990 >= 500 & IND1990 <= 571 ~ "06. Wholesale Trade",
    (IND1990 >= 580 & IND1990 <= 691) & IND1990 != 641 ~ "07. Retail Trade",
    IND1990 >= 400 & IND1990 <= 432 ~ "08. Transportation & Warehousing",
    IND1990 >= 440 & IND1990 <= 442 ~ "09. Information",
    IND1990 >= 700 & IND1990 <= 711 ~ "10. Finance & Insurance",
    IND1990 == 712 ~ "11. Real Estate & Rental/Leasing",
    IND1990 == 841 | (IND1990 >= 881 & IND1990 <= 893) ~ "12. Professional, Scientific, & Tech Services",
    IND1990 >= 721 & IND1990 <= 741 ~ "13. Administrative, Support, & Waste Mgmt",
    IND1990 >= 842 & IND1990 <= 860 ~ "14. Educational Services",
    (IND1990 >= 812 & IND1990 <= 840) | (IND1990 >= 861 & IND1990 <= 873) ~ "15. Health Care & Social Assistance",
    IND1990 >= 800 & IND1990 <= 810 ~ "16. Arts, Entertainment, & Recreation",
    IND1990 %in% c(641, 762, 770) ~ "17. Accommodation & Food Services",
    IND1990 %in% c(750, 751, 752, 761, 771, 772, 781, 791) ~ "18. Other Services (Except Public Admin)",
    IND1990 >= 900 & IND1990 <= 932 ~ "19. Public Administration",
    IND1990 >= 940 & IND1990 <= 960 ~ "20. Military",
    TRUE ~ "21. Unclassified / Not in Labor Force"
  )) %>%
  group_by(YEAR, ind_sector) %>%
  summarise(
    us_avg_wage = sum(INCWAGE * CPI99 * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE),
    # us_avg_wage = sum(INCWAGE * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE),
    total_workers = sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  collect() %>%
  mutate(ind_sector = replace_na(ind_sector, "21. Unclassified / Not in Labor Force"))
```

### Visualize Income trajectory 

```{r PR Income Trajectory by industry, message=FALSE, warning=FALSE}
ggplot(pr_income_ind, aes(x = YEAR, y = avg_wage, color = ind_sector)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~ind_sector, ncol = 4) +
  my_theme +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 8, face = "bold")
  ) +
  # Format Y-axis as dollars
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
  labs(
    title = "Average Annual Wage for Puerto Rican Workers by Industry",
    subtitle = "Nominal Dollars.",
    x = "Year",
    y = "Average Wage ($)"
  )
```

### Best paying industries for Puerto Ricans

```{r Best paying industry 2024, message=FALSE,warning=FALSE}
# Filter for just the final year
pr_income_2024 <- pr_income_ind %>%
  filter(YEAR == 2024, ind_sector != "21. Unclassified / Not in Labor Force")

ggplot(pr_income_2024, aes(x = avg_wage, y = reorder(ind_sector, avg_wage))) +
  geom_col(fill = "#1b9e77", alpha = 0.9) +
  my_theme +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Highest Paying Industries for Puerto Rican Workers (2024)",
    subtitle = "Average Annual Wage by Sector",
    x = "Average Annual Wage ($)",
    y = NULL
  )
```

### Best paying industries for total pop

```{r Best paying industry 2024 total pop, message=FALSE,warning=FALSE}
# Filter for just the final year
us_income_2024 <- us_income_ind %>%
  filter(YEAR == 2024, ind_sector != "21. Unclassified / Not in Labor Force")

ggplot(us_income_2024, aes(x = us_avg_wage, y = reorder(ind_sector, us_avg_wage))) +
  geom_col(fill = "#1b9e77", alpha = 0.9) +
  my_theme +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Highest Paying Industries for Total Pop Workers (2024)",
    subtitle = "Average Annual Wage by Sector",
    x = "Average Annual Wage ($)",
    y = NULL
  )
```

## Calculate wage gap by industry for Puerto Ricans

```{r Wage gap calculation, message=FALSE,warning=FALSE}
# Join the datasets and calculate the wage gap
wage_gap_ind <- pr_income_ind %>%
  # Rename pr_income_ind's avg_wage to pr_avg_wage for clarity before joining
  select(YEAR, ind_sector, pr_avg_wage = avg_wage) %>%
  inner_join(
    us_income_ind %>% select(YEAR, ind_sector, us_avg_wage),
    by = c("YEAR", "ind_sector")
  ) %>%
  mutate(
    # A ratio of 1.0 means exact wage parity
    wage_ratio = pr_avg_wage / us_avg_wage,

    # Optional: Convert to percentage difference (e.g., -15% or +5%)
    wage_gap_pct = (wage_ratio - 1) * 100
  ) %>%
  filter(
    ind_sector != "21. Unclassified / Not in Labor Force",
    ind_sector != "20. Military"
  )
```

```{r Wage gap vizualization, message=FALSE,warning=FALSE}
ggplot(wage_gap_ind, aes(x = YEAR, y = wage_ratio, color = ind_sector)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~ind_sector, ncol = 4) +

  # The Parity Baseline: 1.0 = Equal Pay
  geom_hline(yintercept = 1.0, linetype = "dashed", color = "black", linewidth = 0.8) +
  my_theme +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 8, face = "bold")
  ) +
  # Format Y-axis to show ratios cleanly (e.g., 0.75, 1.00)
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
  labs(
    title = "Industry Wage Gap for Puerto Rican Workers (1950-2024)",
    subtitle = "Ratio of PR Average Wage to US Average Wage. 1.0 = Wage Parity.",
    x = "Year",
    y = "Wage Ratio (PR / US)"
  )
```

## Wage Tier Analysis: Are Puerto Ricans gaining upward mobility

### Wage tier data prep

```{r Wage Tier Calculation, message = FALSE, warning = FALSE}
# 1. Create the Wage Tiers based on the National Economy
industry_tiers <- us_income_ind %>%
  group_by(YEAR) %>%
  mutate(
    # ntile(3) splits the industries into bottom, middle, and top thirds based on wage
    tier_num = ntile(us_avg_wage, 3),
    wage_tier = case_when(
      tier_num == 3 ~ "High-Paying Industries",
      tier_num == 2 ~ "Middle-Paying Industries",
      tier_num == 1 ~ "Low-Paying Industries"
    ),
    # Let's make it a factor so it plots in a logical order
    wage_tier = factor(wage_tier, levels = c(
      "High-Paying Industries",
      "Middle-Paying Industries",
      "Low-Paying Industries"
    ))
  ) %>%
  select(YEAR, ind_sector, wage_tier) %>%
  ungroup()

# 2. Calculate Tier Distribution for Puerto Rican Workers
pr_tier_trends <- ind_sector_data %>%
  inner_join(industry_tiers, by = c("YEAR", "ind_sector")) %>%
  # Group by year and the new wage tier
  group_by(YEAR, wage_tier) %>%
  summarise(tier_workers = sum(workers, na.rm = TRUE), .groups = "drop_last") %>%
  mutate(
    pct_in_tier = (tier_workers / sum(tier_workers)) * 100,
    demographic = "Puerto Rican Workers" # Label for the final plot
  ) %>%
  ungroup()

# 3. Calculate Tier Distribution for Total US Workers
us_tier_trends <- ind_sector_data_us %>%
  inner_join(industry_tiers, by = c("YEAR", "ind_sector")) %>%
  group_by(YEAR, wage_tier) %>%
  summarise(tier_workers = sum(workers, na.rm = TRUE), .groups = "drop_last") %>%
  mutate(
    pct_in_tier = (tier_workers / sum(tier_workers)) * 100,
    demographic = "Total US Workers" # Label for the final plot
  ) %>%
  ungroup()

# 4. Combine them into one dataframe for easy plotting
combined_tier_trends <- bind_rows(pr_tier_trends, us_tier_trends)
```

### Wage tier Visualization

```{r Visualize Wage tier analysis, message = FALSE, warning = FALSE}
ggplot(combined_tier_trends, aes(x = YEAR, y = pct_in_tier, color = demographic)) +
  geom_line(linewidth = 0.75) +
  geom_point(size = 2) +
  facet_wrap(~wage_tier, ncol = 3) +

  # Use distinct colors to clearly separate the two groups
  scale_color_manual(values = c(
    "Puerto Rican Workers" = "#d95f02",
    "Total US Workers" = "#7570b3"
  )) +
  my_theme +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 10, face = "bold")
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
  labs(
    title = "Workforce Distribution Across Industry Wage Tiers (1950-2024)",
    subtitle = "Comparing PR workforce composition against the US baseline",
    x = "Year",
    y = "Share of Workforce (%)",
    color = NULL
  )
```

### Share of Puerto Rican workers by Wage Tiers

```{r Join Puerto Rican workers to Wage tier, message=FALSE, warning=FALSE}
wage_tier_pct_pr_wrk <- industry_tiers %>%
  inner_join(
    ind_sector_data %>% select(YEAR, ind_sector, pr_pct = pct_workers),
    by = c("YEAR", "ind_sector")
  )
```

```{r}
# 1. Aggregate the industry percentages into the three wage tiers
pr_tier_summary <- wage_tier_pct_pr_wrk %>%
  group_by(YEAR, wage_tier) %>%
  summarise(
    total_tier_pct = sum(pr_pct, na.rm = TRUE),
    .groups = "drop"
  )

# 2. Plot the stacked area chart
ggplot(pr_tier_summary, aes(x = YEAR, y = total_tier_pct, fill = wage_tier)) +
  # geom_area automatically stacks the values; color = "white" adds clean borders
  geom_area(alpha = 0.85, color = "white", linewidth = 0.4) +

  # Optional: Define specific colors for Low, Middle, High tiers
  scale_fill_manual(values = c(
    "High-Paying Industries" = "#1b9e77",
    "Middle-Paying Industries" = "#d95f02",
    "Low-Paying Industries" = "#7570b3"
  )) +
  my_theme +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
  labs(
    title = "Puerto Rican Workforce Distribution by Wage Tier",
    subtitle = "Percentage of workers in High, Middle, and Low paying industries",
    x = "Year",
    y = "Share of Workforce (%)"
  )
```

```{r}
ggplot(pr_tier_summary, aes(x = YEAR, y = total_tier_pct, color = wage_tier)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c(
    "High-Paying Industries" = "#1b9e77",
    "Middle-Paying Industries" = "#d95f02",
    "Low-Paying Industries" = "#7570b3"
  )) +
  my_theme +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
  labs(
    title = "Puerto Rican Workforce Trajectory by Wage Tier",
    subtitle = "Tracking the share of workers in each tier over time",
    x = "Year",
    y = "Share of Workforce (%)"
  )
```

### Change in Puerto Rican Worker Share and Wage Tier

```{r}
ggplot(wage_tier_pct_pr_wrk, aes(x = YEAR, y = pr_pct)) +
  # 1. Draw a continuous neutral line to show the trend in workforce share
  geom_line(color = "black", linewidth = 0.75) +

  # 2. Add large points colored by the wage tier for that specific decade
  geom_point(aes(color = wage_tier), size = 2) +

  # 3. Facet by industry so each has its own panel
  facet_wrap(~ind_sector, ncol = 4) +

  # 4. Apply a distinct color palette for the tiers
  scale_color_manual(values = c(
    "High-Paying Industries" = "#1b9e77", # Green
    "Middle-Paying Industries" = "#d95f02", # Orange
    "Low-Paying Industries" = "#7570b3" # Purple
  )) +
  my_theme +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    strip.text = element_text(size = 8, face = "bold")
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
  labs(
    title = "Puerto Rican Workforce Share and Industry Wage Tiers",
    subtitle = "Tracking industry participation. Dot colors indicate the national wage tier for that decade.",
    x = "Year",
    y = "Share of Puerto Rican Workforce (%)"
  )
```

```{r Puerto Rican workforce share by industry by wage tier, message=, warning=FALSE}
wage_tier_pct_pr_wrk %>%
  filter(ind_sector != "21. Unclassified / Not in Labor Force") %>%
  ggplot(aes(x = pr_pct, y = ind_sector, fill = wage_tier)) +
  geom_col() +

  # Facet by YEAR
  facet_wrap(~YEAR, ncol = 3) +

  # FIXED: Changed to scale_fill_manual to match the aes() mapping
  scale_fill_manual(values = c(
    "High-Paying Industries" = "#1b9e77", # Green
    "Middle-Paying Industries" = "#d95f02", # Orange
    "Low-Paying Industries" = "#7570b3" # Purple
  )) +
  my_theme +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 7)
  ) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +

  # FIXED: Updated labels to match the horizontal bar chart orientation
  labs(
    title = "Puerto Rican Workforce Share and Industry Wage Tiers",
    subtitle = "Tracking industry participation. Bar colors indicate the national wage tier for that decade.",
    x = "Share of Puerto Rican Workforce (%)",
    y = NULL
  )
```

## Occupation Analysis

### Initial data prep

```{r Occupation data prep, message=FALSE,warning=FALSE}
occ_cat_data <- data_hist %>%
  filter(HISPAN == 2 |
           HISPAND == 200 | ANCESTR1 == 261, AGE >= 16, EMPSTAT == 1) %>%
  # 1. SERVER SIDE: Create the text labels using case_when
  mutate(
    occ_cat1 = case_when(
      (OCC2010 >= 10 &
         OCC2010 <= 960) |
        (OCC2010 >= 2600 &
           OCC2010 <= 2960) ~ "Management, Business, Entertainment & Media",
      OCC2010 >= 1000 & OCC2010 <= 1965 ~ "STEM",
      (OCC2010 >= 2000 &
         OCC2010 <= 2550) |
        (OCC2010 >= 4300 &
           OCC2010 <= 4650) ~ "Education, Community & Social Services",
      OCC2010 >= 3000 & OCC2010 <= 3655 ~ "Health & Medical",
      (OCC2010 >= 3700 &
         OCC2010 <= 3955) |
        (OCC2010 >= 9000 &
           OCC2010 <= 9420) ~ "Law Enforcement & Transportation",
      OCC2010 >= 4000 &
        OCC2010 <= 4250 ~ "Food, Maintenance, Building & Grounds Cleaning",
      OCC2010 >= 4700 & OCC2010 <= 4965 ~ "Sales and Related",
      OCC2010 >= 5000 &
        OCC2010 <= 5940 ~ "Administrative and Office Support",
      OCC2010 >= 6005 &
        OCC2010 <= 7630 ~ "Natural Resources, Construction & Maintenance",
      (OCC2010 >= 7700 &
         OCC2010 <= 8965) |
        (OCC2010 >= 9500 &
           OCC2010 <= 9750) ~ "Production and Material Moving",
      OCC2010 >= 9800 & OCC2010 <= 9830 ~ "Military",
      OCC2010 %in% c(9920, 9999) ~ "Unclassified",
      TRUE ~ "Unclassified"
    )
  ) %>%
  group_by(YEAR, occ_cat1) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  # 2. BRING TO LOCAL MEMORY
  collect() %>%
  
  # 3. LOCAL R SIDE: Now it is safe to use R-specific functions
  mutate(
    # Catch any NAs before factoring
    occ_cat1 = replace_na(occ_cat1, "Unclassified"),
    
    # Convert to factor to lock in your custom plotting order
    occ_cat1 = factor(
      occ_cat1,
      levels = c(
        "Management, Business, Entertainment & Media",
        "STEM",
        "Education, Community & Social Services",
        "Health & Medical",
        "Law Enforcement & Transportation",
        "Food, Maintenance, Building & Grounds Cleaning",
        "Sales and Related",
        "Administrative and Office Support",
        "Natural Resources, Construction & Maintenance",
        "Production and Material Moving",
        "Military",
        "Unclassified"
      )
    )
  ) %>%
  # 4. Finish your calculations
  group_by(YEAR) %>%
  mutate(pct_workers = workers / sum(workers) * 100) %>%
  ungroup()

occ_cat_data_us <- data_hist %>%
  filter(AGE >= 16, EMPSTAT == 1) %>%
  mutate(
    occ_cat1 = case_when(
      (OCC2010 >= 10 &
         OCC2010 <= 960) |
        (OCC2010 >= 2600 &
           OCC2010 <= 2960) ~ "Management, Business, Entertainment & Media",
      OCC2010 >= 1000 & OCC2010 <= 1965 ~ "STEM",
      (OCC2010 >= 2000 &
         OCC2010 <= 2550) |
        (OCC2010 >= 4300 &
           OCC2010 <= 4650) ~ "Education, Community & Social Services",
      OCC2010 >= 3000 & OCC2010 <= 3655 ~ "Health & Medical",
      (OCC2010 >= 3700 &
         OCC2010 <= 3955) |
        (OCC2010 >= 9000 &
           OCC2010 <= 9420) ~ "Law Enforcement & Transportation",
      OCC2010 >= 4000 &
        OCC2010 <= 4250 ~ "Food, Maintenance, Building & Grounds Cleaning",
      OCC2010 >= 4700 & OCC2010 <= 4965 ~ "Sales and Related",
      OCC2010 >= 5000 &
        OCC2010 <= 5940 ~ "Administrative and Office Support",
      OCC2010 >= 6005 &
        OCC2010 <= 7630 ~ "Natural Resources, Construction & Maintenance",
      (OCC2010 >= 7700 &
         OCC2010 <= 8965) |
        (OCC2010 >= 9500 &
           OCC2010 <= 9750) ~ "Production and Material Moving",
      OCC2010 >= 9800 & OCC2010 <= 9830 ~ "Military",
      OCC2010 %in% c(9920, 9999) ~ "Unclassified",
      TRUE ~ "Unclassified"
    )
  ) %>%
  group_by(YEAR, occ_cat1) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  # 2. BRING TO LOCAL MEMORY
  collect() %>%
  
  # 3. LOCAL R SIDE: Now it is safe to use R-specific functions
  mutate(
    # Catch any NAs before factoring
    occ_cat1 = replace_na(occ_cat1, "Unclassified"),
    
    # Convert to factor to lock in your custom plotting order
    occ_cat1 = factor(
      occ_cat1,
      levels = c(
        "Management, Business, Entertainment & Media",
        "STEM",
        "Education, Community & Social Services",
        "Health & Medical",
        "Law Enforcement & Transportation",
        "Food, Maintenance, Building & Grounds Cleaning",
        "Sales and Related",
        "Administrative and Office Support",
        "Natural Resources, Construction & Maintenance",
        "Production and Material Moving",
        "Military",
        "Unclassified"
      )
    )
  ) %>%
  # 4. Finish your calculations
  group_by(YEAR) %>%
  mutate(pct_workers = workers / sum(workers) * 100) %>%
  ungroup()
```


### Visualize Puerto Rican workers by occupation

```{r Visualize PR workers by Occ, message=FALSE,warning=FALSE}
ggplot(occ_cat_data, aes(x = YEAR, y = workers, color = occ_cat1)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap( ~ occ_cat1, ncol = 4) +
  my_theme +
  theme(legend.position = "none",
        strip.text = element_text(size = 8, face = "bold")) +
  labs(
    title = "Total PR Workers by Occupational Category",
    subtitle = "Estimated workers over time. Note: Y-axis scales vary by panel to show specific trends.",
    x = "Year",
    y = "Total Workers"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```


### Visualize Total workers by occupation

```{r Visualize Total workers by Occ, message=FALSE,warning=FALSE}
ggplot(occ_cat_data_us, aes(x = YEAR, y = workers, color = occ_cat1)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap( ~ occ_cat1, ncol = 4) +
  my_theme +
  theme(legend.position = "none",
        strip.text = element_text(size = 8, face = "bold")) +
  labs(
    title = "Total US Workers by Occupational Category",
    subtitle = "Estimated workers over time. Note: Y-axis scales vary by panel to show specific trends.",
    x = "Year",
    y = "Total Workers"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

### Percentage point difference

```{r Pct Point Diff Calculation, message=FALSE, warning=FALSE}
occ_change <- occ_cat_data %>%
  group_by(occ_cat1) %>%
  arrange(YEAR) %>%
  summarise(
    first_year = first(YEAR),
    last_year = last(YEAR),
    first_share = first(pct_workers),
    last_share = last(pct_workers),
    pp_change = last_share - first_share,
    .groups = "drop"
  ) %>%
  arrange(desc(pp_change))

print(head(occ_change, 3))
print(tail(occ_change, 3))

occ_change_us <- occ_cat_data_us %>%
  group_by(occ_cat1) %>%
  arrange(YEAR) %>%
  summarise(
    first_year = first(YEAR),
    last_year = last(YEAR),
    first_share = first(pct_workers),
    last_share = last(pct_workers),
    pp_change = last_share - first_share,
    .groups = "drop"
  ) %>%
  arrange(desc(pp_change))

print(head(occ_change_us, 3))
print(tail(occ_change_us, 3))
```

```{r Pct point change PR Visualization, message=FALSE, warning=FALSE}
ggplot(occ_change, aes(
  x = reorder(occ_cat1, pp_change),
  y = pp_change,
  fill = pp_change > 0
)) +
  geom_col() +
  coord_flip() +
  my_theme +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("FALSE" = "#d95f02", "TRUE" = "#1b9e77")) +
  labs(
    title = "Change in Occupational Composition among Puerto Rican Workers",
    subtitle = "Percentage-point change in workforce share between first and last year",
    x = NULL,
    y = "Percentage Point Change"
  )
```

```{r Pct point change Total Visualization, message=FALSE, warning=FALSE}

ggplot(occ_change_us, aes(
  x = reorder(occ_cat1, pp_change),
  y = pp_change,
  fill = pp_change > 0
)) +
  geom_col() +
  coord_flip() +
  my_theme +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("FALSE" = "#d95f02", "TRUE" = "#1b9e77")) +
  labs(
    title = "Change in Occupational Composition among Total Workers",
    subtitle = "Percentage-point change in workforce share between first and last year",
    x = NULL,
    y = "Percentage Point Change"
  )

```

### Workforce Share by Occupation in 1950

```{r Workforce Share 1950, message=FALSE, warning=FALSE}
ggplot(occ_change, aes(x = first_share, y = reorder(occ_cat1, first_share))) +
  geom_col(fill = "#2c3e50", alpha = 0.9) +
  my_theme +
  theme(legend.position = "none") +
  scale_x_continuous(
    labels = function(x) {
      paste0(x, "%")
    }
  ) +
  labs(
    title = "Occupational Share among Puerto Rican Workers, 1950",
    subtitle = "Workforce Share by Occupation",
    x = "Share of Workforce",
    y = NULL
  )
```

### Workforce Share by Occupation 1950 to 2010

```{r PR Work share by Occ, message=FALSE, warning=FALSE}
ggplot(occ_cat_data, aes(x = pct_workers, y = occ_cat1, fill = occ_cat1)) +
  geom_col(fill = "#2c3e50", alpha = 0.9) +
  facet_wrap( ~ YEAR, ncol = 3) +
  my_theme +
  theme(legend.position = "none",
        strip.text = element_text(size = 10, face = "bold")) +
  scale_x_continuous(
    labels = function(x) {
      paste0(x, "%")
    }
  ) +
  labs(
    title = "Puerto Rican Workforce Share by Occupation",
    subtitle = "Percentage of total workforce across decennial years",
    x = "Share of Workforce (%)",
    y = NULL
  )
```

### Calculate Growth index

```{r Growth index calculation, message=FALSE,warning=FALSE}
occ_index <- occ_cat_data %>%
  filter(occ_cat1 != "11. Military", occ_cat1 != "Unclassified") %>%
  group_by(occ_cat1) %>%
  arrange(YEAR) %>%
  mutate(index = workers / first(workers) * 100) %>%
  ungroup()

occ_index_us <- occ_cat_data_us %>%
  filter(occ_cat1 != "11. Military", occ_cat1 != "Unclassified") %>%
  group_by(occ_cat1) %>%
  arrange(YEAR) %>%
  mutate(index = workers / first(workers) * 100) %>%
  ungroup()
```

### Visualize Growth Index

#### Puerto Rican

```{r Growth Index visualize, message=FALSE,warning=FALSE}
ggplot(occ_index, aes(x = YEAR, y = index, color = occ_cat1)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1) +
  facet_wrap( ~ occ_cat1, ncol = 4) +
  my_theme +
  theme(legend.position = "none",
        strip.text = element_text(size = 8, face = "bold")) +
  geom_hline(
    yintercept = 100,
    linetype = "dashed",
    color = "black",
    alpha = 0.5
  ) +
  labs(
    title = "Occupational Employment Growth Index",
    subtitle = "First year = 100. Dashed line represents baseline.",
    x = "Year",
    y = "Index"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

#### Total US

```{r Total US Growth Index viz, message=FALSE, warning=FALSE}
ggplot(occ_index_us, aes(x = YEAR, y = index, color = occ_cat1)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1) +
  facet_wrap( ~ occ_cat1, ncol = 4) +
  my_theme +
  theme(legend.position = "none",
        strip.text = element_text(size = 8, face = "bold")) +
  geom_hline(
    yintercept = 100,
    linetype = "dashed",
    color = "black",
    alpha = 0.5
  ) +
  labs(
    title = "Occupational Employment Growth Index",
    subtitle = "First year = 100. Dashed line represents baseline.",
    x = "Year",
    y = "Index"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```


### Over / Under Representation Index

```{r Over/under data prep, message=FALSE, warning=FALSE}
occ_rep_data <- occ_cat_data %>%
  select(YEAR, occ_cat1, pr_pct = pct_workers) %>%
  inner_join(
    occ_cat_data_us %>% select(YEAR, occ_cat1, us_pct = pct_workers),
    by = c("YEAR", "occ_cat1")
  ) %>%
  mutate(rep_index = pr_pct / us_pct) %>%
  filter(occ_cat1 != "Unclassified", occ_cat1 != "11. Military")
```

```{r Over/under vizualization, message=FALSE,warning=FALSE}
ggplot(occ_rep_data, aes(x = YEAR, y = rep_index, color = occ_cat1)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~occ_cat1, ncol = 4, scales = "free_y") +
  geom_hline(
    yintercept = 1,
    linetype = "dashed",
    color = "black",
    linewidth = 0.8
  ) +
  my_theme +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 8, face = "bold")
  ) +
  labs(
    title = "Puerto Rican Worker Representation Index by Occupation",
    subtitle = "Index > 1.0 = Overrepresented vs. US Average | Index < 1.0 = Underrepresented",
    x = "Year",
    y = "Representation Index (PR Share / US Share)"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

### Over / Under Representation Index Adjusted for Small Denominator

```{r Over/under adjusted data prep, message=FALSE,warning=FALSE}
occ_rep_data_national_controlled <- occ_cat_data %>%
  select(YEAR, occ_cat1, pr_pct = pct_workers) %>%
  inner_join(
    occ_cat_data_us %>% select(YEAR, occ_cat1, us_pct = pct_workers),
    by = c("YEAR", "occ_cat1")
  ) %>%
  filter(us_pct >= 1.0) %>%
  mutate(rep_index = pr_pct / us_pct) %>%
  filter(occ_cat1 != "Unclassified", occ_cat1 != "11. Military")
```


### Visualize Over / under adjusted

```{r Visualize Over/under adjusted, message=FALSE,warning=FALSE}
ggplot(
  occ_rep_data_national_controlled,
  aes(x = YEAR, y = rep_index, color = occ_cat1)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~occ_cat1, ncol = 4, scales = "free_y") +
  geom_hline(
    yintercept = 1,
    linetype = "dashed",
    color = "black",
    linewidth = 0.8
  ) +
  my_theme +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 8, face = "bold")
  ) +
  labs(
    title = "Puerto Rican Worker Representation Index by Occupation",
    subtitle = "Compared to National Baseline | Excludes occupations < 1% of total US workforce",
    x = "Year",
    y = "Representation Index (PR Share / Total US Share)"
  )
```


## Occupation Analysis by Sex

### Calculate Occupation metrics by Sex

```{r Occupation by sex data prep, message=FALSE,warning=FALSE}
pr_occ_sex <- data_hist %>%
  filter(HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261,
         AGE >= 16,
         EMPSTAT == 1,
         SEX != 9) %>%
  # 1. SERVER SIDE: Create the text labels using case_when
  mutate(
    sex_label = case_when(SEX == 1 ~ "Male", SEX == 2 ~ "Female"),
    occ_cat1 = case_when(
      (OCC2010 >= 10 &
         OCC2010 <= 960) |
        (OCC2010 >= 2600 &
           OCC2010 <= 2960) ~ "Management, Business, Entertainment & Media",
      OCC2010 >= 1000 & OCC2010 <= 1965 ~ "STEM",
      (OCC2010 >= 2000 &
         OCC2010 <= 2550) |
        (OCC2010 >= 4300 &
           OCC2010 <= 4650) ~ "Education, Community & Social Services",
      OCC2010 >= 3000 & OCC2010 <= 3655 ~ "Health & Medical",
      (OCC2010 >= 3700 &
         OCC2010 <= 3955) |
        (OCC2010 >= 9000 &
           OCC2010 <= 9420) ~ "Law Enforcement & Transportation",
      OCC2010 >= 4000 &
        OCC2010 <= 4250 ~ "Food, Maintenance, Building & Grounds Cleaning",
      OCC2010 >= 4700 & OCC2010 <= 4965 ~ "Sales and Related",
      OCC2010 >= 5000 &
        OCC2010 <= 5940 ~ "Administrative and Office Support",
      OCC2010 >= 6005 &
        OCC2010 <= 7630 ~ "Natural Resources, Construction & Maintenance",
      (OCC2010 >= 7700 &
         OCC2010 <= 8965) |
        (OCC2010 >= 9500 &
           OCC2010 <= 9750) ~ "Production and Material Moving",
      OCC2010 >= 9800 & OCC2010 <= 9830 ~ "Military",
      OCC2010 %in% c(9920, 9999) ~ "Unclassified",
      TRUE ~ "Unclassified"
    )
  ) %>%
  group_by(YEAR, sex_label, occ_cat1) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  # 2. BRING TO LOCAL MEMORY
  collect() %>%
  # 3. LOCAL R SIDE: Apply R-specific functions
  mutate(
    # Catch any missing sectors first
    occ_cat1 = replace_na(occ_cat1, "Unclassified"),
    # Convert to factor to lock in your plotting order
    occ_cat1 = factor(
      occ_cat1,
      levels = c(
        "Management, Business, Entertainment & Media",
        "STEM",
        "Education, Community & Social Services",
        "Health & Medical",
        "Law Enforcement & Transportation",
        "Food, Maintenance, Building & Grounds Cleaning",
        "Sales and Related",
        "Administrative and Office Support",
        "Natural Resources, Construction & Maintenance",
        "Production and Material Moving",
        "Military",
        "Unclassified"
      )
    )
  ) %>%
  # 4. Finish calculations
  group_by(YEAR, sex_label) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100) %>%
  ungroup()

us_occ_sex <- data_hist %>%
  filter(AGE >= 16, EMPSTAT == 1, SEX != 9) %>%
  # 1. SERVER SIDE: Create the text labels using case_when
  mutate(
    sex_label = case_when(SEX == 1 ~ "Male", SEX == 2 ~ "Female"),
    occ_cat1 = case_when(
      (OCC2010 >= 10 &
         OCC2010 <= 960) |
        (OCC2010 >= 2600 &
           OCC2010 <= 2960) ~ "Management, Business, Entertainment & Media",
      OCC2010 >= 1000 & OCC2010 <= 1965 ~ "STEM",
      (OCC2010 >= 2000 &
         OCC2010 <= 2550) |
        (OCC2010 >= 4300 &
           OCC2010 <= 4650) ~ "Education, Community & Social Services",
      OCC2010 >= 3000 & OCC2010 <= 3655 ~ "Health & Medical",
      (OCC2010 >= 3700 &
         OCC2010 <= 3955) |
        (OCC2010 >= 9000 &
           OCC2010 <= 9420) ~ "Law Enforcement & Transportation",
      OCC2010 >= 4000 &
        OCC2010 <= 4250 ~ "Food, Maintenance, Building & Grounds Cleaning",
      OCC2010 >= 4700 & OCC2010 <= 4965 ~ "Sales and Related",
      OCC2010 >= 5000 &
        OCC2010 <= 5940 ~ "Administrative and Office Support",
      OCC2010 >= 6005 &
        OCC2010 <= 7630 ~ "Natural Resources, Construction & Maintenance",
      (OCC2010 >= 7700 &
         OCC2010 <= 8965) |
        (OCC2010 >= 9500 &
           OCC2010 <= 9750) ~ "Production and Material Moving",
      OCC2010 >= 9800 & OCC2010 <= 9830 ~ "Military",
      OCC2010 %in% c(9920, 9999) ~ "Unclassified",
      TRUE ~ "Unclassified"
    )
  ) %>%
  group_by(YEAR, sex_label, occ_cat1) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  # 2. BRING TO LOCAL MEMORY
  collect() %>%
  # 3. LOCAL R SIDE: Apply R-specific functions
  mutate(
    # Catch any missing sectors first
    occ_cat1 = replace_na(occ_cat1, "Unclassified"),
    # Convert to factor to lock in your plotting order
    occ_cat1 = factor(
      occ_cat1,
      levels = c(
        "Management, Business, Entertainment & Media",
        "STEM",
        "Education, Community & Social Services",
        "Health & Medical",
        "Law Enforcement & Transportation",
        "Food, Maintenance, Building & Grounds Cleaning",
        "Sales and Related",
        "Administrative and Office Support",
        "Natural Resources, Construction & Maintenance",
        "Production and Material Moving",
        "Military",
        "Unclassified"
      )
    )
  ) %>%
  # 4. Finish calculations
  group_by(YEAR, sex_label) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100) %>%
  ungroup()
```

### Visualize Puerto Rican workers by Occupation by Sex

```{r}
ggplot(pr_occ_sex, aes(x = YEAR, y = pct_workers, color = sex_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~occ_cat1, ncol = 4) +
  scale_color_manual(values = c("Male" = "#1f78b4", "Female" = "#e31a1c")) +
  my_theme +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 8, face = "bold")
  ) +
  scale_y_continuous(
    labels = function(x) {
      paste0(x, "%")
    }
  ) +
  labs(
    title = "Puerto Rican Workforce Share by Occupation and Sex (1950-2010)",
    subtitle = "Percentage of male vs. female workforce in each category.",
    x = "Year",
    y = "Share of Workforce (%)",
    color = "Sex"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

### Visualize Total workers by Occupation by Sex

```{r}
ggplot(us_occ_sex, aes(x = YEAR, y = pct_workers, color = sex_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~occ_cat1, ncol = 4) +
  scale_color_manual(values = c("Male" = "#1f78b4", "Female" = "#e31a1c")) +
  my_theme +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 8, face = "bold")
  ) +
  scale_y_continuous(
    labels = function(x) {
      paste0(x, "%")
    }
  ) +
  labs(
    title = "Total Workforce Share by Occupation and Sex (1950-2010)",
    subtitle = "Percentage of male vs. female workforce in each category.",
    x = "Year",
    y = "Share of Workforce (%)",
    color = "Sex"
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10))
```

## Occupation analysis by Income

### Calculate Occupation Income

```{r}
pr_income_occ <- data_hist %>%
  filter(
    HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261,
    AGE >= 16,
    EMPSTAT == 1,
    INCWAGE > 0,
    INCWAGE < 999999
  ) %>%
  # 1. SERVER SIDE: Create the text labels using case_when
  mutate(
    occ_cat1 = case_when(
      (OCC2010 >= 10 &
         OCC2010 <= 960) |
        (OCC2010 >= 2600 &
           OCC2010 <= 2960) ~ "Management, Business, Entertainment & Media",
      OCC2010 >= 1000 & OCC2010 <= 1965 ~ "STEM",
      (OCC2010 >= 2000 &
         OCC2010 <= 2550) |
        (OCC2010 >= 4300 &
           OCC2010 <= 4650) ~ "Education, Community & Social Services",
      OCC2010 >= 3000 & OCC2010 <= 3655 ~ "Health & Medical",
      (OCC2010 >= 3700 &
         OCC2010 <= 3955) |
        (OCC2010 >= 9000 &
           OCC2010 <= 9420) ~ "Law Enforcement & Transportation",
      OCC2010 >= 4000 &
        OCC2010 <= 4250 ~ "Food, Maintenance, Building & Grounds Cleaning",
      OCC2010 >= 4700 & OCC2010 <= 4965 ~ "Sales and Related",
      OCC2010 >= 5000 &
        OCC2010 <= 5940 ~ "Administrative and Office Support",
      OCC2010 >= 6005 &
        OCC2010 <= 7630 ~ "Natural Resources, Construction & Maintenance",
      (OCC2010 >= 7700 &
         OCC2010 <= 8965) |
        (OCC2010 >= 9500 &
           OCC2010 <= 9750) ~ "Production and Material Moving",
      OCC2010 >= 9800 & OCC2010 <= 9830 ~ "Military",
      OCC2010 %in% c(9920, 9999) ~ "Unclassified",
      TRUE ~ "Unclassified"
    )
  ) %>%
  group_by(YEAR, occ_cat1) %>%
  summarise(
    avg_wage = sum(INCWAGE * CPI99 * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE),
    total_workers = sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # 2. BRING TO LOCAL MEMORY
  collect() %>%
  # 3. LOCAL R SIDE: Replace NAs and convert to Factor
  mutate(
    occ_cat1 = replace_na(occ_cat1, "Unclassified"),
    occ_cat1 = factor(
      occ_cat1,
      levels = c(
        "Management, Business, Entertainment & Media",
        "STEM",
        "Education, Community & Social Services",
        "Health & Medical",
        "Law Enforcement & Transportation",
        "Food, Maintenance, Building & Grounds Cleaning",
        "Sales and Related",
        "Administrative and Office Support",
        "Natural Resources, Construction & Maintenance",
        "Production and Material Moving",
        "Military",
        "Unclassified"
      )
    )
  )

us_income_occ <- data_hist %>%
  filter(AGE >= 16, EMPSTAT == 1, INCWAGE > 0, INCWAGE < 999999) %>%
  # 1. SERVER SIDE: Create the text labels using case_when
  mutate(
    occ_cat1 = case_when(
      (OCC2010 >= 10 &
         OCC2010 <= 960) |
        (OCC2010 >= 2600 &
           OCC2010 <= 2960) ~ "Management, Business, Entertainment & Media",
      OCC2010 >= 1000 & OCC2010 <= 1965 ~ "STEM",
      (OCC2010 >= 2000 &
         OCC2010 <= 2550) |
        (OCC2010 >= 4300 &
           OCC2010 <= 4650) ~ "Education, Community & Social Services",
      OCC2010 >= 3000 & OCC2010 <= 3655 ~ "Health & Medical",
      (OCC2010 >= 3700 &
         OCC2010 <= 3955) |
        (OCC2010 >= 9000 &
           OCC2010 <= 9420) ~ "Law Enforcement & Transportation",
      OCC2010 >= 4000 &
        OCC2010 <= 4250 ~ "Food, Maintenance, Building & Grounds Cleaning",
      OCC2010 >= 4700 & OCC2010 <= 4965 ~ "Sales and Related",
      OCC2010 >= 5000 &
        OCC2010 <= 5940 ~ "Administrative and Office Support",
      OCC2010 >= 6005 &
        OCC2010 <= 7630 ~ "Natural Resources, Construction & Maintenance",
      (OCC2010 >= 7700 &
         OCC2010 <= 8965) |
        (OCC2010 >= 9500 &
           OCC2010 <= 9750) ~ "Production and Material Moving",
      OCC2010 >= 9800 & OCC2010 <= 9830 ~ "Military",
      OCC2010 %in% c(9920, 9999) ~ "Unclassified",
      TRUE ~ "Unclassified"
    )
  ) %>%
  group_by(YEAR, occ_cat1) %>%
  summarise(
    us_avg_wage = sum(INCWAGE * CPI99 * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE),
    total_workers = sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # 2. BRING TO LOCAL MEMORY
  collect() %>%
  # 3. LOCAL R SIDE: Replace NAs and convert to Factor
  mutate(
    occ_cat1 = replace_na(occ_cat1, "Unclassified"),
    occ_cat1 = factor(
      occ_cat1,
      levels = c(
        "Management, Business, Entertainment & Media",
        "STEM",
        "Education, Community & Social Services",
        "Health & Medical",
        "Law Enforcement & Transportation",
        "Food, Maintenance, Building & Grounds Cleaning",
        "Sales and Related",
        "Administrative and Office Support",
        "Natural Resources, Construction & Maintenance",
        "Production and Material Moving",
        "Military",
        "Unclassified"
      )
    )
  )
```

### Visualize Income trajectory

```{r}
ggplot(pr_income_occ, aes(x = YEAR, y = avg_wage, color = occ_cat1)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~occ_cat1, ncol = 4) +
  my_theme +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 8, face = "bold")
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
  labs(
    title = "Average Annual Wage for Puerto Rican Workers by Occupation",
    subtitle = "Nominal Dollars.",
    x = "Year",
    y = "Average Wage ($)"
  )
```

### Best paying occupations for Puerto Ricans

```{r}
pr_income_2024 <- pr_income_occ %>%
  filter(YEAR == 2024, occ_cat1 != "Unclassified")

ggplot(pr_income_2024, aes(x = avg_wage, y = reorder(occ_cat1, avg_wage))) +
  geom_col(fill = "#1b9e77", alpha = 0.9) +
  my_theme +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Highest Paying Occupations for Puerto Rican Workers (2024)",
    subtitle = "Average Annual Wage by Category",
    x = "Average Annual Wage ($)",
    y = NULL
  )
```

### Best paying occupations for total pop

```{r}
us_income_2024 <- us_income_occ %>%
  filter(YEAR == 2024, occ_cat1 != "Unclassified")

ggplot(us_income_2024, aes(x = us_avg_wage, y = reorder(occ_cat1, us_avg_wage))) +
  geom_col(fill = "#1b9e77", alpha = 0.9) +
  my_theme +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Highest Paying Occupations for Total Pop Workers (2024)",
    subtitle = "Average Annual Wage by Category",
    x = "Average Annual Wage ($)",
    y = NULL
  )
```


### Calculate wage gap by occupation for Puerto Ricans

```{r}
wage_gap_occ <- pr_income_occ %>%
  select(YEAR, occ_cat1, pr_avg_wage = avg_wage) %>%
  inner_join(us_income_occ %>% select(YEAR, occ_cat1, us_avg_wage),
    by = c("YEAR", "occ_cat1")
  ) %>%
  mutate(
    wage_ratio = pr_avg_wage / us_avg_wage,
    wage_gap_pct = (wage_ratio - 1) * 100
  ) %>%
  filter(occ_cat1 != "Unclassified", occ_cat1 != "11. Military")
```

### Visualize wage gap for Puuerto Ricans

```{r}
ggplot(wage_gap_occ, aes(x = YEAR, y = wage_ratio, color = occ_cat1)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~occ_cat1, ncol = 4) +
  geom_hline(
    yintercept = 1.0,
    linetype = "dashed",
    color = "black",
    linewidth = 0.8
  ) +
  my_theme +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 8, face = "bold")
  ) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
  labs(
    title = "Occupational Wage Gap for Puerto Rican Workers (1950-2024)",
    subtitle = "Ratio of PR Average Wage to US Average Wage. 1.0 = Wage Parity.",
    x = "Year",
    y = "Wage Ratio (PR / US)"
  )
```

### Wage Tier Analysis:Are Puerto Ricans gaining upward mobility

#### Wage tier data prep

```{r}
occupation_tiers <- us_income_occ %>%
  group_by(YEAR) %>%
  mutate(
    tier_num = ntile(us_avg_wage, 3),
    wage_tier = case_when(
      tier_num == 3 ~ "High-Paying Occupations",
      tier_num == 2 ~ "Middle-Paying Occupations",
      tier_num == 1 ~ "Low-Paying Occupations"
    ),
    wage_tier = factor(
      wage_tier,
      levels = c(
        "High-Paying Occupations",
        "Middle-Paying Occupations",
        "Low-Paying Occupations"
      )
    )
  ) %>%
  select(YEAR, occ_cat1, wage_tier) %>%
  ungroup()

pr_tier_trends <- occ_cat_data %>%
  inner_join(occupation_tiers, by = c("YEAR", "occ_cat1")) %>%
  group_by(YEAR, wage_tier) %>%
  summarise(
    tier_workers = sum(workers, na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  mutate(
    pct_in_tier = (tier_workers / sum(tier_workers)) * 100,
    demographic = "Puerto Rican Workers"
  ) %>%
  ungroup()

us_tier_trends <- occ_cat_data_us %>%
  inner_join(occupation_tiers, by = c("YEAR", "occ_cat1")) %>%
  group_by(YEAR, wage_tier) %>%
  summarise(
    tier_workers = sum(workers, na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  mutate(
    pct_in_tier = (tier_workers / sum(tier_workers)) * 100,
    demographic = "Total US Workers"
  ) %>%
  ungroup()

combined_tier_trends <- bind_rows(pr_tier_trends, us_tier_trends)
```

### Wage tier Visualization

```{r}
ggplot(
  combined_tier_trends,
  aes(x = YEAR, y = pct_in_tier, color = demographic)
) +
  geom_line(linewidth = 0.75) +
  geom_point(size = 2) +
  facet_wrap(~wage_tier, ncol = 3) +
  scale_color_manual(values = c(
    "Puerto Rican Workers" = "#d95f02",
    "Total US Workers" = "#7570b3"
  )) +
  my_theme +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 10, face = "bold")
  ) +
  scale_y_continuous(
    labels = function(x) {
      paste0(x, "%")
    }
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
  labs(
    title = "Workforce Distribution Across Occupational Wage Tiers (1950-2024)",
    subtitle = "Comparing PR workforce composition against the US baseline",
    x = "Year",
    y = "Share of Workforce (%)",
    color = NULL
  )
```

### Share of Puerto Rican workers by Wage Tiers

```{r}
wage_tier_pct_pr_occ <- occupation_tiers %>%
  inner_join(occ_cat_data %>% select(YEAR, occ_cat1, pr_pct = pct_workers),
    by = c("YEAR", "occ_cat1")
  )

pr_tier_summary <- wage_tier_pct_pr_occ %>%
  group_by(YEAR, wage_tier) %>%
  summarise(
    total_tier_pct = sum(pr_pct, na.rm = TRUE),
    .groups = "drop"
  )
```

### Visualize Share of Puerto Rican Workers by Wage Tiers

```{r}
ggplot(
  pr_tier_summary,
  aes(x = YEAR, y = total_tier_pct, fill = wage_tier)
) +
  geom_area(
    alpha = 0.85,
    color = "white",
    linewidth = 0.4
  ) +
  scale_fill_manual(
    values = c(
      "High-Paying Occupations" = "#1b9e77",
      "Middle-Paying Occupations" = "#d95f02",
      "Low-Paying Occupations" = "#7570b3"
    )
  ) +
  my_theme +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  scale_y_continuous(
    labels = function(x) {
      paste0(x, "%")
    }
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
  labs(
    title = "Puerto Rican Workforce Distribution by Wage Tier",
    subtitle = "Percentage of workers in High, Middle, and Low paying occupations",
    x = "Year",
    y = "Share of Workforce (%)"
  )
```

```{r}
ggplot(
  pr_tier_summary,
  aes(x = YEAR, y = total_tier_pct, color = wage_tier)
) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = c(
      "High-Paying Occupations" = "#1b9e77",
      "Middle-Paying Occupations" = "#d95f02",
      "Low-Paying Occupations" = "#7570b3"
    )
  ) +
  my_theme +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  scale_y_continuous(
    labels = function(x) {
      paste0(x, "%")
    }
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
  labs(
    title = "Puerto Rican Workforce Trajectory by Wage Tier",
    subtitle = "Tracking the share of workers in each tier over time",
    x = "Year",
    y = "Share of Workforce (%)"
  )
```
### Change in Puerto Rican Worker Share and Wage Tier

```{r}
ggplot(wage_tier_pct_pr_occ, aes(x = YEAR, y = pr_pct)) +
  geom_line(color = "black", linewidth = 0.75) +
  geom_point(aes(color = wage_tier), size = 2) +
  facet_wrap(~occ_cat1, ncol = 4) +
  scale_color_manual(
    values = c(
      "High-Paying Occupations" = "#1b9e77",
      "Middle-Paying Occupations" = "#d95f02",
      "Low-Paying Occupations" = "#7570b3"
    )
  ) +
  my_theme +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    strip.text = element_text(size = 8, face = "bold")
  ) +
  scale_y_continuous(
    labels = function(x) {
      paste0(x, "%")
    }
  ) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
  labs(
    title = "Puerto Rican Workforce Share and Occupational Wage Tiers",
    subtitle = "Tracking occupational participation. Dot colors indicate the national wage tier for that decade.",
    x = "Year",
    y = "Share of Puerto Rican Workforce (%)"
  )
```

```{r}
wage_tier_pct_pr_occ %>%
  filter(occ_cat1 != "Unclassified") %>%
  ggplot(aes(x = pr_pct, y = occ_cat1, fill = wage_tier)) +
  geom_col() +
  facet_wrap(~YEAR, ncol = 3) +
  scale_fill_manual(
    values = c(
      "High-Paying Occupations" = "#1b9e77",
      "Middle-Paying Occupations" = "#d95f02",
      "Low-Paying Occupations" = "#7570b3"
    )
  ) +
  my_theme +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 7)
  ) +
  scale_x_continuous(
    labels = function(x) {
      paste0(x, "%")
    }
  ) +
  labs(
    title = "Puerto Rican Workforce Share and Occupational Wage Tiers",
    subtitle = "Tracking occupational participation. Bar colors indicate the national wage tier for that decade.",
    x = "Share of Puerto Rican Workforce (%)",
    y = NULL
  )
```

## AI Exposed Occupations

