---
title: "011_AI_vulnerable_occupaitons"
author: "Jorge R. Soldevila Irizarry"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 999)
selected_years <- c(2005, 2010, 2015, 2020)

# Source central setup file
source("00_project_setup.R")
```

## AI Exposed Occupations Analysis

### Map industry codes and AI vulnerable Occupations

```{r Mapping Industry and Occupaiton Categories, message=FALSE,warning=FALSE}
mapped_workers_db <- data_acs %>%
  # Apply base employment filter, correcting the missing GQ != 3
  filter(AGE >= 16, EMPSTAT == 1, GQ != 3) %>%
  mutate(
    # Create a boolean flag for PR status to split the data later
    is_pr = case_when(
      HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261 ~ TRUE,
      TRUE ~ FALSE
    ),
    
    # Map Industries
    ind_sector = case_when(
    IND1990 >= 10 & IND1990 <= 32 ~ "01. Agriculture, Forestry, Fishing & Hunting",
    IND1990 >= 40 & IND1990 <= 50 ~ "02. Mining, Quarrying, & Oil/Gas Extraction",
    IND1990 >= 450 & IND1990 <= 472 ~ "03. Utilities",
    IND1990 == 60 ~ "04. Construction",
    IND1990 >= 100 & IND1990 <= 392 ~ "05. Manufacturing",
    IND1990 >= 500 & IND1990 <= 571 ~ "06. Wholesale Trade",

    # Excludes 641 (Eating/Drinking Places) to align with modern NAICS grouping
    (IND1990 >= 580 & IND1990 <= 691) & IND1990 != 641 ~ "07. Retail Trade",
    IND1990 >= 400 & IND1990 <= 432 ~ "08. Transportation & Warehousing",
    IND1990 >= 440 & IND1990 <= 442 ~ "09. Information", # Maps 1990 Communications here
    IND1990 >= 700 & IND1990 <= 711 ~ "10. Finance & Insurance",
    IND1990 == 712 ~ "11. Real Estate & Rental/Leasing",

    # Merges 1990 Engineering/Architecture/Accounting with related professional codes
    IND1990 == 841 | (IND1990 >= 881 & IND1990 <= 893) ~ "12. Professional, Scientific, & Tech Services",
    IND1990 >= 721 & IND1990 <= 741 ~ "13. Administrative, Support, & Waste Mgmt",
    IND1990 >= 842 & IND1990 <= 860 ~ "14. Educational Services",
    (IND1990 >= 812 & IND1990 <= 840) | (IND1990 >= 861 & IND1990 <= 873) ~ "15. Health Care & Social Assistance",
    IND1990 >= 800 & IND1990 <= 810 ~ "16. Arts, Entertainment, & Recreation",

    # Combines 1990 Eating/Drinking places (641) with Hotels/Lodging (762, 770)
    IND1990 %in% c(641, 762, 770) ~ "17. Accommodation & Food Services",
    IND1990 %in% c(750, 751, 752, 761, 771, 772, 781, 791) ~ "18. Other Services (Except Public Admin)",
    IND1990 >= 900 & IND1990 <= 932 ~ "19. Public Administration",
    IND1990 >= 940 & IND1990 <= 960 ~ "20. Military",
      TRUE ~ "21. Unclassified / Not in Labor Force"
    ),
    
    # Map Occupations
    # Map Occupations
    occ_cat1 = case_when(
      (OCC2010 >= 10 &
         OCC2010 <= 960) |
        (OCC2010 >= 2600 &
           OCC2010 <= 2960) ~ "Management, Business, Entertainment & Media",
      OCC2010 >= 1000 & OCC2010 <= 1965 ~ "STEM",
      (OCC2010 >= 2000 &
         OCC2010 <= 2550) |
        (OCC2010 >= 4300 &
           OCC2010 <= 4650) ~ "Education, Community & Social Services",
      OCC2010 >= 3000 & OCC2010 <= 3655 ~ "Health & Medical",
      (OCC2010 >= 3700 &
         OCC2010 <= 3955) |
        (OCC2010 >= 9000 &
           OCC2010 <= 9420) ~ "Law Enforcement & Transportation",
      OCC2010 >= 4000 &
        OCC2010 <= 4250 ~ "Food, Maintenance, Building & Grounds Cleaning",
      OCC2010 >= 4700 & OCC2010 <= 4965 ~ "Sales and Related",
      OCC2010 >= 5000 &
        OCC2010 <= 5940 ~ "Administrative and Office Support",
      OCC2010 >= 6005 &
        OCC2010 <= 7630 ~ "Natural Resources, Construction & Maintenance",
      (OCC2010 >= 7700 &
         OCC2010 <= 8965) |
        (OCC2010 >= 9500 &
           OCC2010 <= 9750) ~ "Production and Material Moving",
      OCC2010 >= 9800 & OCC2010 <= 9830 ~ "Military",
      OCC2010 %in% c(9920, 9999) ~ "Unclassified",
      TRUE ~ "Unclassified"
    ),
    # Map AI vulnerable occupaitons 
    ai_vuln = case_when(
  OCC2010 %in% c(10, 20, 110, 120, 800, 1010, 1020, 3255, 4850, 5240, 5400, 5700, 7340) ~ "AI Vulnerable",
  OCC2010 %in% c(9920, 9999) ~ "Unclassified",
  TRUE ~ "Not AI Vulnerable"
  )
)
```

## Puerto Rican concentration by specific occupation

```{r Top 10 occupations for PR, message = FALSE, warning = FALSE}
top_10_pr_occ_by_year <- mapped_workers_db %>%
  # 1. ADD THIS FILTER to isolate Puerto Rican workers
  filter(is_pr == TRUE) %>%
  
  # 2. Group by Year, Occupation Code, AND ai_vuln so it doesn't get dropped
  group_by(YEAR, OCC2010, ai_vuln) %>%
  
  # 3. Sum the person weights to get the estimated total workers
  summarise(
    total_estimated_workers = sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  # 4. Bring the summary into local memory
  collect() %>%
  
  # (Notice the redundant ai_vuln mutate is gone!)
  
  # 5. Group LOCALLY by Year so our top 10 function applies to each year individually
  group_by(YEAR) %>%
  
  # 6. Extract the top 10 occupations with the highest worker count for each year
  slice_max(order_by = total_estimated_workers, n = 10) %>%
  
  # 7. Ungroup and sort cleanly so it's easy to read
  ungroup() %>%
  arrange(YEAR, desc(total_estimated_workers))

```

### Add labels to occupation categories

```{r Add labeles to Occupation, message=FALSE,warning=FALSE}
ddi <- read_ipums_ddi("./usa_00121.xml")
occ_labels <- ipums_val_labels(ddi, OCC2010)

# Join the text labels to your top 10 list
top_10_pr_occ_labeled <- top_10_pr_occ_by_year %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  select(YEAR,OCC2010, occupation_name = lbl, total_estimated_workers, ai_vuln)

print(top_10_pr_occ_labeled)
```
### Visualize Top 10 occupations by year

```{r Plot top 10 occupations in timeseries, message=FALSE,warning=FALSE}
# Create the faceted time series plot
ggplot(top_10_pr_occ_labeled, aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name # Ensures lines connect properly
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Create a separate panel for each occupation
  # scales = "free_y" allows each plot to zoom in on its specific trend
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Apply distinct colors based on AI vulnerability
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      # Bold Orange
    "Not AI Vulnerable" = "#1f78b4",  # Cool Blue
    "Unclassified" = "grey70"         # Neutral Grey
  )) +
  
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    
    # Make the facet headers smaller and bold so long occupation names fit
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) # Tilts years to prevent overlap
  ) +
  
  scale_y_continuous(labels = label_comma()) +
  
  labs(
    title = "Top Occupations for Puerto Rican Workers Over Time",
    subtitle = "Estimated workforce volume by individual occupation",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```


```{r Top 10 occupations for PR visualization, message=FALSE, warning=FALSE}
ggplot(top_10_pr_occ_labeled, aes(
  x = total_estimated_workers, 
  # reorder_within sorts the bars largest-to-smallest INSIDE each facet
  y = tidytext::reorder_within(OCC2010, total_estimated_workers, YEAR), 
  fill = ai_vuln
)) +
  geom_col(alpha = 0.9) +
  
  # scales = "free_y" is crucial here so only that year's top 10 show in the panel
  facet_wrap(~YEAR, ncol = 4, scales = "free_y") +
  
  # This cleans up the Y-axis text that reorder_within generates
  tidytext::scale_y_reordered() +
  
  # Custom colors for vulnerability
  scale_fill_manual(values = c(
    "AI Vulnerable" = "#d95f02",      # Warning orange
    "Not AI Vulnerable" = "#2c3e50",  # Your standard navy blue
    "Unclassified" = "grey70"         # Neutral grey
  )) +
  my_theme +
  theme(
    # We bring the legend back so the AI colors make sense
    legend.position = "bottom",
    strip.text = element_text(size = 10, face = "bold"),
    # Shrink the y-axis text slightly so long occupation names fit
    axis.text.y = element_text(size = 7) 
  ) +
  scale_x_continuous(labels = scales::comma_format()) +
  labs(
    title = "Top 10 Occupations for Puerto Rican Workers by Year",
    subtitle = "Ranked by estimated workforce volume and colored by AI exposure",
    x = "Estimated Total Workers",
    y = NULL, 
    fill = "AI Vulnerability Risk" 
  )
```

```{r Pivot top 10 occupations wide, message=FALSE,warning=FALSE}
top_10_pr_occ_wide <- top_10_pr_occ_labeled %>%
  pivot_wider(names_from = "YEAR",values_from = total_estimated_workers)

```

### Top Occupations by sex

```{r Determine Occupations by sex, message=FALSE, warning=FALSE}

top_10_pr_occ_by_sex <- mapped_workers_db %>%
  # 1. Isolate Puerto Rican workers AND remove unknown sex codes (SEX == 9)
  filter(is_pr == TRUE, SEX != 9) %>%
  
  # 2. Create a clean label for the sexes
  mutate(sex_label = case_when(
    SEX == 1 ~ "Male",
    SEX == 2 ~ "Female"
  )) %>%
  
  # 3. Add sex_label to the grouping before summarizing
  group_by(YEAR, sex_label, OCC2010, ai_vuln) %>%
  
  # 4. Sum the person weights
  summarise(
    total_estimated_workers = sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  # 5. Bring into local memory
  collect() %>%
  
  # 6. Group LOCALLY by both Year AND Sex
  group_by(YEAR, sex_label) %>%
  
  # 7. Extract the top 10 for each specific demographic group
  slice_max(order_by = total_estimated_workers, n = 10) %>%
  
  # 8. Clean up and sort
  ungroup() %>%
  arrange(YEAR, sex_label, desc(total_estimated_workers))

# Join the text labels so it is readable
top_10_pr_occ_sex_labeled <- top_10_pr_occ_by_sex %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  select(YEAR, sex_label, OCC2010, occupation_name = lbl, 
         total_estimated_workers, ai_vuln)

```

### Visualize top 10 occupations for Males

```{r}
# Create the faceted time series plot
ggplot(top_10_pr_occ_sex_labeled %>% filter(sex_label == "Male"), aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name # Ensures lines connect properly
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Create a separate panel for each occupation
  # scales = "free_y" allows each plot to zoom in on its specific trend
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Apply distinct colors based on AI vulnerability
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      # Bold Orange
    "Not AI Vulnerable" = "#1f78b4",  # Cool Blue
    "Unclassified" = "grey70"         # Neutral Grey
  )) +
  
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    
    # Make the facet headers smaller and bold so long occupation names fit
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) # Tilts years to prevent overlap
  ) +
  
  scale_y_continuous(labels = label_comma()) +
  
  labs(
    title = "Top Occupations for Puerto Rican Male Workers Over Time",
    subtitle = "Estimated workforce volume by individual occupation",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```

### Visualize top 10 occupations for Females

```{r}
# Create the faceted time series plot
ggplot(top_10_pr_occ_sex_labeled %>% filter(sex_label == "Female"), aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name # Ensures lines connect properly
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Create a separate panel for each occupation
  # scales = "free_y" allows each plot to zoom in on its specific trend
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Apply distinct colors based on AI vulnerability
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      # Bold Orange
    "Not AI Vulnerable" = "#1f78b4",  # Cool Blue
    "Unclassified" = "grey70"         # Neutral Grey
  )) +
  
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    
    # Make the facet headers smaller and bold so long occupation names fit
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) # Tilts years to prevent overlap
  ) +
  
  scale_y_continuous(labels = label_comma()) +
  
  labs(
    title = "Top Occupations for Puerto Rican Female Workers Over Time",
    subtitle = "Estimated workforce volume by individual occupation",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```
### Continous visualization of Male and Female Top 10 Occupations

```{r Continuos top 10 occupations PR Males, message=FALSE, warning=FALSE}
# 1. Identify which occupations ever made the Top 10 for males
top_male_occs <- top_10_pr_occ_sex_labeled %>%
  filter(sex_label == "Male") %>%
  pull(OCC2010) %>%
  unique()

# 2. Pull the continuous history for ONLY those specific occupations
continuous_male_trends <- mapped_workers_db %>%
  filter(is_pr == TRUE, SEX == 1, OCC2010 %in% top_male_occs) %>%
  group_by(YEAR, OCC2010, ai_vuln) %>%
  summarise(total_estimated_workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  select(YEAR, OCC2010, occupation_name = lbl, ai_vuln, total_estimated_workers) %>%
   mutate(occupation_name = str_wrap(occupation_name, width = 35))

# 3. Plot the continuous data
ggplot(continuous_male_trends, aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name 
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      
    "Not AI Vulnerable" = "#1f78b4",  
    "Unclassified" = "grey70"         
  )) +
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) 
  ) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Top Occupations for Puerto Rican Male Workers Over Time",
    subtitle = "Continuous workforce volume for any occupation that ever reached the Top 10",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```


```{r Continuos top 10 occupations PR Males, message=FALSE, warning=FALSE}
# 1. Identify which occupations ever made the Top 10 for females
top_female_occs <- top_10_pr_occ_sex_labeled %>%
  filter(sex_label == "Female") %>%
  pull(OCC2010) %>%
  unique()

# 2. Pull the continuous history for ONLY those specific occupations
continuous_female_trends <- mapped_workers_db %>%
  # Note: SEX == 2 filters specifically for female workers in the database
  filter(is_pr == TRUE, SEX == 2, OCC2010 %in% top_female_occs) %>%
  group_by(YEAR, OCC2010, ai_vuln) %>%
  summarise(total_estimated_workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  select(YEAR, OCC2010, occupation_name = lbl, ai_vuln, total_estimated_workers) %>%
   mutate(occupation_name = str_wrap(occupation_name, width = 35))

# 3. Plot the continuous data for females
ggplot(continuous_female_trends, aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name 
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      
    "Not AI Vulnerable" = "#1f78b4",  
    "Unclassified" = "grey70"         
  )) +
  
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) 
  ) +
  
  scale_y_continuous(labels = scales::label_comma()) +
  
  labs(
    title = "Top Occupations for Puerto Rican Female Workers Over Time",
    subtitle = "Continuous workforce volume for any occupation that ever reached the Top 10",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```

### Exposure over time (2001 to 2024)

```{r Calculate and visualize exposure, message=FALSE, warning=FALSE}

ai_exposure_data <- mapped_workers_db %>%
  filter(ai_vuln != "Unclassified") %>%
  group_by(YEAR, is_pr, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  group_by(YEAR, is_pr) %>%
  mutate(
    pct_workers = (workers / sum(workers)) * 100,
    demographic = ifelse(is_pr, "Puerto Rican Workers", "Total US Workers")
  ) %>%
  ungroup()
```

```{r }
ggplot(ai_exposure_data %>% filter(ai_vuln == "AI Vulnerable"), 
       aes(x = YEAR, y = pct_workers, color = demographic)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  my_theme +
  scale_color_manual(values = c("Puerto Rican Workers" = "#d95f02", "Total US Workers" = "#7570b3")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Workforce Exposure to AI Technologies (2001-2024)",
    subtitle = "Percentage of workers employed in AI-vulnerable occupations",
    x = "Year",
    y = "Share of Workforce (%)"
  ) +
  scale_x_continuous(breaks = seq(2001, 2024, by = 5))
```

### Wage Premium or Penalty for AI vulnerable occupations

```{r Wage preiums calculation and visualization, message=FALSE, warning=FALSE}
ai_wage_data <- mapped_workers_db %>%
  filter(ai_vuln != "Unclassified", INCWAGE > 0, INCWAGE < 999999) %>%
  group_by(YEAR, is_pr, ai_vuln) %>%
  summarise(
    avg_wage = sum(INCWAGE * CPI99 * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  collect() %>%
  mutate(demographic = ifelse(is_pr, "Puerto Rican Workers", "Total US Workers"))

ggplot(ai_wage_data, aes(x = YEAR, y = avg_wage, color = demographic, linetype = ai_vuln)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  my_theme +
  scale_color_manual(values = c("Puerto Rican Workers" = "#d95f02", "Total US Workers" = "#7570b3")) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Historical Wage Trajectories by AI Vulnerability",
    subtitle = "Real 1999 Dollars",
    x = "Year",
    y = "Average Annual Wage ($)"
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5))
```

### Demographic variations in AI vulnerable occupations

```{r Demographic difference calculation and visualization, message=FALSE, warning=FALSE}
ai_demographics <- mapped_workers_db %>%
  filter(is_pr == TRUE, ai_vuln != "Unclassified", SEX != 9) %>%
  mutate(sex_label = ifelse(SEX == 1, "Male", "Female")) %>%
  group_by(YEAR, sex_label, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  group_by(YEAR, sex_label) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100) %>%
  ungroup()

ggplot(ai_demographics %>% filter(ai_vuln == "AI Vulnerable"), 
       aes(x = YEAR, y = pct_workers, color = sex_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  my_theme +
  scale_color_manual(values = c("Male" = "#1f78b4", "Female" = "#e31a1c")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Puerto Rican AI Vulnerability by Sex",
    subtitle = "Percentage of each sex's workforce in AI-vulnerable occupations",
    x = "Year",
    y = "Exposure Rate (%)"
  )
```

### Demographic variations in AI vulnerable occupations for total population
```{r}
ai_demographics_us <- mapped_workers_db %>%
  filter(ai_vuln != "Unclassified", SEX != 9) %>%
  mutate(sex_label = ifelse(SEX == 1, "Male", "Female")) %>%
  group_by(YEAR, sex_label, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  group_by(YEAR, sex_label) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100) %>%
  ungroup()

ggplot(ai_demographics %>% filter(ai_vuln == "AI Vulnerable"), 
       aes(x = YEAR, y = pct_workers, color = sex_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  my_theme +
  scale_color_manual(values = c("Male" = "#1f78b4", "Female" = "#e31a1c")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "AI Vulnerability by Sex",
    subtitle = "Percentage of each sex's workforce in AI-vulnerable occupations",
    x = "Year",
    y = "Exposure Rate (%)"
  )
```

## AI Vulnerable Occupations Puerto Rican trends

```{r Calculate Puerto Ricans in AI Vulnerable Occ, message=FALSE,warning=FALSE}

ai_exposure_by_occ <- mapped_workers_db %>%
  filter(ai_vuln == "AI Vulnerable") %>%
  group_by(YEAR, is_pr, OCC2010) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val"))%>%
  group_by(YEAR,is_pr) %>%
  mutate(
    pct_workers = (workers / sum(workers)) * 100,
    demographic = ifelse(is_pr, "Puerto Rican Workers", "Total US Workers")
  ) %>%
  ungroup()

```

```{r}
# 1. Clean the labels before plotting
ai_exposure_plot_data <- ai_exposure_by_occ %>%
  # Rename 'lbl' (from the left_join) so it's easier to reference
  rename(occupation_name = lbl) %>%
  # Wrap long IPUMS occupation text so it stacks neatly in the facet headers
  mutate(occupation_name = str_wrap(occupation_name, width = 35))

# 2. Plot the comparison trends
ggplot(ai_exposure_plot_data, aes(
  x = YEAR, 
  y = pct_workers, 
  color = demographic,
  group = demographic
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Facet by occupation to give each job its own comparison panel
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Distinct colors for the demographics
  scale_color_manual(values = c(
    "Puerto Rican Workers" = "#d95f02", # Orange for emphasis
    "Total US Workers" = "#7570b3"      # Cooler purple/blue for baseline
  )) +
  
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(), # Hides "demographic" as it's self-explanatory
    legend.text = element_text(size = 10, face = "bold"),
    
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  ) +
  
  # Format Y-axis as percentages
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  
  labs(
    title = "Concentration of AI-Vulnerable Occupations",
    subtitle = "Comparing the distribution of exposed Puerto Rican workers vs. the Total US baseline",
    x = "Year",
    y = "Share of AI-Vulnerable Workforce (%)"
  )
```

### Average Wage by Occupaitons vulnerable to AI over time (2001-2024)

```{r Calculate Average Wage per Occupations, message=FALSE,warning=FALSE}
ai_wage_data_by_occ <- mapped_workers_db %>%
  filter(ai_vuln == "AI Vulnerable", INCWAGE > 0, INCWAGE < 999999) %>%
  group_by(YEAR, is_pr, OCC2010) %>%
  summarise(
    avg_wage = sum(INCWAGE * CPI99 * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val"))%>%
  mutate(demographic = ifelse(is_pr, "Puerto Rican Workers", "Total US Workers"))
```

### Visualize Average Wage over time for exposed Occupations

```{r Visualize Average Wage for AI Vulnerable Occ, message=FALSE,warning=FALSE}

# 1. Clean the labels before plotting
ai_wage_data_by_occ <- ai_wage_data_by_occ %>%
  # Rename 'lbl' (from the left_join) so it's easier to reference
  rename(occupation_name = lbl) %>%
  # Wrap long IPUMS occupation text so it stacks neatly in the facet headers
  mutate(occupation_name = str_wrap(occupation_name, width = 35))

# 2. Plot the comparison trends
ggplot(ai_wage_data_by_occ, aes(
  x = YEAR, 
  y = avg_wage, 
  color = demographic,
  group = demographic
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Facet by occupation to give each job its own comparison panel
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Distinct colors for the demographics
  scale_color_manual(values = c(
    "Puerto Rican Workers" = "#d95f02", # Orange for emphasis
    "Total US Workers" = "#7570b3"      # Cooler purple/blue for baseline
  )) +
  
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(), # Hides "demographic" as it's self-explanatory
    legend.text = element_text(size = 10, face = "bold"),
    
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  ) +
  
  # Format Y-axis as percentages
  scale_y_continuous(labels = scales::dollar_format()) +
  
  labs(
    title = "Wage Trajectory for AI Vulnerable Occupatoins",
    subtitle = "Real 1999 Dollars",
    x = "Year",
    y = "Average Annual Wage ($)"
  )
```














