---
title: "011_AI_vulnerable_occupaitons"
author: "Jorge R. Soldevila Irizarry"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 999)
selected_years <- c(2005, 2010, 2015, 2020)
ddi <- read_ipums_ddi("./usa_00121.xml")
occ_labels <- ipums_val_labels(ddi, OCC2010)

# Source central setup file
source("00_project_setup.R")
```

## AI Exposed Occupations Analysis

### Map industry codes and AI vulnerable Occupations

```{r Mapping Industry and Occupaiton Categories, message=FALSE,warning=FALSE}
mapped_workers_db <- data_acs %>%
  # Apply base employment filter, correcting the missing GQ != 3
  filter(AGE >= 16, EMPSTAT == 1, GQ != 3) %>%
  mutate(
    # Create a boolean flag for PR status to split the data later
    is_pr = case_when(
      HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261 ~ TRUE,
      TRUE ~ FALSE
    ),
    
# Map Industries
    ind_sector = case_when(
      IND1990 >= 10 & IND1990 <= 32 ~ "Ag, Forestry, Fish & Hunt",
      IND1990 >= 40 & IND1990 <= 50 ~ "Mining, Quarry, Oil/Gas",
      IND1990 >= 450 & IND1990 <= 472 ~ "Utilities",
      IND1990 == 60 ~ "Construction",
      IND1990 >= 100 & IND1990 <= 392 ~ "Manufacturing",
      IND1990 >= 500 & IND1990 <= 571 ~ "Wholesale Trade",
      
      # Excludes 641 (Eating/Drinking Places) to align with modern NAICS grouping
      (IND1990 >= 580 &
         IND1990 <= 691) & IND1990 != 641 ~ "Retail Trade",
      IND1990 >= 400 &
        IND1990 <= 432 ~ "Transport & Warehousing",
      IND1990 >= 440 &
        IND1990 <= 442 ~ "Information",
      # Maps 1990 Communications here
      IND1990 >= 700 & IND1990 <= 711 ~ "Finance & Insurance",
      IND1990 == 712 ~ "Real Estate & Leasing",
      
      # Merges 1990 Engineering/Architecture/Accounting with related professional codes
      IND1990 == 841 |
        (IND1990 >= 881 & IND1990 <= 893) ~ "Prof, Sci, & Tech Svcs",
      IND1990 >= 721 &
        IND1990 <= 741 ~ "Admin, Support, & Waste Mgmt",
      IND1990 >= 842 & IND1990 <= 860 ~ "Educational Svcs",
      (IND1990 >= 812 &
         IND1990 <= 840) |
        (IND1990 >= 861 & IND1990 <= 873) ~ "Health & Social Asst",
      IND1990 >= 800 &
        IND1990 <= 810 ~ "Arts, Entertain, & Rec",
      
      # Combines 1990 Eating/Drinking places (641) with Hotels/Lodging (762, 770)
      IND1990 %in% c(641, 762, 770) ~ "Accom & Food Svcs",
      IND1990 %in% c(750, 751, 752, 761, 771, 772, 
                     781, 791) ~ "Other Svcs (excl. Pub Admin)",
      IND1990 >= 900 & IND1990 <= 932 ~ "Public Admin",
      IND1990 >= 940 & IND1990 <= 960 ~ "Military",
      TRUE ~ "Unclassified / NILF"
    ),
    
    # Map Occupations
    occ_cat1 = case_when(
      (OCC2010 >= 10 &
         OCC2010 <= 960) |
        (OCC2010 >= 2600 &
           OCC2010 <= 2960) ~ "Mgmt, Bus, Entertain & Media",
      OCC2010 >= 1000 & OCC2010 <= 1965 ~ "STEM",
      (OCC2010 >= 2000 &
         OCC2010 <= 2550) |
        (OCC2010 >= 4300 &
           OCC2010 <= 4650) ~ "Educ, Comm & Social Svcs",
      OCC2010 >= 3000 & OCC2010 <= 3655 ~ "Health & Medical",
      (OCC2010 >= 3700 &
         OCC2010 <= 3955) |
        (OCC2010 >= 9000 &
           OCC2010 <= 9420) ~ "Law Enf & Transport",
      OCC2010 >= 4000 &
        OCC2010 <= 4250 ~ "Food, Maint, Bldg & Grounds",
      OCC2010 >= 4700 & OCC2010 <= 4965 ~ "Sales & Related",
      OCC2010 >= 5000 &
        OCC2010 <= 5940 ~ "Admin & Office Support",
      OCC2010 >= 6005 &
        OCC2010 <= 7630 ~ "Nat Res, Const & Maint",
      (OCC2010 >= 7700 &
         OCC2010 <= 8965) |
        (OCC2010 >= 9500 &
           OCC2010 <= 9750) ~ "Prod & Material Moving",
      OCC2010 >= 9800 & OCC2010 <= 9830 ~ "Military",
      OCC2010 %in% c(9920, 9999) ~ "Unclassified",
      TRUE ~ "Unclassified"
    ),
    # Map AI vulnerable occupaitons 
    ai_vuln = case_when(
  OCC2010 %in% c(10, 20, 110, 120, 150, 160, 430, 540, 565, 630, 650, 700, 735, 
                 740, 800, 910, 1005, 1006, 1007, 1010, 1020, 1050, 1105, 1106, 
                 1400, 1430, 1460, 1530, 1965, 2825, 3255, 3510, 3655, 4830, 4850, 
                 4940, 5100, 5110, 5120, 5160, 5240, 5400, 5520, 5600, 5700, 
                 5840,5860) ~ "AI Vulnerable",
  OCC2010 %in% c(9920, 9999) ~ "Unclassified",
  TRUE ~ "Not AI Vulnerable"
  ),
  ai_vul_grp = case_when(
    OCC2010 %in% c(10,20,110,120,150,160,430) ~ "Management",
    OCC2010 %in% c(540, 565, 630, 650, 700, 735,740) ~ "Business Operations",
    OCC2010 %in% c(800, 910) ~ "Financial Specialists",
    OCC2010 %in% c(1005, 1006, 1007, 1010, 1020, 1050,
                   1105, 1106, 1400, 1430, 1460, 1530,
                   1965) ~ "STEM",
    OCC2010 %in% c(2825) ~ "Entertainment and Media",
    OCC2010 %in% c(3255, 3510, 3655) ~ "Health Care",
    OCC2010 %in% c(4830, 4850, 4940) ~ "Sales and Related",
    OCC2010 %in% c(5100, 5110, 5120, 5160, 5240, 5400,
                   5520, 5600, 5700, 5840,5860) ~ "Admin & Office Support",
    TRUE ~ "Not AI Vulnerable"
  )

)
```

## Puerto Rican concentration by specific occupation

```{r Top 10 occupations for PR, message = FALSE, warning = FALSE}
top_10_pr_occ_by_year <- mapped_workers_db %>%
  # 1. ADD THIS FILTER to isolate Puerto Rican workers
  filter(is_pr == TRUE) %>%
  
  # 2. Group by Year, Occupation Code, AND ai_vuln so it doesn't get dropped
  group_by(YEAR, OCC2010, ai_vuln) %>%
  
  # 3. Sum the person weights to get the estimated total workers
  summarise(
    total_estimated_workers = sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  # 4. Bring the summary into local memory
  collect() %>%
  
  # (Notice the redundant ai_vuln mutate is gone!)
  
  # 5. Group LOCALLY by Year so our top 10 function applies to each year individually
  group_by(YEAR) %>%
  
  # 6. Extract the top 10 occupations with the highest worker count for each year
  slice_max(order_by = total_estimated_workers, n = 10) %>%
  
  # 7. Ungroup and sort cleanly so it's easy to read
  ungroup() %>%
  arrange(YEAR, desc(total_estimated_workers))

```

### Add labels to occupation categories

```{r Add labeles to Occupation, message=FALSE,warning=FALSE}

# Join the text labels to your top 10 list
top_10_pr_occ_labeled <- top_10_pr_occ_by_year %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  select(YEAR,OCC2010, occupation_name = lbl, total_estimated_workers, ai_vuln)

print(top_10_pr_occ_labeled)
```

### Visualize Top 10 occupations by year

```{r Plot top 10 occupations in timeseries, message=FALSE,warning=FALSE}
# Create the faceted time series plot
ggplot(top_10_pr_occ_labeled, aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name # Ensures lines connect properly
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Create a separate panel for each occupation
  # scales = "free_y" allows each plot to zoom in on its specific trend
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Apply distinct colors based on AI vulnerability
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      # Bold Orange
    "Not AI Vulnerable" = "#1f78b4",  # Cool Blue
    "Unclassified" = "grey70"         # Neutral Grey
  )) +
  
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    
    # Make the facet headers smaller and bold so long occupation names fit
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) # Tilts years to prevent overlap
  ) +
  
  scale_y_continuous(labels = label_comma()) +
  
  labs(
    title = "Top Occupations for Puerto Rican Workers Over Time",
    subtitle = "Estimated workforce volume by individual occupation",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```

```{r Top 10 occupations for PR visualization, message=FALSE, warning=FALSE}
ggplot(top_10_pr_occ_labeled, aes(
  x = total_estimated_workers, 
  # reorder_within sorts the bars largest-to-smallest INSIDE each facet
  y = tidytext::reorder_within(OCC2010, total_estimated_workers, YEAR), 
  fill = ai_vuln
)) +
  geom_col(alpha = 0.9) +
  
  # scales = "free_y" is crucial here so only that year's top 10 show in the panel
  facet_wrap(~YEAR, ncol = 4, scales = "free_y") +
  
  # This cleans up the Y-axis text that reorder_within generates
  tidytext::scale_y_reordered() +
  
  # Custom colors for vulnerability
  scale_fill_manual(values = c(
    "AI Vulnerable" = "#d95f02",      # Warning orange
    "Not AI Vulnerable" = "#2c3e50",  # Your standard navy blue
    "Unclassified" = "grey70"         # Neutral grey
  )) +
  my_theme +
  theme(
    # We bring the legend back so the AI colors make sense
    legend.position = "bottom",
    strip.text = element_text(size = 10, face = "bold"),
    # Shrink the y-axis text slightly so long occupation names fit
    axis.text.y = element_text(size = 7) 
  ) +
  scale_x_continuous(labels = scales::comma_format()) +
  labs(
    title = "Top 10 Occupations for Puerto Rican Workers by Year",
    subtitle = "Ranked by estimated workforce volume and colored by AI exposure",
    x = "Estimated Total Workers",
    y = NULL, 
    fill = "AI Vulnerability Risk" 
  )
```

```{r Pivot top 10 occupations wide, message=FALSE,warning=FALSE}
top_10_pr_occ_wide <- top_10_pr_occ_labeled %>%
  pivot_wider(names_from = "YEAR",values_from = total_estimated_workers)

```

### Top Occupations by sex

```{r Determine Occupations by sex, message=FALSE, warning=FALSE}

top_10_pr_occ_by_sex <- mapped_workers_db %>%
  # 1. Isolate Puerto Rican workers AND remove unknown sex codes (SEX == 9)
  filter(is_pr == TRUE, SEX != 9) %>%
  
  # 2. Create a clean label for the sexes
  mutate(sex_label = case_when(
    SEX == 1 ~ "Male",
    SEX == 2 ~ "Female"
  )) %>%
  
  # 3. Add sex_label to the grouping before summarizing
  group_by(YEAR, sex_label, OCC2010, ai_vuln) %>%
  
  # 4. Sum the person weights
  summarise(
    total_estimated_workers = sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  # 5. Bring into local memory
  collect() %>%
  
  # 6. Group LOCALLY by both Year AND Sex
  group_by(YEAR, sex_label) %>%
  
  # 7. Extract the top 10 for each specific demographic group
  slice_max(order_by = total_estimated_workers, n = 10) %>%
  
  # 8. Clean up and sort
  ungroup() %>%
  arrange(YEAR, sex_label, desc(total_estimated_workers))

# Join the text labels so it is readable
top_10_pr_occ_sex_labeled <- top_10_pr_occ_by_sex %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  select(YEAR, sex_label, OCC2010, occupation_name = lbl, 
         total_estimated_workers, ai_vuln)

```

### Visualize top 10 occupations for Males

```{r}
# Create the faceted time series plot
ggplot(top_10_pr_occ_sex_labeled %>% filter(sex_label == "Male"), aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name # Ensures lines connect properly
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Create a separate panel for each occupation
  # scales = "free_y" allows each plot to zoom in on its specific trend
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Apply distinct colors based on AI vulnerability
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      # Bold Orange
    "Not AI Vulnerable" = "#1f78b4",  # Cool Blue
    "Unclassified" = "grey70"         # Neutral Grey
  )) +
  
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    
    # Make the facet headers smaller and bold so long occupation names fit
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) # Tilts years to prevent overlap
  ) +
  
  scale_y_continuous(labels = label_comma()) +
  
  labs(
    title = "Top Occupations for Puerto Rican Male Workers Over Time",
    subtitle = "Estimated workforce volume by individual occupation",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```

### Visualize top 10 occupations for Females

```{r}
# Create the faceted time series plot
ggplot(top_10_pr_occ_sex_labeled %>% filter(sex_label == "Female"), aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name # Ensures lines connect properly
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Create a separate panel for each occupation
  # scales = "free_y" allows each plot to zoom in on its specific trend
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Apply distinct colors based on AI vulnerability
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      # Bold Orange
    "Not AI Vulnerable" = "#1f78b4",  # Cool Blue
    "Unclassified" = "grey70"         # Neutral Grey
  )) +
  
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    
    # Make the facet headers smaller and bold so long occupation names fit
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) # Tilts years to prevent overlap
  ) +
  
  scale_y_continuous(labels = label_comma()) +
  
  labs(
    title = "Top Occupations for Puerto Rican Female Workers Over Time",
    subtitle = "Estimated workforce volume by individual occupation",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```

### Continous visualization of Male and Female Top 10 Occupations

```{r Continuos top 10 occupations PR Males, message=FALSE, warning=FALSE}
# 1. Identify which occupations ever made the Top 10 for males
top_male_occs <- top_10_pr_occ_sex_labeled %>%
  filter(sex_label == "Male") %>%
  pull(OCC2010) %>%
  unique()

# 2. Pull the continuous history for ONLY those specific occupations
continuous_male_trends <- mapped_workers_db %>%
  filter(is_pr == TRUE, SEX == 1, OCC2010 %in% top_male_occs) %>%
  group_by(YEAR, OCC2010, ai_vuln) %>%
  summarise(total_estimated_workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  select(YEAR, OCC2010, occupation_name = lbl, ai_vuln, total_estimated_workers) %>%
   mutate(occupation_name = str_wrap(occupation_name, width = 35))

# 3. Plot the continuous data
ggplot(continuous_male_trends, aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name 
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      
    "Not AI Vulnerable" = "#1f78b4",  
    "Unclassified" = "grey70"         
  )) +
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) 
  ) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Top Occupations for Puerto Rican Male Workers Over Time",
    subtitle = "Continuous workforce volume for any occupation that ever reached the Top 10",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```

```{r Continuos top 10 occupations PR Males, message=FALSE, warning=FALSE}
# 1. Identify which occupations ever made the Top 10 for females
top_female_occs <- top_10_pr_occ_sex_labeled %>%
  filter(sex_label == "Female") %>%
  pull(OCC2010) %>%
  unique()

# 2. Pull the continuous history for ONLY those specific occupations
continuous_female_trends <- mapped_workers_db %>%
  # Note: SEX == 2 filters specifically for female workers in the database
  filter(is_pr == TRUE, SEX == 2, OCC2010 %in% top_female_occs) %>%
  group_by(YEAR, OCC2010, ai_vuln) %>%
  summarise(total_estimated_workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  select(YEAR, OCC2010, occupation_name = lbl, ai_vuln, total_estimated_workers) %>%
   mutate(occupation_name = str_wrap(occupation_name, width = 35))

# 3. Plot the continuous data for females
ggplot(continuous_female_trends, aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name 
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      
    "Not AI Vulnerable" = "#1f78b4",  
    "Unclassified" = "grey70"         
  )) +
  
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) 
  ) +
  
  scale_y_continuous(labels = scales::label_comma()) +
  
  labs(
    title = "Top Occupations for Puerto Rican Female Workers Over Time",
    subtitle = "Continuous workforce volume for any occupation that ever reached the Top 10",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```

# Exposure over time (2001 to 2024)

```{r Calculate and visualize exposure, message=FALSE, warning=FALSE}

ai_exposure_pr <- mapped_workers_db %>%
  filter(is_pr == TRUE, ai_vuln != "Unclassified") %>%
  group_by(YEAR, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  group_by(YEAR) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100, demographic = "Puerto Rican Workers") %>%
  ungroup()

ai_exposure_us <- mapped_workers_db %>%
  filter(ai_vuln != "Unclassified") %>%
  group_by(YEAR, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  group_by(YEAR) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100, demographic = "Total US Workers") %>%
  ungroup()

ai_exposure_data <- bind_rows(ai_exposure_pr, ai_exposure_us)
```

```{r }
p <- ggplot(ai_exposure_data %>% filter(ai_vuln == "AI Vulnerable"), 
       aes(x = YEAR, y = pct_workers, color = demographic)) +
  geom_line(linewidth = 1.5) +  
  geom_point(size = 3) +    
  my_theme +
  scale_color_manual(values = c("Puerto Rican Workers" = "#d95f02", "Total US Workers" = "#7570b3")) +
  scale_y_continuous(
    limits = c(0, 30),
    breaks = seq(0,100, by = 5),
    labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(2001, 2024, by = 5)) +
  labs(
    title = "Workforce Exposure to AI Technologies", 
    subtitle = "Percentage of workers employed in AI-vulnerable occupations (2001-2024)", 
    x = NULL, 
    y = "Share of Workforce", 
    caption = "Source: IPUMS USA, ACS (2001-2024)" 
  )

print(p)

# ggsave(
#   filename = "figures/ai_exposure_presentation.png",
#   plot = p,
#   width = 10, # Width in inches (standard for 16:9 slides)
#   height = 5.625, # Height in inches (10 / (16/9) = 5.625)
#   dpi = 300, # 300 DPI ensures it won't look pixelated on a projector
#   bg = "white" # Ensures the background is solid white, not transparent
# )
```

### Wage Premium or Penalty for AI vulnerable occupations

```{r Wage preiums calculation and visualization, message=FALSE, warning=FALSE}
ai_wage_pr <- mapped_workers_db %>%
  filter(is_pr == TRUE, ai_vuln != "Unclassified", INCWAGE > 0, INCWAGE < 999999) %>%
  group_by(YEAR, ai_vuln) %>%
  summarise(avg_wage = sum(INCWAGE * CPI99 * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  mutate(demographic = "Puerto Rican Workers")

ai_wage_us <- mapped_workers_db %>%
  filter(ai_vuln != "Unclassified", INCWAGE > 0, INCWAGE < 999999) %>%
  group_by(YEAR, ai_vuln) %>%
  summarise(avg_wage = sum(INCWAGE * CPI99 * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  mutate(demographic = "Total US Workers")

ai_wage_data <- bind_rows(ai_wage_pr, ai_wage_us)
```

```{r}
ggplot(ai_wage_data, aes(x = YEAR, y = avg_wage, color = demographic, linetype = ai_vuln)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  my_theme +
  scale_color_manual(values = c("Puerto Rican Workers" = "#d95f02", "Total US Workers" = "#7570b3")) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Historical Wage Trajectories by AI Vulnerability",
    subtitle = "Real 1999 Dollars",
    x = "Year",
    y = "Average Annual Wage ($)"
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5))
```

### Demographic variations in AI vulnerable occupations

```{r Demographic difference calculation and visualization, message=FALSE, warning=FALSE}
# Puerto Rican Data
ai_demographics_pr <- mapped_workers_db %>%
  filter(is_pr == TRUE, ai_vuln != "Unclassified", SEX != 9) %>%
  mutate(sex_label = ifelse(SEX == 1, "Male", "Female")) %>%
  group_by(YEAR, sex_label, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  group_by(YEAR, sex_label) %>%
  mutate(
    pct_workers = (workers / sum(workers)) * 100,
    demographic = "Puerto Rican Workers" # Added identifier
  ) %>%
  ungroup()

# Total US Data
ai_demographics_us <- mapped_workers_db %>%
  filter(ai_vuln != "Unclassified", SEX != 9) %>%
  mutate(sex_label = ifelse(SEX == 1, "Male", "Female")) %>%
  group_by(YEAR, sex_label, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  group_by(YEAR, sex_label) %>%
  mutate(
    pct_workers = (workers / sum(workers)) * 100,
    demographic = "Total US Workers" # Added identifier
  ) %>%
  ungroup()

# Combine datasets
ai_demographics <- bind_rows(ai_demographics_pr, ai_demographics_us)

```

### Demographic variations in AI vulnerable occupations for total population

```{r}
p <- ggplot(
  ai_demographics %>% filter(ai_vuln == "AI Vulnerable"), 
  aes(
    x = YEAR, 
    y = pct_workers, 
    color = demographic, 
    linetype = sex_label, 
    shape = sex_label # Makes the points distinct too!
  )
) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  
  my_theme +
  
  # Assign colors to the demographics
  scale_color_manual(values = c(
    "Puerto Rican Workers" = "#d95f02", 
    "Total US Workers" = "#7570b3"
  )) +
  
  # Format axes
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(1950, 2024, by = 10)) +
  
  labs(
    title = "AI Vulnerability by Sex: Puerto Rican vs. Total US Workers",
    subtitle = "Percentage of each sex's workforce in AI-vulnerable occupations",
    x = NULL, 
    y = "Exposure Rate",
    color = "Demographic", # Titles for the legends
    linetype = "Sex",
    shape = "Sex",
    caption = "Source: IPUMS USA, Decennial Census and ACS (1950-2024)"
  )

ggsave(
  filename = "figures/ai_exposure_gender_presentation.png",
  plot = p,
  width = 10, # Width in inches (standard for 16:9 slides)
  height = 5.625, # Height in inches (10 / (16/9) = 5.625)
  dpi = 300, # 300 DPI ensures it won't look pixelated on a projector
  bg = "white" # Ensures the background is solid white, not transparent
)

print(p)
```

## AI Vulnerable Occupations Puerto Rican trends

```{r Calculate Puerto Ricans in AI Vulnerable Occ, message=FALSE,warning=FALSE}
ai_occ_exposure_pr <- mapped_workers_db %>%
  filter(is_pr == TRUE, ai_vuln == "AI Vulnerable") %>%
  group_by(YEAR, OCC2010) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  group_by(YEAR) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100, demographic = "Puerto Rican Workers") %>%
  ungroup()

ai_occ_exposure_us <- mapped_workers_db %>%
  filter(ai_vuln == "AI Vulnerable") %>%
  group_by(YEAR, OCC2010) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  group_by(YEAR) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100, demographic = "Total US Workers") %>%
  ungroup()

ai_exposure_by_occ <- bind_rows(ai_occ_exposure_pr, ai_occ_exposure_us)

```

## Share of AI Workers per Occupation

```{r Share of AI Workers by Occupation}
# 1. Clean the labels before plotting
ai_exposure_plot_data <- ai_exposure_by_occ %>%
  # Rename 'lbl' (from the left_join) so it's easier to reference
  rename(occupation_name = lbl) %>%
  # Wrap long IPUMS occupation text so it stacks neatly in the facet headers
  mutate(occupation_name = str_wrap(occupation_name, width = 35))

# 2. Plot the comparison trends
ggplot(ai_exposure_plot_data, aes(
  x = YEAR, 
  y = pct_workers, 
  color = demographic,
  group = demographic
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Facet by occupation to give each job its own comparison panel
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Distinct colors for the demographics
  scale_color_manual(values = c(
    "Puerto Rican Workers" = "#d95f02", # Orange for emphasis
    "Total US Workers" = "#7570b3"      # Cooler purple/blue for baseline
  )) +
  
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(), # Hides "demographic" as it's self-explanatory
    legend.text = element_text(size = 10, face = "bold"),
    
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  ) +
  
  # Format Y-axis as percentages
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  
  labs(
    title = "Concentration of AI-Vulnerable Occupations",
    subtitle = "Comparing the distribution of exposed Puerto Rican workers vs. the Total US baseline",
    x = "Year",
    y = "Share of AI-Vulnerable Workforce (%)"
  )
```

```{r}
ai_exposure_2024 <- ai_exposure_by_occ %>%
  filter(YEAR == 2024) %>%
  mutate(
    # Create short, presentation-friendly names based on the OCC2010 codes
    short_occ_name = case_when(
      OCC2010 == 10 ~ "Execs & Legislators",
      OCC2010 == 20 ~ "Gen & Ops Managers",
      OCC2010 == 110 ~ "IT Managers",
      OCC2010 == 120 ~ "Financial Managers",
      OCC2010 == 800 ~ "Accountants & Auditors",
      OCC2010 == 1010 ~ "Computer Programmers",
      OCC2010 == 1020 ~ "Software Developers",
      OCC2010 == 3255 ~ "Registered Nurses",
      OCC2010 == 4850 ~ "Wholesale/Mfg Sales",
      OCC2010 == 5240 ~ "Customer Service Reps",
      OCC2010 == 5400 ~ "Receptionists",
      OCC2010 == 5700 ~ "Secretaries & Admins",
      OCC2010 == 7340 ~ "Maintenance & Repair",
      TRUE ~ lbl # Fallback to the original label just in case
    ),
    # Reorder the abbreviated names based on the percentage for a clean descending chart
    short_occ_name = fct_reorder(short_occ_name, pct_workers, .fun = max)
  )

p <- ggplot(ai_exposure_2024, aes(
  x = pct_workers, 
  y = short_occ_name, 
  fill = demographic 
)) +
  # Draw side-by-side bars
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  
  # NEW: Add direct percentage labels to the ends of the dodged bars
  geom_text(
    aes(label = paste0(round(pct_workers, 1), "%")),
    position = position_dodge(width = 0.8), # Must match the geom_col dodge width
    hjust = -0.15,                          # Pushes text slightly outside the bar
    size = 3.75, 
    fontface = "bold", 
    color = "gray30"
  ) +
  
  # Apply your distinct demographic colors
  scale_fill_manual(values = c(
    "Puerto Rican Workers" = "#d95f02", 
    "Total US Workers" = "#7570b3"      
  )) +
  
  my_theme +
  
  # Format X-axis as percentages and expand the right side to fit the text labels
  scale_x_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.15)) 
  ) +
  
  labs(
    title = "Concentration of AI-Vulnerable Occupations (2024)",
    subtitle = "Comparing the distribution of exposed Puerto Rican workers vs. the Total US baseline",
    x = "Share of AI-Vulnerable Workforce",
    y = NULL, 
    fill = NULL,
    caption = "Source: IPUMS USA, Decennial Census and ACS (1950-2024)" # Standardized caption
  ) +
  
  # Presentation styling overrides
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14, color = "grey30"),
    axis.text.y = element_text(size = 12, face = "bold"), 
    axis.text.x = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    
    # Move legend to the top and increase text size
    legend.position = "top",
    legend.text = element_text(size = 14),
    
    # Remove horizontal grid lines to keep the background behind the bars clean
    panel.grid.major.y = element_blank(),
    plot.caption = element_text(size = 10, color = "grey50", hjust = 1)
  )

ggsave(
  filename = "figures/vuln_occ_share_presentation.png",
  plot = p,
  width = 10, # Width in inches (standard for 16:9 slides)
  height = 5.625, # Height in inches (10 / (16/9) = 5.625)
  dpi = 300, # 300 DPI ensures it won't look pixelated on a projector
  bg = "white" # Ensures the background is solid white, not transparent
)

print(p)
```

### Average Wage by Occupaitons vulnerable to AI over time (2001-2024)

```{r Calculate Average Wage per Occupations, message=FALSE,warning=FALSE}
ai_occ_wage_pr <- mapped_workers_db %>%
  filter(is_pr == TRUE, ai_vuln == "AI Vulnerable", INCWAGE > 0, INCWAGE < 999999) %>%
  group_by(YEAR, OCC2010) %>%
  summarise(avg_wage = sum(INCWAGE * CPI99 * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  mutate(demographic = "Puerto Rican Workers")

ai_occ_wage_us <- mapped_workers_db %>%
  filter(ai_vuln == "AI Vulnerable", INCWAGE > 0, INCWAGE < 999999) %>%
  group_by(YEAR, OCC2010) %>%
  summarise(avg_wage = sum(INCWAGE * CPI99 * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  mutate(demographic = "Total US Workers")

ai_wage_data_by_occ <- bind_rows(ai_occ_wage_pr, ai_occ_wage_us)
```

### Visualize Average Wage over time for exposed Occupations

```{r Visualize Average Wage for AI Vulnerable Occ, message=FALSE,warning=FALSE}

# 1. Clean the labels before plotting
ai_wage_data_by_occ <- ai_wage_data_by_occ %>%
  # Rename 'lbl' (from the left_join) so it's easier to reference
  rename(occupation_name = lbl) %>%
  # Wrap long IPUMS occupation text so it stacks neatly in the facet headers
  mutate(occupation_name = str_wrap(occupation_name, width = 35))

# 2. Plot the comparison trends
ggplot(ai_wage_data_by_occ, aes(
  x = YEAR, 
  y = avg_wage, 
  color = demographic,
  group = demographic
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Facet by occupation to give each job its own comparison panel
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Distinct colors for the demographics
  scale_color_manual(values = c(
    "Puerto Rican Workers" = "#d95f02", # Orange for emphasis
    "Total US Workers" = "#7570b3"      # Cooler purple/blue for baseline
  )) +
  
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(), # Hides "demographic" as it's self-explanatory
    legend.text = element_text(size = 10, face = "bold"),
    
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  ) +
  
  # Format Y-axis as percentages
  scale_y_continuous(labels = scales::dollar_format()) +
  
  labs(
    title = "Wage Trajectory for AI Vulnerable Occupatoins",
    subtitle = "Real 1999 Dollars",
    x = "Year",
    y = "Average Annual Wage ($)"
  )
```

## AI vulnerability comparison with other race categories

```{r}
# Calculate AI exposure by Race, Ethnicity, and Puerto Rican Origin
ai_race_exposure <- mapped_workers_db %>%
  filter(ai_vuln != "Unclassified") %>%
  mutate(
    race_cat = case_when(
      HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261 ~ "Puerto Rican",
      HISPAN > 0 & HISPAN < 9 ~ "Other Hispanic",
      RACE == 1 ~ "White",
      RACE == 2 ~ "Black/African American",
      RACE == 3 ~ "AIAN",
      RACE >= 4 & RACE <= 6 ~ "Asian/PI",
      RACE >= 7 ~ "Other",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(race_cat)) %>%
  group_by(YEAR, race_cat, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  
  # Pull the summarized data into local R memory FIRST
  collect() %>%
  
  # NOW apply the factor conversion locally
  mutate(
    race_cat = factor(race_cat, levels = c(
      "Puerto Rican", "Other Hispanic", "White", "Black/African American", 
      "Asian/PI", "AIAN", "Other"
    ))
  ) %>%
  
  # Proceed with the percentage calculation
  group_by(YEAR, race_cat) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100) %>%
  ungroup()
```

## Visualize share of vulnerable workers

```{r}

p <- ggplot(ai_race_exposure %>% filter(YEAR == 2024, ai_vuln == "AI Vulnerable"), 
       aes(x = pct_workers, y = race_cat, fill = race_cat)) +
  geom_col(alpha = 0.9) +
  
  # Add direct percentage labels to the end of each bar for instant readability
  geom_text(
    aes(label = paste0(round(pct_workers, 1), "%")),
    hjust = -0.2, # Pushes the text slightly outside the bar
    size = 5, 
    fontface = "bold", 
    color = "gray20"
  ) +
  
  my_theme +
  
  # Keep the consistent colors, highlighting Puerto Ricans in orange
  scale_fill_manual(values = c(
    "Puerto Rican" = "#d95f02",           # Your signature highlight orange
    "Other Hispanic" = "#bdc3c7",         # Muted gray
    "White" = "#bdc3c7",
    "Black/African American" = "#bdc3c7",
    "Asian/PI" = "#bdc3c7",
    "AIAN" = "#bdc3c7",
    "Other" = "#bdc3c7"
  )) +
  
  # Expand the X-axis slightly so the text labels don't get cut off on the right
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  
  labs(
    title = "AI Vulnerability by Race and Ethnicity (2024)",
    subtitle = "Share of workers employed in AI-vulnerable occupations",
    x = NULL,
    y = NULL, # The categories are self-explanatory, no y-axis title needed
    caption = "Source: IPUMS USA, ACS (2024)"
  ) +
  
  # Aggressive decluttering for the presentation slide
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14, color = "grey30"),
    axis.text.y = element_text(size = 14, face = "bold"),
    
    # Hide the legend, X-axis text, and gridlines to keep focus on the bars and labels
    legend.position = "none",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    
    plot.caption = element_text(size = 10, color = "grey50", hjust = 1) 
  )

# ggsave(
#   filename = "figures/ai_race_comp_presentation.png",
#   plot = p,
#   width = 10, # Width in inches (standard for 16:9 slides)
#   height = 5.625, # Height in inches (10 / (16/9) = 5.625)
#   dpi = 300, # 300 DPI ensures it won't look pixelated on a projector
#   bg = "white" # Ensures the background is solid white, not transparent
# )
```

## Wage tier analysis for Top 10 Occupations

```{r}
# 1. Calculate National Wage Tiers for Occupations by Year
us_occ_wage_tiers <- mapped_workers_db %>%
  filter(INCWAGE > 0, INCWAGE < 999999) %>%
  group_by(YEAR, OCC2010) %>%
  summarise(
    us_avg_wage = sum(INCWAGE * CPI99 * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  collect() %>% # Bring to local memory before using ntile()
  group_by(YEAR) %>%
  mutate(
    tier_num = ntile(us_avg_wage, 3),
    wage_tier = case_when(
      tier_num == 3 ~ "High-Paying",
      tier_num == 2 ~ "Middle-Paying",
      tier_num == 1 ~ "Low-Paying"
    ),
    wage_tier = factor(wage_tier, levels = c("High-Paying", "Middle-Paying", "Low-Paying"))
  ) %>%
  ungroup() %>%
  select(YEAR, OCC2010, wage_tier)

# 2. Join Wage Tiers to the Continuous Male Trends
continuous_male_trends_tiered <- continuous_male_trends %>%
  left_join(us_occ_wage_tiers, by = c("YEAR", "OCC2010"))

# 3. Join Wage Tiers to the Continuous Female Trends
continuous_female_trends_tiered <- continuous_female_trends %>%
  left_join(us_occ_wage_tiers, by = c("YEAR", "OCC2010"))

# Define the abbreviation logic
abbreviate_occupations <- function(df) {
  df %>%
    mutate(
      short_occ_name = case_when(
        # Replace these examples with the actual OCC2010 codes in your Top 10
        OCC2010 == 7200 ~ "Mechanics",
        OCC2010 == 9130 ~ "Truck Drivers",
        OCC2010 == 4720 ~ "Cashiers",
        OCC2010 == 4700 ~ "Retail Supervisors",
        OCC2010 == 4060 ~ "Food Service Workers",
        OCC2010 == 4600 ~ "Childcare Workers",
        OCC2010 == 3255 ~ "Registered Nurses",
        OCC2010 == 4110 ~ "Waiters & Waitresses",
        OCC2010 == 4760 ~ "Retail Salespersons",
        OCC2010 == 5700 ~ "Secretaries & Admins",
        OCC2010 == 5000 ~ "Office Supervisors",
        OCC2010 == 9620 ~ "Freight and Stock Movers",
        OCC2010 == 430 ~ "Other Managers",
        OCC2010 == 8965 ~ "Other Production Workers",
        
        
        # Fallback: keep the original name if you didn't explicitly abbreviate it
        TRUE ~ occupation_name 
      ),
      # Optional: Wrap the text just in case the abbreviation is still a bit wide
      short_occ_name = str_wrap(short_occ_name, width = 25)
    )
}

# Apply it to both of your continuous trend dataframes
continuous_male_trends_tiered <- abbreviate_occupations(continuous_male_trends_tiered)
continuous_female_trends_tiered <- abbreviate_occupations(continuous_female_trends_tiered)
```

```{r}
p <- ggplot(continuous_male_trends_tiered, aes(x = YEAR, y = total_estimated_workers)) +
  
  # Draw a neutral line to establish the workforce volume trend
  geom_line(color = "gray60", linewidth = 1.2) +
  
  # Add points colored by the wage tier to show economic shifts over time
  geom_point(aes(color = wage_tier), size = 2) +
  
  facet_wrap(~ short_occ_name, ncol = 4, scales = "free_y") +
  
  # Distinct colors for wage tiers
  scale_color_manual(values = c(
    "High-Paying" = "#1b9e77",   # Green
    "Middle-Paying" = "#d95f02", # Orange
    "Low-Paying" = "#7570b3"     # Purple
  )) +
  
  my_theme +
  
  scale_y_continuous(labels = scales::label_comma()) +
  scale_x_continuous(breaks = seq(1950, 2024, by = 10)) +
  
  labs(
    title = "Top Occupations for Puerto Rican Male Workers Over Time",
    subtitle = "Continuous workforce volume, colored by the national wage tier of the occupation",
    x = NULL,
    y = "Estimated Total Workers",
    color = "National Wage Tier",
    caption = "Source: IPUMS USA, ACS (2001-2024)"
  ) +
  
  # Presentation styling
  theme(# Reduce header text size (e.g., down to 7pt or 8pt) and trim padding
  strip.text = element_text(
    size = 7.5, 
    face = "bold", 
    margin = margin(t = 2, b = 2) # Minimal top and bottom padding
  ),
  
  # Shrink axis text slightly to give panels more room
  axis.text.x = element_text(size = 7, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 7),
  
  # Reduce spacing between individual facet panels
  panel.spacing = unit(0.5, "lines"),
  
  # Legend & Title Sizing
  plot.title = element_text(size = 16, face = "bold"),
  plot.subtitle = element_text(size = 12, color = "grey30"),
  legend.position = "top",
  legend.box = "vertical",
  legend.text = element_text(size = 10),
  plot.caption = element_text(size = 9, color = "grey50", hjust = 1)
  )

ggsave(
  filename = "figures/top_occ_male_presentation.png",
  plot = p,
  width = 10, # Width in inches (standard for 16:9 slides)
  height = 5.625, # Height in inches (10 / (16/9) = 5.625)
  dpi = 300, # 300 DPI ensures it won't look pixelated on a projector
  bg = "white" # Ensures the background is solid white, not transparent
)

print(p)
```

```{r}
p <- ggplot(continuous_female_trends_tiered, aes(x = YEAR, y = total_estimated_workers)) +
  
  # Draw a neutral line to establish the workforce volume trend
  geom_line(color = "gray60", linewidth = 1.2) +
  
  # Add points colored by the wage tier to show economic shifts over time
  geom_point(aes(color = wage_tier), size = 2) +
  
  facet_wrap(~ short_occ_name, ncol = 4, scales = "free_y") +
  
  # Distinct colors for wage tiers
  scale_color_manual(values = c(
    "High-Paying" = "#1b9e77",   # Green
    "Middle-Paying" = "#d95f02", # Orange
    "Low-Paying" = "#7570b3"     # Purple
  )) +
  
  my_theme +
  
  scale_y_continuous(labels = scales::label_comma()) +
  scale_x_continuous(breaks = seq(1950, 2024, by = 10)) +
  
  labs(
    title = "Top Occupations for Puerto Rican Female Workers Over Time",
    subtitle = "Continuous workforce volume, colored by the national wage tier of the occupation",
    x = NULL,
    y = "Estimated Total Workers",
    color = "National Wage Tier",
    caption = "Source: IPUMS USA, ACS (2001-2024)"
  ) +
  
  # Presentation styling
  theme(# Reduce header text size (e.g., down to 7pt or 8pt) and trim padding
  strip.text = element_text(
    size = 7.5, 
    face = "bold", 
    margin = margin(t = 2, b = 2) # Minimal top and bottom padding
  ),
  
  # Shrink axis text slightly to give panels more room
  axis.text.x = element_text(size = 7, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 7),
  
  # Reduce spacing between individual facet panels
  panel.spacing = unit(0.5, "lines"),
  
  # Legend & Title Sizing
  plot.title = element_text(size = 16, face = "bold"),
  plot.subtitle = element_text(size = 12, color = "grey30"),
  legend.position = "top",
  legend.box = "vertical",
  legend.text = element_text(size = 10),
  plot.caption = element_text(size = 9, color = "grey50", hjust = 1)
  )

ggsave(
  filename = "figures/top_occ_female_presentation.png",
  plot = p,
  width = 10, # Width in inches (standard for 16:9 slides)
  height = 5.625, # Height in inches (10 / (16/9) = 5.625)
  dpi = 300, # 300 DPI ensures it won't look pixelated on a projector
  bg = "white" # Ensures the background is solid white, not transparent
)
```

```{r}
# 1. Add a sex identifier to both datasets and combine them
combined_gender_trends <- bind_rows(
  continuous_male_trends_tiered %>% mutate(sex_label = "Male"),
  continuous_female_trends_tiered %>% mutate(sex_label = "Female")
)

# 2. Plot the combined data
ggplot(combined_gender_trends, aes(x = YEAR, y = total_estimated_workers, group = sex_label)) +
  
  # Map line type to sex so the audience can tell them apart
  geom_line(aes(linetype = sex_label), color = "gray60", linewidth = 1) +
  
  # Map color to wage tier, and shape to sex
  geom_point(aes(color = wage_tier, shape = sex_label), size = 2.5) +
  
  facet_wrap(~ short_occ_name, ncol = 4, scales = "free_y") +
  
  scale_color_manual(values = c(
    "High-Paying" = "#1b9e77",   
    "Middle-Paying" = "#d95f02", 
    "Low-Paying" = "#7570b3"     
  )) +
  
  my_theme +
  
  scale_y_continuous(labels = scales::label_comma()) +
  scale_x_continuous(breaks = seq(1950, 2024, by = 10)) +
  
  labs(
    title = "Top Occupations for Puerto Rican Workers by Sex",
    subtitle = "Comparing male and female workforce volumes across major occupations",
    x = NULL,
    y = "Estimated Total Workers",
    color = "National Wage Tier",
    linetype = "Sex",
    shape = "Sex",
    caption = "Source: IPUMS USA, Decennial Census and ACS (1950-2024)"
  ) +
  
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14, color = "grey30"),
    
    # We need the legends at the top to be clear and distinct
    legend.position = "top",
    legend.box = "vertical", # Stacks the two legends so they don't crash into each other
    legend.text = element_text(size = 12, face = "bold"),
    legend.title = element_text(size = 12),
    
    strip.text = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    plot.caption = element_text(size = 10, color = "grey50", hjust = 1)
  )
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 999)
selected_years <- c(2005, 2010, 2015, 2020)

# Source central setup file
source("00_project_setup.R")
```

## AI Exposed Occupations Analysis

### Map industry codes and AI vulnerable Occupations

```{r Mapping Industry and Occupaiton Categories, message=FALSE,warning=FALSE}
mapped_workers_db <- data_acs %>%
  # Apply base employment filter, correcting the missing GQ != 3
  filter(AGE >= 16, EMPSTAT == 1, GQ != 3) %>%
  mutate(
    # Create a boolean flag for PR status to split the data later
    is_pr = case_when(
      HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261 ~ TRUE,
      TRUE ~ FALSE
    ),
    
    # Map Industries
    ind_sector = case_when(
    IND1990 >= 10 & IND1990 <= 32 ~ "01. Agriculture, Forestry, Fishing & Hunting",
    IND1990 >= 40 & IND1990 <= 50 ~ "02. Mining, Quarrying, & Oil/Gas Extraction",
    IND1990 >= 450 & IND1990 <= 472 ~ "03. Utilities",
    IND1990 == 60 ~ "04. Construction",
    IND1990 >= 100 & IND1990 <= 392 ~ "05. Manufacturing",
    IND1990 >= 500 & IND1990 <= 571 ~ "06. Wholesale Trade",

    # Excludes 641 (Eating/Drinking Places) to align with modern NAICS grouping
    (IND1990 >= 580 & IND1990 <= 691) & IND1990 != 641 ~ "07. Retail Trade",
    IND1990 >= 400 & IND1990 <= 432 ~ "08. Transportation & Warehousing",
    IND1990 >= 440 & IND1990 <= 442 ~ "09. Information", # Maps 1990 Communications here
    IND1990 >= 700 & IND1990 <= 711 ~ "10. Finance & Insurance",
    IND1990 == 712 ~ "11. Real Estate & Rental/Leasing",

    # Merges 1990 Engineering/Architecture/Accounting with related professional codes
    IND1990 == 841 | (IND1990 >= 881 & IND1990 <= 893) ~ "12. Professional, Scientific, & Tech Services",
    IND1990 >= 721 & IND1990 <= 741 ~ "13. Administrative, Support, & Waste Mgmt",
    IND1990 >= 842 & IND1990 <= 860 ~ "14. Educational Services",
    (IND1990 >= 812 & IND1990 <= 840) | (IND1990 >= 861 & IND1990 <= 873) ~ "15. Health Care & Social Assistance",
    IND1990 >= 800 & IND1990 <= 810 ~ "16. Arts, Entertainment, & Recreation",

    # Combines 1990 Eating/Drinking places (641) with Hotels/Lodging (762, 770)
    IND1990 %in% c(641, 762, 770) ~ "17. Accommodation & Food Services",
    IND1990 %in% c(750, 751, 752, 761, 771, 772, 781, 791) ~ "18. Other Services (Except Public Admin)",
    IND1990 >= 900 & IND1990 <= 932 ~ "19. Public Administration",
    IND1990 >= 940 & IND1990 <= 960 ~ "20. Military",
      TRUE ~ "21. Unclassified / Not in Labor Force"
    ),
    
    # Map Occupations
    # Map Occupations
    occ_cat1 = case_when(
      (OCC2010 >= 10 &
         OCC2010 <= 960) |
        (OCC2010 >= 2600 &
           OCC2010 <= 2960) ~ "Management, Business, Entertainment & Media",
      OCC2010 >= 1000 & OCC2010 <= 1965 ~ "STEM",
      (OCC2010 >= 2000 &
         OCC2010 <= 2550) |
        (OCC2010 >= 4300 &
           OCC2010 <= 4650) ~ "Education, Community & Social Services",
      OCC2010 >= 3000 & OCC2010 <= 3655 ~ "Health & Medical",
      (OCC2010 >= 3700 &
         OCC2010 <= 3955) |
        (OCC2010 >= 9000 &
           OCC2010 <= 9420) ~ "Law Enforcement & Transportation",
      OCC2010 >= 4000 &
        OCC2010 <= 4250 ~ "Food, Maintenance, Building & Grounds Cleaning",
      OCC2010 >= 4700 & OCC2010 <= 4965 ~ "Sales and Related",
      OCC2010 >= 5000 &
        OCC2010 <= 5940 ~ "Administrative and Office Support",
      OCC2010 >= 6005 &
        OCC2010 <= 7630 ~ "Natural Resources, Construction & Maintenance",
      (OCC2010 >= 7700 &
         OCC2010 <= 8965) |
        (OCC2010 >= 9500 &
           OCC2010 <= 9750) ~ "Production and Material Moving",
      OCC2010 >= 9800 & OCC2010 <= 9830 ~ "Military",
      OCC2010 %in% c(9920, 9999) ~ "Unclassified",
      TRUE ~ "Unclassified"
    ),
    # Map AI vulnerable occupaitons 
    ai_vuln = case_when(
  OCC2010 %in% c(10, 20, 110, 120, 150, 630, 430, 650, 735, 740, 800, 910, 1005, 1006,
                 1010, 1020, 1050, 1105, 1430, 1530, 1965, 3255, 3655, 4850, 4940, 5110,
                 5120, 5160, 5240, 
                 5400, 5600, 5700, 5860, 7340) ~ "AI Vulnerable",
  OCC2010 %in% c(9920, 9999) ~ "Unclassified",
  TRUE ~ "Not AI Vulnerable"
  )
)
```

## Puerto Rican concentration by specific occupation

```{r Top 10 occupations for PR, message = FALSE, warning = FALSE}
top_10_pr_occ_by_year <- mapped_workers_db %>%
  # 1. ADD THIS FILTER to isolate Puerto Rican workers
  filter(is_pr == TRUE) %>%
  
  # 2. Group by Year, Occupation Code, AND ai_vuln so it doesn't get dropped
  group_by(YEAR, OCC2010, ai_vuln) %>%
  
  # 3. Sum the person weights to get the estimated total workers
  summarise(
    total_estimated_workers = sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  # 4. Bring the summary into local memory
  collect() %>%
  
  # (Notice the redundant ai_vuln mutate is gone!)
  
  # 5. Group LOCALLY by Year so our top 10 function applies to each year individually
  group_by(YEAR) %>%
  
  # 6. Extract the top 10 occupations with the highest worker count for each year
  slice_max(order_by = total_estimated_workers, n = 10) %>%
  
  # 7. Ungroup and sort cleanly so it's easy to read
  ungroup() %>%
  arrange(YEAR, desc(total_estimated_workers))

```

### Add labels to occupation categories

```{r Add labeles to Occupation, message=FALSE,warning=FALSE}
ddi <- read_ipums_ddi("./usa_00121.xml")
occ_labels <- ipums_val_labels(ddi, OCC2010)

# Join the text labels to your top 10 list
top_10_pr_occ_labeled <- top_10_pr_occ_by_year %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  select(YEAR,OCC2010, occupation_name = lbl, total_estimated_workers, ai_vuln)

print(top_10_pr_occ_labeled)
```

### Visualize Top 10 occupations by year

```{r Plot top 10 occupations in timeseries, message=FALSE,warning=FALSE}
# Create the faceted time series plot
ggplot(top_10_pr_occ_labeled, aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name # Ensures lines connect properly
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Create a separate panel for each occupation
  # scales = "free_y" allows each plot to zoom in on its specific trend
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Apply distinct colors based on AI vulnerability
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      # Bold Orange
    "Not AI Vulnerable" = "#1f78b4",  # Cool Blue
    "Unclassified" = "grey70"         # Neutral Grey
  )) +
  
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    
    # Make the facet headers smaller and bold so long occupation names fit
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) # Tilts years to prevent overlap
  ) +
  
  scale_y_continuous(labels = label_comma()) +
  
  labs(
    title = "Top Occupations for Puerto Rican Workers Over Time",
    subtitle = "Estimated workforce volume by individual occupation",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```

```{r Top 10 occupations for PR visualization, message=FALSE, warning=FALSE}
ggplot(top_10_pr_occ_labeled, aes(
  x = total_estimated_workers, 
  # reorder_within sorts the bars largest-to-smallest INSIDE each facet
  y = tidytext::reorder_within(OCC2010, total_estimated_workers, YEAR), 
  fill = ai_vuln
)) +
  geom_col(alpha = 0.9) +
  
  # scales = "free_y" is crucial here so only that year's top 10 show in the panel
  facet_wrap(~YEAR, ncol = 4, scales = "free_y") +
  
  # This cleans up the Y-axis text that reorder_within generates
  tidytext::scale_y_reordered() +
  
  # Custom colors for vulnerability
  scale_fill_manual(values = c(
    "AI Vulnerable" = "#d95f02",      # Warning orange
    "Not AI Vulnerable" = "#2c3e50",  # Your standard navy blue
    "Unclassified" = "grey70"         # Neutral grey
  )) +
  my_theme +
  theme(
    # We bring the legend back so the AI colors make sense
    legend.position = "bottom",
    strip.text = element_text(size = 10, face = "bold"),
    # Shrink the y-axis text slightly so long occupation names fit
    axis.text.y = element_text(size = 7) 
  ) +
  scale_x_continuous(labels = scales::comma_format()) +
  labs(
    title = "Top 10 Occupations for Puerto Rican Workers by Year",
    subtitle = "Ranked by estimated workforce volume and colored by AI exposure",
    x = "Estimated Total Workers",
    y = NULL, 
    fill = "AI Vulnerability Risk" 
  )
```

```{r Pivot top 10 occupations wide, message=FALSE,warning=FALSE}
top_10_pr_occ_wide <- top_10_pr_occ_labeled %>%
  pivot_wider(names_from = "YEAR",values_from = total_estimated_workers)

```

### Top Occupations by sex

```{r Determine Occupations by sex, message=FALSE, warning=FALSE}

top_10_pr_occ_by_sex <- mapped_workers_db %>%
  # 1. Isolate Puerto Rican workers AND remove unknown sex codes (SEX == 9)
  filter(is_pr == TRUE, SEX != 9) %>%
  
  # 2. Create a clean label for the sexes
  mutate(sex_label = case_when(
    SEX == 1 ~ "Male",
    SEX == 2 ~ "Female"
  )) %>%
  
  # 3. Add sex_label to the grouping before summarizing
  group_by(YEAR, sex_label, OCC2010, ai_vuln) %>%
  
  # 4. Sum the person weights
  summarise(
    total_estimated_workers = sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  # 5. Bring into local memory
  collect() %>%
  
  # 6. Group LOCALLY by both Year AND Sex
  group_by(YEAR, sex_label) %>%
  
  # 7. Extract the top 10 for each specific demographic group
  slice_max(order_by = total_estimated_workers, n = 10) %>%
  
  # 8. Clean up and sort
  ungroup() %>%
  arrange(YEAR, sex_label, desc(total_estimated_workers))

# Join the text labels so it is readable
top_10_pr_occ_sex_labeled <- top_10_pr_occ_by_sex %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  select(YEAR, sex_label, OCC2010, occupation_name = lbl, 
         total_estimated_workers, ai_vuln)

```

### Visualize top 10 occupations for Males

```{r}
# Create the faceted time series plot
ggplot(top_10_pr_occ_sex_labeled %>% filter(sex_label == "Male"), aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name # Ensures lines connect properly
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Create a separate panel for each occupation
  # scales = "free_y" allows each plot to zoom in on its specific trend
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Apply distinct colors based on AI vulnerability
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      # Bold Orange
    "Not AI Vulnerable" = "#1f78b4",  # Cool Blue
    "Unclassified" = "grey70"         # Neutral Grey
  )) +
  
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    
    # Make the facet headers smaller and bold so long occupation names fit
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) # Tilts years to prevent overlap
  ) +
  
  scale_y_continuous(labels = label_comma()) +
  
  labs(
    title = "Top Occupations for Puerto Rican Male Workers Over Time",
    subtitle = "Estimated workforce volume by individual occupation",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```

### Visualize top 10 occupations for Females

```{r}
# Create the faceted time series plot
ggplot(top_10_pr_occ_sex_labeled %>% filter(sex_label == "Female"), aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name # Ensures lines connect properly
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Create a separate panel for each occupation
  # scales = "free_y" allows each plot to zoom in on its specific trend
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Apply distinct colors based on AI vulnerability
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      # Bold Orange
    "Not AI Vulnerable" = "#1f78b4",  # Cool Blue
    "Unclassified" = "grey70"         # Neutral Grey
  )) +
  
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    
    # Make the facet headers smaller and bold so long occupation names fit
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) # Tilts years to prevent overlap
  ) +
  
  scale_y_continuous(labels = label_comma()) +
  
  labs(
    title = "Top Occupations for Puerto Rican Female Workers Over Time",
    subtitle = "Estimated workforce volume by individual occupation",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```

### Continous visualization of Male and Female Top 10 Occupations

```{r Continuos top 10 occupations PR Males, message=FALSE, warning=FALSE}
# 1. Identify which occupations ever made the Top 10 for males
top_male_occs <- top_10_pr_occ_sex_labeled %>%
  filter(sex_label == "Male") %>%
  pull(OCC2010) %>%
  unique()

# 2. Pull the continuous history for ONLY those specific occupations
continuous_male_trends <- mapped_workers_db %>%
  filter(is_pr == TRUE, SEX == 1, OCC2010 %in% top_male_occs) %>%
  group_by(YEAR, OCC2010, ai_vuln) %>%
  summarise(total_estimated_workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  select(YEAR, OCC2010, occupation_name = lbl, ai_vuln, total_estimated_workers) %>%
   mutate(occupation_name = str_wrap(occupation_name, width = 35))

# 3. Plot the continuous data
ggplot(continuous_male_trends, aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name 
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      
    "Not AI Vulnerable" = "#1f78b4",  
    "Unclassified" = "grey70"         
  )) +
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) 
  ) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Top Occupations for Puerto Rican Male Workers Over Time",
    subtitle = "Continuous workforce volume for any occupation that ever reached the Top 10",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```

```{r Continuos top 10 occupations PR Males, message=FALSE, warning=FALSE}
# 1. Identify which occupations ever made the Top 10 for females
top_female_occs <- top_10_pr_occ_sex_labeled %>%
  filter(sex_label == "Female") %>%
  pull(OCC2010) %>%
  unique()

# 2. Pull the continuous history for ONLY those specific occupations
continuous_female_trends <- mapped_workers_db %>%
  # Note: SEX == 2 filters specifically for female workers in the database
  filter(is_pr == TRUE, SEX == 2, OCC2010 %in% top_female_occs) %>%
  group_by(YEAR, OCC2010, ai_vuln) %>%
  summarise(total_estimated_workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  select(YEAR, OCC2010, occupation_name = lbl, ai_vuln, total_estimated_workers) %>%
   mutate(occupation_name = str_wrap(occupation_name, width = 35))

# 3. Plot the continuous data for females
ggplot(continuous_female_trends, aes(
  x = YEAR, 
  y = total_estimated_workers, 
  color = ai_vuln,
  group = occupation_name 
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  scale_color_manual(values = c(
    "AI Vulnerable" = "#d95f02",      
    "Not AI Vulnerable" = "#1f78b4",  
    "Unclassified" = "grey70"         
  )) +
  
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1) 
  ) +
  
  scale_y_continuous(labels = scales::label_comma()) +
  
  labs(
    title = "Top Occupations for Puerto Rican Female Workers Over Time",
    subtitle = "Continuous workforce volume for any occupation that ever reached the Top 10",
    x = "Year",
    y = "Estimated Total Workers",
    color = "AI Vulnerability Risk"
  )
```

### Exposure over time (2001 to 2024)

```{r Calculate and visualize exposure, message=FALSE, warning=FALSE}

ai_exposure_data <- mapped_workers_db %>%
  filter(ai_vuln != "Unclassified") %>%
  group_by(YEAR, is_pr, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  group_by(YEAR, is_pr) %>%
  mutate(
    pct_workers = (workers / sum(workers)) * 100,
    demographic = ifelse(is_pr, "Puerto Rican Workers", "Total US Workers")
  ) %>%
  ungroup()
```

```{r }
ggplot(ai_exposure_data %>% filter(ai_vuln == "AI Vulnerable"), 
       aes(x = YEAR, y = pct_workers, color = demographic)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  my_theme +
  scale_color_manual(values = c("Puerto Rican Workers" = "#d95f02", "Total US Workers" = "#7570b3")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Workforce Exposure to AI Technologies (2001-2024)",
    subtitle = "Percentage of workers employed in AI-vulnerable occupations",
    x = "Year",
    y = "Share of Workforce (%)"
  ) +
  scale_x_continuous(breaks = seq(2001, 2024, by = 5))
```

### Wage Premium or Penalty for AI vulnerable occupations

```{r Wage preiums calculation and visualization, message=FALSE, warning=FALSE}
ai_wage_data <- mapped_workers_db %>%
  filter(ai_vuln != "Unclassified", INCWAGE > 0, INCWAGE < 999999) %>%
  group_by(YEAR, is_pr, ai_vuln) %>%
  summarise(
    avg_wage = sum(INCWAGE * CPI99 * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  collect() %>%
  mutate(demographic = ifelse(is_pr, "Puerto Rican Workers", "Total US Workers"))

ggplot(ai_wage_data, aes(x = YEAR, y = avg_wage, color = demographic, linetype = ai_vuln)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  my_theme +
  scale_color_manual(values = c("Puerto Rican Workers" = "#d95f02", "Total US Workers" = "#7570b3")) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Historical Wage Trajectories by AI Vulnerability",
    subtitle = "Real 1999 Dollars",
    x = "Year",
    y = "Average Annual Wage ($)"
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5))
```

### Demographic variations in AI vulnerable occupations

```{r Demographic difference calculation and visualization, message=FALSE, warning=FALSE}
ai_demographics <- mapped_workers_db %>%
  filter(is_pr == TRUE, ai_vuln != "Unclassified", SEX != 9) %>%
  mutate(sex_label = ifelse(SEX == 1, "Male", "Female")) %>%
  group_by(YEAR, sex_label, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  group_by(YEAR, sex_label) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100) %>%
  ungroup()

ggplot(ai_demographics %>% filter(ai_vuln == "AI Vulnerable"), 
       aes(x = YEAR, y = pct_workers, color = sex_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  my_theme +
  scale_color_manual(values = c("Male" = "#1f78b4", "Female" = "#e31a1c")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Puerto Rican AI Vulnerability by Sex",
    subtitle = "Percentage of each sex's workforce in AI-vulnerable occupations",
    x = "Year",
    y = "Exposure Rate (%)"
  )
```

### Demographic variations in AI vulnerable occupations for total population

```{r}
ai_demographics_us <- mapped_workers_db %>%
  filter(ai_vuln != "Unclassified", SEX != 9) %>%
  mutate(sex_label = ifelse(SEX == 1, "Male", "Female")) %>%
  group_by(YEAR, sex_label, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  group_by(YEAR, sex_label) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100) %>%
  ungroup()

ggplot(ai_demographics %>% filter(ai_vuln == "AI Vulnerable"), 
       aes(x = YEAR, y = pct_workers, color = sex_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  my_theme +
  scale_color_manual(values = c("Male" = "#1f78b4", "Female" = "#e31a1c")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "AI Vulnerability by Sex",
    subtitle = "Percentage of each sex's workforce in AI-vulnerable occupations",
    x = "Year",
    y = "Exposure Rate (%)"
  )
```

## AI Vulnerable Occupations Puerto Rican trends

```{r Calculate Puerto Ricans in AI Vulnerable Occ, message=FALSE,warning=FALSE}

ai_exposure_by_occ <- mapped_workers_db %>%
  filter(ai_vuln == "AI Vulnerable") %>%
  group_by(YEAR, is_pr, OCC2010) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val"))%>%
  group_by(YEAR,is_pr) %>%
  mutate(
    pct_workers = (workers / sum(workers)) * 100,
    demographic = ifelse(is_pr, "Puerto Rican Workers", "Total US Workers")
  ) %>%
  ungroup()

```

```{r}
# 1. Clean the labels before plotting
ai_exposure_plot_data <- ai_exposure_by_occ %>%
  # Rename 'lbl' (from the left_join) so it's easier to reference
  rename(occupation_name = lbl) %>%
  # Wrap long IPUMS occupation text so it stacks neatly in the facet headers
  mutate(occupation_name = str_wrap(occupation_name, width = 35))

# 2. Plot the comparison trends
ggplot(ai_exposure_plot_data, aes(
  x = YEAR, 
  y = pct_workers, 
  color = demographic,
  group = demographic
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Facet by occupation to give each job its own comparison panel
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Distinct colors for the demographics
  scale_color_manual(values = c(
    "Puerto Rican Workers" = "#d95f02", # Orange for emphasis
    "Total US Workers" = "#7570b3"      # Cooler purple/blue for baseline
  )) +
  
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(), # Hides "demographic" as it's self-explanatory
    legend.text = element_text(size = 10, face = "bold"),
    
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  ) +
  
  # Format Y-axis as percentages
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  
  labs(
    title = "Concentration of AI-Vulnerable Occupations",
    subtitle = "Comparing the distribution of exposed Puerto Rican workers vs. the Total US baseline",
    x = "Year",
    y = "Share of AI-Vulnerable Workforce (%)"
  )
```

### Average Wage by Occupaitons vulnerable to AI over time (2001-2024)

```{r Calculate Average Wage per Occupations, message=FALSE,warning=FALSE}
ai_wage_data_by_occ <- mapped_workers_db %>%
  filter(ai_vuln == "AI Vulnerable", INCWAGE > 0, INCWAGE < 999999) %>%
  group_by(YEAR, is_pr, OCC2010) %>%
  summarise(
    avg_wage = sum(INCWAGE * CPI99 * PERWT, na.rm = TRUE) / sum(PERWT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val"))%>%
  mutate(demographic = ifelse(is_pr, "Puerto Rican Workers", "Total US Workers"))
```

### Visualize Average Wage over time for exposed Occupations

```{r Visualize Average Wage for AI Vulnerable Occ, message=FALSE,warning=FALSE}

# 1. Clean the labels before plotting
ai_wage_data_by_occ <- ai_wage_data_by_occ %>%
  # Rename 'lbl' (from the left_join) so it's easier to reference
  rename(occupation_name = lbl) %>%
  # Wrap long IPUMS occupation text so it stacks neatly in the facet headers
  mutate(occupation_name = str_wrap(occupation_name, width = 35))

# 2. Plot the comparison trends
ggplot(ai_wage_data_by_occ, aes(
  x = YEAR, 
  y = avg_wage, 
  color = demographic,
  group = demographic
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Facet by occupation to give each job its own comparison panel
  facet_wrap(~ occupation_name, ncol = 4, scales = "free_y") +
  
  # Distinct colors for the demographics
  scale_color_manual(values = c(
    "Puerto Rican Workers" = "#d95f02", # Orange for emphasis
    "Total US Workers" = "#7570b3"      # Cooler purple/blue for baseline
  )) +
  
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(), # Hides "demographic" as it's self-explanatory
    legend.text = element_text(size = 10, face = "bold"),
    
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  ) +
  
  # Format Y-axis as percentages
  scale_y_continuous(labels = scales::dollar_format()) +
  
  labs(
    title = "Wage Trajectory for AI Vulnerable Occupatoins",
    subtitle = "Real 1999 Dollars",
    x = "Year",
    y = "Average Annual Wage ($)"
  )
```
