---
title: "02_emploment_populaiton_ratio"
author: "Jorge R. Soldevila Irizarry"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
selected_years <- c(2005, 2010, 2015, 2020)

# Source central setup file
source("00_project_setup.R")
```

# Puerto Rican Labor Force

## Estimate the Puerto Rican population in the Labor Force and other calculations

```{r Total Puerto Ricans in Labor Force, message=FALSE,warning=FALSE}

# We use the function in 00_project_setup to generate calculations for Employment-population ratio
# Puerto Ricans in the US
pr_lf_rate_us <- calculate_lfpr(data_acs, HISPAN == 2 | HISPAND == 200 | 
                                  ANCESTR1 == 261)
```

```{r Visualize Total Puerto Ricans in Labor Force, message=FALSE,warning=FALSE}

# We can now visualize the Puerto Rican population in the labor force since 2001.

ggplot() +
  geom_line(data = pr_lf_rate_us, aes(x = YEAR, y = etop_ratio, 
                                      color = "PR in US"), size = 1) +
  geom_point(data = pr_lf_rate_us, aes(x = YEAR, y = etop_ratio, 
                                       color = "PR in US")) +
  important_years_lines +
  my_theme +
  labs(
    title = "Trends in Employment-population ratio of Puerto Ricans 2001-2024",
    x = "Year",
    y = "Ratio"
  )
```

# Compare with Total Population and Other Groups

## Employment-population ratio for Total Population, Hispanics, Other Race Groups, and Hispanic Origin Groups

```{r Estimates for populations in Labor Force, message=FALSE,warning=FALSE}

# We can use the same function in 00_project_setup to obtain estimates for all
# other groups of interest

# 1. Total U.S. Population
lf_rate_us <- calculate_lfpr(data_acs)

# 2. Non-Hispanics in US
non_hisp_lf_rate_us <- calculate_lfpr(data_acs, HISPAN == 0)

# 3. Hispanics in US
hisp_lf_rate_us <- calculate_lfpr(data_acs, HISPAN > 0 & HISPAN < 9)

# 4. Non-Hispanic Race Groups in US
lf_by_race <- data_acs |>
  filter(
    HISPAN == 0,
    AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3,
    RACE < 8
  ) |> 
  mutate(race_cat = case_when(
    RACE == 1 ~ "01. White",
    RACE == 2 ~ "02. Black/African American",
    RACE == 3 ~ "03. AIAN",
    RACE >= 4 & RACE <= 6 ~ "04. Asian/PI",
    RACE == 7 ~ "05. Other",
    TRUE ~ NA
  )
         ) |> 
  filter(!is.na(race_cat)) |>
  group_by(YEAR, race_cat) |>
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp     = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    labfor_rate    = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate     = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor-unemp)/total_pop * 100
  ) |>
  collect()

  

# 5. Hispanic Origin Groups in US
lf_by_hisp_ori <- data_acs |>
  filter(
    HISPAN > 0 & HISPAN < 9,
    AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3
  ) |> 
  mutate(hisp_ori = case_when(
    HISPAND >= 100 & HISPAND <= 107 ~ "Mexican",
    HISPAND == 200 ~ "Puerto Rican",
    HISPAND == 300 ~ "Cuban",
    HISPAND >= 400 & HISPAND <= 417 ~ "Central American",
    HISPAND >= 420 & HISPAND <= 429 ~ "South American",
    HISPAND == 460 ~ "Dominican",
    TRUE ~ NA
  )
  ) |> 
  filter(!is.na(hisp_ori)) |>
  group_by(YEAR, hisp_ori) |>
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp     = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    labfor_rate    = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate     = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor-unemp)/total_pop * 100
  ) |>
  collect()

```

### We can now visualize Estimates for Labor Force for the total US population and other race and ethnicity groups

#### Total Population vs Hispanics vs Non-Hispanics

```{r Visualize Total US Emp-Pop Ratio with Hispanics and Non Hispanics, message=FALSE, warning=FALSE}

ggplot() +
  geom_line(data = lf_rate_us, 
            aes(x = YEAR, y = etop_ratio, color = "Total"), size = 1) +
  geom_point(data = lf_rate_us, 
             aes(x = YEAR, y = etop_ratio, color = "Total")) +
  
  geom_line(data = hisp_lf_rate_us, 
            aes(x = YEAR, y = etop_ratio, color = "Hispanics"), size = 1) +
  geom_point(data = hisp_lf_rate_us, 
             aes(x = YEAR, y = etop_ratio, color = "Hispanics")) +
  
  geom_line(data = non_hisp_lf_rate_us, 
            aes(x = YEAR, y = etop_ratio, color = "Non-Hispanics"), size = 1) +
  geom_point(data = non_hisp_lf_rate_us, 
             aes(x = YEAR, y = etop_ratio, color = "Non-Hispanics"), linetype = 2) +
  
  geom_line(data = pr_lf_rate_us, aes(x = YEAR, y = etop_ratio, 
                                      color = "Puerto Rican"), size = 1) +
  geom_point(data = pr_lf_rate_us, aes(x = YEAR, y = etop_ratio, 
                                       color = "Puerto Rican")) +
  
  important_years_lines +
  my_theme +
  scale_color_manual(values = c("Total" = "blue", 
                                "Hispanics" = "orange", 
                                "Non-Hispanics" = "darkgreen",
                                "Puerto Rican" = "red")) +
  labs(
    title = "Trends in Employment-populaiton ratio 2001-2024",
    x = "Year",
    y = "Ratio"
  )

```

#### Race Categories vs Puerto Ricans

```{r Visualize Labor FOrce for Race Categories, message=FALSE, warning=FALSE}

ggplot() +
  geom_line(data = lf_by_race, 
            aes(x = YEAR, y = etop_ratio, color = race_cat), 
            linewidth = 0.6, alpha = 0.4) +
  geom_point(data = lf_by_race, 
             aes(x = YEAR, y = etop_ratio, color = race_cat),
             size = 1.5, alpha = 0.4) +
  geom_line(data = pr_lf_rate_us, aes(x = YEAR, y = etop_ratio, 
                                      color = "Puerto Rican"), size = 1) +
  geom_point(data = pr_lf_rate_us, aes(x = YEAR, y = etop_ratio, 
                                       color = "Puerto Rican")) +

  important_years_lines +
  my_theme +
  # scale_color_manual(values = c("Total" = "blue", 
                                # "Hispanics" = "orange", 
                                # "Non-Hispanics" = "darkgreen")) +
  labs(
    title = "Trends in Employment-populaiton ratio 2001-2024",
    x = "Year",
    y = "Ratio"
  )

```

#### Hispanic Origin Groups

```{r Visualize Labor Force for Hispanic Origin, message=FALSE, warning=FALSE}

ggplot() +
  geom_line(data = lf_by_hisp_ori, 
            aes(x = YEAR, y = etop_ratio, color = hisp_ori), size = 1) +
  geom_point(data = lf_by_hisp_ori, 
             aes(x = YEAR, y = etop_ratio, color = hisp_ori)) +
  important_years_lines +
  my_theme +
  labs(
    title = "Trends in Employment-populaiton ratio 2001-2024",
    x = "Year",
    y = "Ratio"
  )
```

## Estimate Puerto Rican Employment-populaiton ratio by Sex

```{r Puerto Ricans Employment-population ratio by Sex}
pr_lf_by_sex <- data_acs %>%
    filter(HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261,
           AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3,
           SEX != 9) %>%
    group_by(YEAR,SEX) %>%
    summarise(
      total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
      in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
      out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
      unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      sex = case_when(SEX == 1 ~ "Male",
                      SEX == 2 ~ "Female"),
      labfor_rate = (in_labfor / total_pop) * 100,
      out_labfor_pct = (out_labfor / total_pop) * 100,
      unemp_rate = (unemp / in_labfor) * 100,
      etop_ratio = (in_labfor-unemp)/total_pop * 100
    ) %>%
    collect()
```

### Puerto Rican Employment-population Ratio by Sex

```{r Distribution of Puerto Ricans by sex, message=FALSE, warning=FALSE}
# Distribution of Puerto Ricans by sex
pr_lf_by_sex <- pr_lf_by_sex |>
  group_by(YEAR) |>
  mutate(pct = (total_pop/sum(total_pop)*100))%>%
  ungroup()

ggplot(pr_lf_by_sex, aes(x = YEAR, y = etop_ratio, fill = sex)) +
  geom_col(position = "dodge") +
  labs(title = "Population Percentage by Sex Over Time",
    x = "Year",
    y = "Percentage (%)",
    fill = "Sex"
  ) +
  # A clean, professional background theme
  my_theme
  
```

### Visualize trends by sex for Puerto Ricans

```{r}

# We can now visualize the Puerto Rican population in the labor force by sex since 2001.

ggplot() +
  geom_line(data = pr_lf_by_sex, aes(x = YEAR, y = etop_ratio, 
                                      color = sex)) +
  geom_point(data = pr_lf_by_sex, aes(x = YEAR, y = etop_ratio, 
                                      color = sex)) +
  important_years_lines +
  my_theme +
  labs(
    title = "Trends in Employment-population ratio by Sex 2001-2024",
    x = "Year",
    y = "Ratio"
  )
```

## Estimate Puerto Rican Labor Force by Place of Birth

```{r Puerto Ricans in Labor Force by Sex}
pr_lf_by_bpl <- data_acs %>%
    filter(HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261,
           AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3,
           SEX != 9) %>%
    mutate(
      plc_bth =  case_when(
        BPL <= 56 ~ "US",
        BPL == 110 ~ "Puerto Rico",
        TRUE ~ NA
      )) %>%
    filter(!is.na(plc_bth)) %>%
    group_by(YEAR,plc_bth) %>%
    summarise(
      total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
      in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
      out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
      unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      labfor_rate = (in_labfor / total_pop) * 100,
      out_labfor_pct = (out_labfor / total_pop) * 100,
      unemp_rate = (unemp / in_labfor) * 100,
      etop_ratio = (in_labfor - unemp) / total_pop * 100
    ) %>%
    collect()
```

### Visualize trends by place of birth for Puerto Ricans

```{r}

# We can now visualize the Puerto Rican population in the labor force by sex since 2001.

ggplot() +
  geom_line(data = pr_lf_by_bpl, aes(x = YEAR, y = etop_ratio, 
                                      color = plc_bth)) +
  geom_point(data = pr_lf_by_bpl, aes(x = YEAR, y = etop_ratio, 
                                      color = plc_bth)) +
  important_years_lines +
  my_theme +
  labs(
    title = "Trends in Employment-populaiton ratio by Place of Birth 2001-2024",
    x = "Year",
    y = "Ratio"
  )
```

<!-- ## Estimate Puerto Rican populaiton in Labor Force by State -->

<!-- ```{r} -->

<!-- pr_lf_by_state <- data_acs %>% -->

<!--   filter( -->

<!--     (HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261), -->

<!--     AGE >= 16,  -->

<!--     EMPSTAT != 0,  -->

<!--     !(EMPSTATD %in% 13:15),  -->

<!--     GQ != 3, -->

<!--     SEX %in% c(1, 2) -->

<!--   ) %>% -->

<!--   mutate( -->

<!--     sex = ifelse(SEX == 1, "Male", "Female") -->

<!--   ) %>% -->

<!--   filter(!is.na(STATEFIP)) %>% -->

<!--   group_by(YEAR, STATEFIP, sex) %>% -->

<!--   summarise( -->

<!--     total_pop  = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE), -->

<!--     in_labfor  = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE), -->

<!--     out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE), -->

<!--     unemp      = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE), -->

<!--     .groups    = "drop" -->

<!--   ) %>% -->

<!--   mutate( -->

<!--     labfor_rate    = ifelse(total_pop == 0, NA, (in_labfor / total_pop) * 100), -->

<!--     out_labfor_pct = ifelse(total_pop == 0, NA, (out_labfor / total_pop) * 100), -->

<!--     unemp_rate     = ifelse(in_labfor == 0, NA, (unemp / in_labfor) * 100) -->

<!--   ) %>% -->

<!--   collect() -->

<!-- ``` -->

<!-- ```{r} -->

<!-- pr_lf_by_state$st_name <- fips_info(pr_lf_by_state$STATEFIP)$full -->

<!-- ``` -->

<!-- ```{r} -->

<!-- ``` -->
