---
title: "02_LFPR_analysis"
author: "Jorge R. Soldevila Irizarry"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(scipen = 999)
selected_years <- c(2005, 2010, 2015, 2020)

# Source central setup file
source("00_project_setup.R")
```

## Overall Participation Rate

```{r Rates calculation, message =  FALSE, warning = FALSE}
# 1. Puerto Ricans in the US
pr_lf_rate_us <- calculate_lfpr(data_acs, HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261)

# 2. Total US Population
lf_rate_us <- calculate_lfpr(data_acs)

# 3. Puerto Rico Population
lf_rate_pr <- calculate_lfpr(data_prcs)
```

## LFPR Visualization

```{r Rates Visusalization, message = FALSE, warning=FALSE}
ggplot() +
  geom_line(data = pr_lf_rate_us, aes(x = YEAR, y = labfor_rate, color = "PR in US"), size = 1) +
  geom_point(data = pr_lf_rate_us, aes(x = YEAR, y = labfor_rate, color = "PR in US")) +
  
  geom_line(data = lf_rate_us, aes(x = YEAR, y = labfor_rate, color = "Total US"), size = 1) +
  geom_point(data = lf_rate_us, aes(x = YEAR, y = labfor_rate, color = "Total US")) +
  
  # geom_line(data = lf_rate_pr, aes(x = YEAR, y = labfor_rate, color = "PR in PR"), size = 1) +
  # geom_point(data = lf_rate_pr, aes(x = YEAR, y = labfor_rate, color = "PR in PR"), linetype = 2) +
  
  important_years_lines +
  my_theme +
  scale_color_manual(values = c("PR in US" = "blue", "Total US" = "orange", "PR in PR" = "lightblue")) +
  labs(
    title = "Trends in Labor Force Participation Rate 2001-2024",
    x = "Year",
    y = "Participation Rate (%)"
  )
```

# Compare with Total Population and Other Groups

## Calculate for Total Population, Hispanics, Other Race Groups, and Hispanic Origin Groups

```{r LFPR for population groups, message=FALSE,warning=FALSE}

# 1. Hispanics in the US
hisp_lf_rate_us <- calculate_lfpr(data_acs, HISPAN > 0 & HISPAN < 9)

# 2. Non-Hispanics in the US  
non_hisp_lf_rate_us <- calculate_lfpr(data_acs, HISPAN == 0)

# 3. Non-Hispanic Race Groups
lf_by_race <- data_acs |>
  filter(
    HISPAN == 0,
    AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3,
    RACE < 8
  ) |> 
  mutate(race_cat = case_when(
    RACE == 1 ~ "01. White",
    RACE == 2 ~ "02. Black/African American",
    RACE == 3 ~ "03. AIAN",
    RACE >= 4 & RACE <= 6 ~ "04. Asian/PI",
    RACE == 7 ~ "05. Other",
    TRUE ~ NA
  )
         ) |> 
  filter(!is.na(race_cat)) |>
  group_by(YEAR, race_cat) |>
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp     = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    labfor_rate    = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate     = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor-unemp)/total_pop * 100
  ) |>
  collect()

  

# 5. Hispanic Origin Groups in US
lf_by_hisp_ori <- data_acs |>
  filter(
    HISPAN > 0 & HISPAN < 9,
    AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3
  ) |> 
  mutate(hisp_ori = case_when(
    HISPAND >= 100 & HISPAND <= 107 ~ "Mexican",
    HISPAND == 200 ~ "Puerto Rican",
    HISPAND == 300 ~ "Cuban",
    HISPAND >= 400 & HISPAND <= 417 ~ "Central American",
    HISPAND >= 420 & HISPAND <= 429 ~ "South American",
    HISPAND == 460 ~ "Dominican",
    TRUE ~ NA
  )
  ) |> 
  filter(!is.na(hisp_ori)) |>
  group_by(YEAR, hisp_ori) |>
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp     = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    labfor_rate    = (in_labfor / total_pop) * 100,
    out_labfor_pct = (out_labfor / total_pop) * 100,
    unemp_rate     = (unemp / in_labfor) * 100,
    etop_ratio = (in_labfor-unemp)/total_pop * 100
  ) |>
  collect()
```

### We can now visualize the Labor Force Participation Rate for the total US population and other race and ethnicity groups.

#### Total Population vs Hispanics vs Non-Hispanics vs Puerto Ricans

```{r Visualize Total US labor Force with Hispanics and Non Hispanics, message=FALSE, warning=FALSE}

ggplot() +
  geom_line(data = lf_rate_us, 
            aes(x = YEAR, y = labfor_rate, color = "Total"), size = 1) +
  geom_point(data = lf_rate_us, 
             aes(x = YEAR, y = labfor_rate, color = "Total")) +
  
  geom_line(data = hisp_lf_rate_us, 
            aes(x = YEAR, y = labfor_rate, color = "Hispanics"), size = 1) +
  geom_point(data = hisp_lf_rate_us, 
             aes(x = YEAR, y = labfor_rate, color = "Hispanics")) +
  
  geom_line(data = non_hisp_lf_rate_us, 
            aes(x = YEAR, y = labfor_rate, color = "Non-Hispanics"), size = 1) +
  geom_point(data = non_hisp_lf_rate_us, 
             aes(x = YEAR, y = labfor_rate, color = "Non-Hispanics")) +
  
  geom_line(data = pr_lf_rate_us, 
            aes(x = YEAR, y = labfor_rate, color = "Puerto Rican"), size = 1) +
  geom_point(data = pr_lf_rate_us, 
             aes(x = YEAR, y = labfor_rate, color = "Puerto Rican")) +
  
  
  important_years_lines +
  my_theme +
  scale_color_manual(values = c("Total" = "blue", 
                                "Hispanics" = "orange", 
                                "Non-Hispanics" = "darkgreen",
                                "Puerto Rican" = "red")) +
  
  labs(
    title = "Trends in Labor Force Participation Rate 2001-2024",
    x = "Year",
    y = "Estimate"
  )

```

#### Race Categories vs Puerto Ricans

```{r Visualize Labor FOrce for Race Categories, message=FALSE, warning=FALSE}

ggplot() +
  geom_line(data = lf_by_race, 
            aes(x = YEAR, y = labfor_rate, color = race_cat), size = 1) +
  geom_point(data = lf_by_race, 
             aes(x = YEAR, y = labfor_rate, color = race_cat)) +
  
  geom_line(data = pr_lf_rate_us, aes(x = YEAR, y = labfor_rate, 
                                      color = "Puerto Rican"), size = 1) +
  geom_point(data = pr_lf_rate_us, aes(x = YEAR, y = labfor_rate, 
                                       color = "Puerto Rican")) +

  important_years_lines +
  my_theme +
  labs(
    title = "Trends in Labor Force Participation 2001-2024",
    x = "Year",
    y = "Estimate"
  )

```

#### Hispanic Origin Groups

```{r Visualize Labor Force for Hispanic Origin, message=FALSE, warning=FALSE}

ggplot() +
  geom_line(data = lf_by_hisp_ori, 
            aes(x = YEAR, y = labfor_rate, color = hisp_ori), size = 1) +
  geom_point(data = lf_by_hisp_ori, 
             aes(x = YEAR, y = labfor_rate, color = hisp_ori)) +
  important_years_lines +
  my_theme +
  labs(
    title = "Trends in Labor Force Participation Rate 2001-2024",
    x = "Year",
    y = "Ratio"
  )
```

## Estimate Puerto Rican Labor Force Participation Rate ratio by Sex

```{r Puerto Ricans LFPR by Sex, message=FALSE, warning=FALSE}
pr_lf_by_sex <- data_acs %>%
    filter(HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261,
           AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3,
           SEX != 9) %>%
    group_by(YEAR,SEX) %>%
    summarise(
      total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
      in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
      out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
      unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      sex = case_when(SEX == 1 ~ "Male",
                      SEX == 2 ~ "Female"),
      labfor_rate = (in_labfor / total_pop) * 100,
      out_labfor_pct = (out_labfor / total_pop) * 100,
      unemp_rate = (unemp / in_labfor) * 100,
      etop_ratio = (in_labfor-unemp)/total_pop * 100
    ) %>%
    collect()
```

### Visualize trends by sex for Puerto Ricans

```{r Visualize LFPR by Sex, message=FALSE, warning=FALSE}
# We can now visualize the Puerto Rican population in the labor force by sex since 2001.

ggplot() +
  geom_line(data = pr_lf_by_sex, aes(x = YEAR, y = labfor_rate, 
                                      color = sex)) +
  geom_point(data = pr_lf_by_sex, aes(x = YEAR, y = labfor_rate, 
                                      color = sex)) +
  important_years_lines +
  my_theme +
  labs(
    title = "Trends in Labor Force Participation Rate by Sex 2001-2024",
    x = "Year",
    y = "Ratio"
  )
```

## Estimate Puerto Rican Labor Force by Place of Birth

```{r Puerto Ricans in Labor Force by Sex}
pr_lf_by_bpl <- data_acs %>%
    filter(HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261,
           AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3,
           SEX != 9) %>%
    mutate(
      plc_bth =  case_when(
        BPL <= 56 ~ "US",
        BPL == 110 ~ "Puerto Rico",
        TRUE ~ NA
      )) %>%
    filter(!is.na(plc_bth)) %>%
    group_by(YEAR,plc_bth) %>%
    summarise(
      total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
      in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
      out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
      unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      labfor_rate = (in_labfor / total_pop) * 100,
      out_labfor_pct = (out_labfor / total_pop) * 100,
      unemp_rate = (unemp / in_labfor) * 100,
      etop_ratio = (in_labfor - unemp) / total_pop * 100
    ) %>%
    collect()
```

### Visualize trends by place of birth for Puerto Ricans

```{r}

# We can now visualize the Puerto Rican population in the labor force by sex since 2001.

ggplot() +
  geom_line(data = pr_lf_by_bpl, aes(x = YEAR, y = etop_ratio, 
                                      color = plc_bth)) +
  geom_point(data = pr_lf_by_bpl, aes(x = YEAR, y = etop_ratio, 
                                      color = plc_bth)) +
  important_years_lines +
  my_theme +
  labs(
    title = "Trends in Employment-populaiton ratio by Place of Birth 2001-2024",
    x = "Year",
    y = "Ratio"
  )
```

# LFPR by Age Groups

```{r LFPR Age GRoups, message=FALSE, warning=FALSE}
# Create a base table with the virtual grouping
acs_with_age <- data_acs %>%
  mutate(
    age_grp = case_when(
      AGE < 16 ~ "minor",
      AGE >= 16 & AGE <= 24 ~ "16 to 24",
      AGE >= 25 & AGE <= 54 ~ "25 to 54",
      AGE >= 55 ~ "55+"
    )
  )

# Calculate for PR in US by Age Group
pr_age_lf_rate_us <- acs_with_age %>%
  filter(HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261, AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3) %>%
  group_by(YEAR, age_grp) %>%
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    labfor_rate = in_labfor / total_pop * 100,
    unemp_rate = unemp / in_labfor * 100
  ) %>%
  collect()
```

## LFPR by Age Visualization

```{r LFPR Age Group Visualization}
ggplot(pr_age_lf_rate_us, aes(x = YEAR, y = labfor_rate, color = age_grp)) +
  geom_line(size = 1) +
  geom_point() +
  important_years_lines +
  my_theme +
  labs(
    title = "Labor Force Participation Rate for Puerto Ricans in the US by Age Group",
    x = "Year",
    y = "Participation Rate (%)"
  )
```
