---
title: "03_age_adjusted_LFPR"
author: "Jorge R. Soldevila Irizarry"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
selected_years <- c(2005, 2010, 2015, 2020)

# Source central setup file
source("00_project_setup.R")
```

# Age Adjusted Labor Force Participation Rate

## LFPR for Puerto Ricans in the US

```{r Total Puerto Ricans in Labor Force, message=FALSE,warning=FALSE}

# We use the function in 00_project_setup to calculate the total Puerto Rican population in the labor force. 

# Puerto Ricans in the US
pr_lf_rate_us <- calculate_age_adjusted_lfpr(data_hist, HISPAN == 2 | HISPAND == 200
                                             | ANCESTR1 == 261)
```

```{r Visualize Puerto Rican Labor Force Participation Rate, message=FALSE,warning=FALSE}

# We can now visualize the Puerto Rican Labor Force Participation Rate since 2001.

ggplot() +
  geom_line(data = pr_lf_rate_us, aes(x = YEAR, y = adjusted_lfpr, 
                                      color = "Adjusted"), size = 1) +
  geom_point(data = pr_lf_rate_us, aes(x = YEAR, y = adjusted_lfpr, 
                                       color = "Adjusted")) +
  geom_line(data = pr_lf_rate_us, aes(x = YEAR, y = crude_lfpr, 
                                      color = "Crude"), size = 1) +
  geom_point(data = pr_lf_rate_us, aes(x = YEAR, y = crude_lfpr, 
                                       color = "Crude")) +
  important_years_lines +
  my_theme +
  labs(
    title = "Crude vs Age-adjusted Labor Force Participation Rate 2001-2024",
    x = "Year",
    y = "Percent"
  )
```

### Change in age composition over time for Puerto Ricans in the US

```{r Calculate Change 2001-2024, message=FALSE, warning=FALSE}

age_dist_pr <- data_acs |>
  filter(AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3,
         HISPAN == 2 | HISPAND == 200 | ANCESTR1 == 261) |>
  mutate(age_grp = case_when(
    AGE >= 16 & AGE <= 24 ~ "16-24",
    AGE >= 25 & AGE <= 54 ~ "25-54",
    AGE >= 55 ~ "55+"
    )
  ) |>
  group_by(YEAR,age_grp) |>
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    labfor_rate = in_labfor / total_pop * 100,
    unemp_rate = unemp / in_labfor * 100
  ) %>%
  collect()

age_dist_pr <- age_dist_pr |>
  group_by(YEAR) |>
  mutate(
    pct_pop = total_pop/sum(total_pop),
    pct_labfor = in_labfor/sum(in_labfor)
  )

age_dist_us <- data_acs |>
  filter(AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3) |>
  mutate(age_grp = case_when(
    AGE >= 16 & AGE <= 24 ~ "16-24",
    AGE >= 25 & AGE <= 54 ~ "25-54",
    AGE >= 55 ~ "55+"
  )
  ) |>
  group_by(YEAR,age_grp) |>
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    unemp = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    labfor_rate = in_labfor / total_pop * 100,
    unemp_rate = unemp / in_labfor * 100
  ) %>%
  collect()

age_dist_us <- age_dist_us |>
  group_by(YEAR) |>
  mutate(
    pct_pop = total_pop/sum(total_pop),
    pct_labfor = in_labfor/sum(in_labfor)
  )
```

### Visualize change in distribution of Puerto Rican population vs Total US

```{r}
bind_rows(
  age_dist_pr |> mutate(population = "Puerto Ricans in U.S."),
  age_dist_us |> mutate(population = "Total U.S.")
) |>
  ggplot(aes(x = YEAR, y = pct_pop, color = age_grp, linetype = population, group = interaction(age_grp, population))) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  important_years_lines +
  my_theme +
  labs(
    title = "Age Distribution Over Time: Puerto Ricans vs. Total U.S.",
    subtitle = "Share of working-age population (16+) by age group, 2001–2024",
    x = "Year",
    y = "Share of Population",
    color = "Age Group",
    linetype = "Population"
  )
```

```{r}
bind_rows(
  age_dist_pr |> mutate(population = "Puerto Ricans in U.S."),
  age_dist_us |> mutate(population = "Total U.S.")
) |>
  ggplot(aes(x = YEAR, y = pct_labfor, color = age_grp, linetype = population, group = interaction(age_grp, population))) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  important_years_lines +
  my_theme +
  labs(
    title = "Age Distribution Over Time: Puerto Ricans vs. Total U.S.",
    subtitle = "Share of working-age population (16+) in labor force by age group, 2001–2024",
    x = "Year",
    y = "Share of Population",
    color = "Age Group",
    linetype = "Population"
  )
```

# Compare with Total Population and Other Groups

## Estimate for Total Population, Hispanics, Other Race Groups, and Hispanic Origin Groups

```{r Estimates for populations in Labor Force, message=FALSE,warning=FALSE}

# We can use the same function in 00_project_setup to obtain estimates for all
# other groups of interest

# 1. Total U.S. Population
lf_rate_us <- calculate_age_adjusted_lfpr(data_acs)

# 2. Non-Hispanics in US
  non_hisp_lf_rate_us <- calculate_age_adjusted_lfpr(data_acs, HISPAN == 0)

# 3. Hispanics in US
hisp_lf_rate_us <- calculate_age_adjusted_lfpr(data_acs, HISPAN > 0 & HISPAN < 9)

# 4. Non-Hispanic Race Groups in US
lf_by_race <- data_acs |>
  filter(
    HISPAN == 0,
    AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3,
    RACE < 8
  ) |> 
  mutate(race_cat = case_when(
    RACE == 1 ~ "01. White",
    RACE == 2 ~ "02. Black/African American",
    RACE == 3 ~ "03. AIAN",
    RACE >= 4 & RACE <= 6 ~ "04. Asian/PI",
    RACE == 7 ~ "05. Other",
    TRUE ~ NA
  ),
  age_bucket = case_when(
        AGE >= 16 & AGE <= 24 ~ "16-24",
        AGE >= 25 & AGE <= 54 ~ "25-54",
        AGE >= 55 ~ "55+"
      )
  ) |> 
  
  filter(!is.na(race_cat)) |>
  
  group_by(YEAR, age_bucket, race_cat) |>
  
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp     = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) |>
  
  collect() |>
  
  mutate(
    age_specific_lfpr = in_labfor / total_pop,
           age_specific_etop = (in_labfor - unemp) / total_pop
           ) %>%
    
    left_join(standard_weights, by = "age_bucket") %>%
    
    group_by(YEAR,race_cat) %>%
  
    summarise(
      crude_lfpr = (sum(in_labfor) / sum(total_pop)) * 100,
      crude_etop = (sum(in_labfor - unemp) / sum(total_pop)) * 100,
      crude_unemp = (sum(unemp) / sum(in_labfor)) * 100,
      
      adjusted_lfpr = sum(age_specific_lfpr * std_weight) * 100,
      adjusted_etop = sum(age_specific_etop * std_weight) * 100,
      .groups = "drop"
    )


  

# 5. Hispanic Origin Groups in US
lf_by_hisp_ori <- data_acs |>
  filter(
    HISPAN > 0 & HISPAN < 9,
    AGE >= 16, EMPSTAT != 0, !(EMPSTATD %in% 13:15), GQ != 3
  ) |> 
  mutate(hisp_ori = case_when(
    HISPAND >= 100 & HISPAND <= 107 ~ "Mexican",
    HISPAND == 200 ~ "Puerto Rican",
    HISPAND == 300 ~ "Cuban",
    HISPAND >= 400 & HISPAND <= 417 ~ "Central American",
    HISPAND >= 420 & HISPAND <= 429 ~ "South American",
    HISPAND == 460 ~ "Dominican",
    TRUE ~ NA
  ),
  age_bucket = case_when(
        AGE >= 16 & AGE <= 24 ~ "16-24",
        AGE >= 25 & AGE <= 54 ~ "25-54",
        AGE >= 55 ~ "55+"
      )
  ) |> 
  
  filter(!is.na(hisp_ori)) |>
  
  group_by(YEAR, age_bucket, hisp_ori) |>
  
  summarise(
    total_pop = sum(ifelse(EMPSTAT %in% c(1, 2, 3), PERWT, 0), na.rm = TRUE),
    in_labfor = sum(ifelse(EMPSTAT %in% c(1, 2), PERWT, 0), na.rm = TRUE),
    out_labfor = sum(ifelse(EMPSTAT == 3, PERWT, 0), na.rm = TRUE),
    unemp     = sum(ifelse(EMPSTAT == 2, PERWT, 0), na.rm = TRUE),
    .groups = "drop"
  ) |>
  
  collect() |>
  
  mutate(
    age_specific_lfpr = in_labfor / total_pop,
           age_specific_etop = (in_labfor - unemp) / total_pop
           ) %>%
    
    left_join(standard_weights, by = "age_bucket") %>%
    
    group_by(YEAR,hisp_ori) %>%
  
    summarise(
      crude_lfpr = (sum(in_labfor) / sum(total_pop)) * 100,
      crude_etop = (sum(in_labfor - unemp) / sum(total_pop)) * 100,
      crude_unemp = (sum(unemp) / sum(in_labfor)) * 100,
      
      adjusted_lfpr = sum(age_specific_lfpr * std_weight) * 100,
      adjusted_etop = sum(age_specific_etop * std_weight) * 100,
      .groups = "drop"
    )
```

### We can now visualize Estimates for Labor Force for the total US population and other race and ethnicity groups

#### Total Population vs Hispanics vs Non-Hispanics

```{r Visualize Total US labor Force with Hispanics and Non Hispanics, message=FALSE, warning=FALSE}

ggplot() +
  geom_line(data = lf_rate_us, 
            aes(x = YEAR, y = adjusted_lfpr, color = "Total"), size = 1) +
  geom_point(data = lf_rate_us, 
             aes(x = YEAR, y = adjusted_lfpr, color = "Total")) +
  
  geom_line(data = hisp_lf_rate_us, 
            aes(x = YEAR, y = adjusted_lfpr, color = "Hispanics"), size = 1) +
  geom_point(data = hisp_lf_rate_us, 
             aes(x = YEAR, y = adjusted_lfpr, color = "Hispanics")) +
  
  geom_line(data = non_hisp_lf_rate_us, 
            aes(x = YEAR, y = adjusted_lfpr, color = "Non-Hispanics"), size = 1) +
  geom_point(data = non_hisp_lf_rate_us, 
             aes(x = YEAR, y = adjusted_lfpr, color = "Non-Hispanics")) +
  geom_line(data = pr_lf_rate_us, 
            aes(x = YEAR, y = adjusted_lfpr, color = "Puerto Rican"), size = 1) +
  geom_point(data = pr_lf_rate_us, 
             aes(x = YEAR, y = adjusted_lfpr, color = "Puerto Rican")) +
  important_years_lines +
  my_theme +
  scale_color_manual(values = c("Total" = "blue", 
                                "Hispanics" = "orange", 
                                "Non-Hispanics" = "darkgreen",
                                "Puerto Rican" = "red")) +
  labs(
    title = "Labor Force Participation Rate 2001-2024",
    subtitle = "Puertoricans vs Total Population vs Hispanics vs Non Hispanics",
    x = "Year",
    y = "Estimate"
  )

```

#### Race Categories vs Puerto Ricans

```{r Visualize Labor FOrce for Race Categories, message=FALSE, warning=FALSE}

ggplot() +
  geom_line(data = lf_by_race, 
            aes(x = YEAR, y = adjusted_lfpr, color = race_cat), size = 1) +
  geom_point(data = lf_by_race, 
             aes(x = YEAR, y = adjusted_lfpr, color = race_cat)) +
  
  geom_line(data = pr_lf_rate_us, aes(x = YEAR, y = adjusted_lfpr, 
                                      color = "Puerto Rican"), size = 1) +
  geom_point(data = pr_lf_rate_us, aes(x = YEAR, y = adjusted_lfpr, 
                                       color = "Puerto Rican")) +

  important_years_lines +
  my_theme +
  # scale_color_manual(values = c("Total" = "blue", 
                                # "Hispanics" = "orange", 
                                # "Non-Hispanics" = "darkgreen")) +
  labs(
    title = "Trends in Labor Force Participation 2001-2024",
    x = "Year",
    y = "Estimate"
  )

```

#### Hispanic Origin Groups

```{r Visualize Labor Force for Hispanic Origin, message=FALSE, warning=FALSE}

ggplot() +
  geom_line(data = lf_by_hisp_ori, 
            aes(x = YEAR, y = adjusted_lfpr, color = hisp_ori), size = 1) +
  geom_point(data = lf_by_hisp_ori, 
             aes(x = YEAR, y = adjusted_lfpr, color = hisp_ori)) +
  important_years_lines +
  my_theme +
  # scale_color_manual(values = c("Total" = "blue", 
  # "Hispanics" = "orange", 
  # "Non-Hispanics" = "darkgreen")) +
  labs(
    title = "Trends in Labor Force Participation 2001-2024",
    x = "Year",
    y = "Estimate"
  )
```

## Adding Some stuff here

```{r Age adjusted calculation}
pr_lf_rate_us <- calculate_age_adjusted_lfpr(data_acs, HISPAN == 2 |
  HISPAND == 200 |
  ANCESTR1 == 261)
lf_rate_us <- calculate_age_adjusted_lfpr(data_acs)
```

### Visualize Age Adjusted labor force participation rate

```{r Age adjusted LFPR visualization}
p <- ggplot() +
  geom_line(
    data = pr_lf_rate_us,
    aes(x = YEAR, y = adjusted_lfpr, color = "Puerto Rican"),
    linewidth = 1.5
  ) +
  geom_point(
    data = pr_lf_rate_us,
    aes(x = YEAR, y = adjusted_lfpr, color = "Puerto Rican"),
    size = 3 
  )  +
  geom_line(
    data = lf_rate_us,
    aes(x = YEAR, y = adjusted_lfpr, color = "Total US"),
    linewidth = 1.5
  ) +
  geom_point(
    data = lf_rate_us,
    aes(x = YEAR, y = adjusted_lfpr, color = "Total US"),
    size = 3
  ) +
  my_theme +
  # Expanded labels for clarity
  scale_color_manual(values = c("Puerto Rican" = "#d95f02", "Total US" = "#7570b3")) +
  # Appends a % sign to the y-axis values
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Age-Adjusted Labor Force Participation Rates and Unemployment Rate",
    subtitle = "Comparing Puerto Rican workers to the total US population (2001-2024)",
    x = NULL, # 'Year' is obvious, so we remove the label to save space
    y = "Participation Rate",
    caption = "Source: IPUMS USA, ACS (2001-2024)"
  ) +
  # Overriding my_theme defaults to make text presentation-sized
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14, color = "grey30"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.text = element_text(size = 14),
    plot.caption = element_text(size = 10, color = "grey50", hjust = 1)
  ) +
  important_years_lines 

print(p)

# ggsave(
#   filename = "figures/age_adj_pr_lfpr_presentation.png",
#   plot = p,
#   width = 10, # Width in inches (standard for 16:9 slides)
#   height = 5.625, # Height in inches (10 / (16/9) = 5.625)
#   dpi = 300, # 300 DPI ensures it won't look pixelated on a projector
#   bg = "white" # Ensures the background is solid white, not transparent
# )
```
