---
title: "Update_AI_vulnerable_occ"
author: "Jorge R. Soldevila Irizarry"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 999)
selected_years <- c(2005, 2010, 2015, 2020)

# Source central setup file
source("00_project_setup.R")

ddi <- read_ipums_ddi("./usa_00121.xml")
occ_labels <- ipums_val_labels(ddi, OCC2010) %>%
  mutate(lbl = case_when(val == 1020 ~ "Software Developer",
                         val == 430 ~ "Miscellaneous Managers",
                         TRUE ~ lbl))
```

# AI Exposed Occupations Analysis

## Map industry codes and AI vulnerable Occupations

```{r Mapping Industry and Occupaiton Categories, message=FALSE,warning=FALSE}
mapped_workers_db <- data_acs %>%
  # Apply base employment filter, correcting the missing GQ != 3
  filter(AGE >= 16, EMPSTAT == 1, GQ != 3) %>%
  mutate(
    # Create a boolean flag for PR status to split the data later
    is_pr = case_when(HISPAN == 2 |
                        HISPAND == 200 | ANCESTR1 == 261 ~ TRUE, TRUE ~ FALSE),
    
    # Map Industries
    ind_sector = case_when(
      IND1990 >= 10 & IND1990 <= 32 ~ "Ag, Forestry, Fish & Hunt",
      IND1990 >= 40 & IND1990 <= 50 ~ "Mining, Quarry, Oil/Gas",
      IND1990 >= 450 & IND1990 <= 472 ~ "Utilities",
      IND1990 == 60 ~ "Construction",
      IND1990 >= 100 & IND1990 <= 392 ~ "Manufacturing",
      IND1990 >= 500 & IND1990 <= 571 ~ "Wholesale Trade",
      (IND1990 >= 580 &
         IND1990 <= 691) & IND1990 != 641 ~ "Retail Trade",
      IND1990 >= 400 &
        IND1990 <= 432 ~ "Transport & Warehousing",
      IND1990 >= 440 &
        IND1990 <= 442 ~ "Information",
      # Maps 1990 Communications here
      IND1990 >= 700 & IND1990 <= 711 ~ "Finance & Insurance",
      IND1990 == 712 ~ "Real Estate & Leasing",
      IND1990 == 841 |
        (IND1990 >= 881 & IND1990 <= 893) ~ "Prof, Sci, & Tech Svcs",
      IND1990 >= 721 &
        IND1990 <= 741 ~ "Admin, Support, & Waste Mgmt",
      IND1990 >= 842 & IND1990 <= 860 ~ "Educational Svcs",
      (IND1990 >= 812 &
         IND1990 <= 840) |
        (IND1990 >= 861 & IND1990 <= 873) ~ "Health & Social Asst",
      IND1990 >= 800 &
        IND1990 <= 810 ~ "Arts, Entertain, & Rec",
      IND1990 %in% c(641, 762, 770) ~ "Accom & Food Svcs",
      IND1990 %in% c(750, 751, 752, 761, 771, 772, 
                     781, 791) ~ "Other Svcs (excl. Pub Admin)",
      IND1990 >= 900 & IND1990 <= 932 ~ "Public Admin",
      IND1990 >= 940 & IND1990 <= 960 ~ "Military",
      TRUE ~ "Unclassified / NILF"
    ),
    
    # Map Occupations
    occ_cat1 = case_when(
      (OCC2010 >= 10 &
         OCC2010 <= 960) |
        (OCC2010 >= 2600 &
           OCC2010 <= 2960) ~ "Mgmt, Bus, Entertain & Media",
      OCC2010 >= 1000 & OCC2010 <= 1965 ~ "STEM",
      (OCC2010 >= 2000 &
         OCC2010 <= 2550) |
        (OCC2010 >= 4300 &
           OCC2010 <= 4650) ~ "Educ, Comm & Social Svcs",
      OCC2010 >= 3000 & OCC2010 <= 3655 ~ "Health & Medical",
      (OCC2010 >= 3700 &
         OCC2010 <= 3955) |
        (OCC2010 >= 9000 &
           OCC2010 <= 9420) ~ "Law Enf & Transport",
      OCC2010 >= 4000 &
        OCC2010 <= 4250 ~ "Food, Maint, Bldg & Grounds",
      OCC2010 >= 4700 & OCC2010 <= 4965 ~ "Sales & Related",
      OCC2010 >= 5000 &
        OCC2010 <= 5940 ~ "Admin & Office Support",
      OCC2010 >= 6005 &
        OCC2010 <= 7630 ~ "Nat Res, Const & Maint",
      (OCC2010 >= 7700 &
         OCC2010 <= 8965) |
        (OCC2010 >= 9500 &
           OCC2010 <= 9750) ~ "Prod & Material Moving",
      OCC2010 >= 9800 & OCC2010 <= 9830 ~ "Military",
      OCC2010 %in% c(9920, 9999) ~ "Unclassified",
      TRUE ~ "Unclassified"
    ), 
    # Map AI vulnerable occupaitons 
ai_vuln = case_when(
  OCC2010 %in% c(10,20,110,120,150,160,430,540, 565, 630, 650, 700, 735,740,
                 800, 910,1005, 1006, 1007, 1010, 1020, 1050,1105, 1106, 1400,
                 1430, 1460, 1530,1965, 2825,3510, 3655,4830, 4850, 
                 4940, 2145, 5100, 5110, 5120, 5160, 5240, 5400,
                 5520, 5600, 5700, 5840, 5860) ~ "AI Vulnerable",
  OCC2010 %in% c(9920, 9999) ~ "Unclassified",
  TRUE ~ "Not AI Vulnerable"
  ),
  ai_vul_grp = case_when(
    OCC2010 %in% c(10,20,110,120,150,160,430) ~ "Management",
    OCC2010 %in% c(540, 565, 630, 650, 700, 735,740) ~ "Business Operations",
    OCC2010 %in% c(800, 910) ~ "Financial Specialists",
    OCC2010 %in% c(1005, 1006, 1007, 1010, 1020, 1050,
                   1105, 1106, 1400, 1430, 1460, 1530,
                   1965) ~ "STEM",
    OCC2010 %in% c(2825) ~ "Entertainment and Media",
    OCC2010 %in% c(3510, 3655) ~ "Health Care",
    OCC2010 %in% c(4830, 4850, 4940) ~ "Sales and Related",
    OCC2010 %in% c(2145, 5100, 5110, 5120, 5160, 5240, 5400,
                   5520, 5600, 5700, 5840, 5860) ~ "Admin & Office Support",
    TRUE ~ "Not AI Vulnerable"
  )

)
```

## Exposure over time (2001 to 2024)

```{r Calculate exposure 2001-2024, message=FALSE, warning=FALSE}

ai_exposure_pr <- mapped_workers_db %>%
  filter(is_pr == TRUE, ai_vuln != "Unclassified") %>%
  group_by(YEAR, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  group_by(YEAR) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100, 
         demographic = "Puerto Rican Workers") %>%
  ungroup()

ai_exposure_us <- mapped_workers_db %>%
  filter(ai_vuln != "Unclassified") %>%
  group_by(YEAR, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  group_by(YEAR) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100, 
         demographic = "Total US Workers") %>%
  ungroup()

ai_exposure_data <- bind_rows(ai_exposure_pr, ai_exposure_us)
```

```{r Visualize exposure, message=FALSE, warning=FALSE}
p <- ggplot(ai_exposure_data %>% filter(ai_vuln == "AI Vulnerable"), 
       aes(x = YEAR, y = pct_workers, color = demographic)) +
  geom_line(linewidth = 1.5) +  
  geom_point(size = 3) +    
  geom_text(data = ai_exposure_data %>% 
      filter(
        ai_vuln == "AI Vulnerable", 
        # Combine the 5-year sequence with the year 2024 manually
        YEAR %in% c(seq(2001, 2024, by = 5), 2024) 
      ),
    aes(
      label = paste0(round(pct_workers, 1), "%"),
      # Conditional vjust: Positive pushes down, negative pushes up
      vjust = ifelse(demographic == "Puerto Rican Workers", 1.8, -1.2)
    ),
    size = 3.5,      
    fontface = "bold",
    show.legend = FALSE
    ) +
  my_theme +
  scale_color_manual(values = c("Puerto Rican Workers" = "#d95f02", "Total US Workers" = "#7570b3")) +
  scale_y_continuous(
    limits = c(0, 35),
    breaks = seq(0,100, by = 5),
    labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(2001, 2024, by = 5)) +
  labs(
    title = "Workforce Exposure to AI Technologies", 
    subtitle = "Percentage of workers employed in AI-vulnerable occupations (2001-2024)", 
    x = NULL, 
    y = "Share of Workforce", 
    caption = "Source: IPUMS USA, ACS (2001-2024)" 
  )

print(p)

ggsave(
  filename = "figures/ai_exposure_presentation.png",
  plot = p,
  width = 10, # Width in inches (standard for 16:9 slides)
  height = 5.625, # Height in inches (10 / (16/9) = 5.625)
  dpi = 300, # 300 DPI ensures it won't look pixelated on a projector
  bg = "white" # Ensures the background is solid white, not transparent
)
```

## AI Vulnerable Occupation Groups trends

```{r Calculate Puerto Ricans in AI Vulnerable Occ, message=FALSE,warning=FALSE}
ai_occ_grp_exposure_pr <- mapped_workers_db %>%
  filter(is_pr == TRUE, ai_vul_grp != "Not AI Vulnerable") %>%
  group_by(YEAR, ai_vul_grp) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  #left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  group_by(YEAR) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100, demographic = "Puerto Rican Workers") %>%
  ungroup()

ai_occ_grp_exposure_us <- mapped_workers_db %>%
  filter(ai_vul_grp != "Not AI Vulnerable") %>%
  group_by(YEAR, ai_vul_grp) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  #left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  group_by(YEAR) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100, demographic = "Total US Workers") %>%
  ungroup()

ai_exposure_by_occ_grp <- bind_rows(ai_occ_grp_exposure_pr, ai_occ_grp_exposure_us)

```

## Share of AI Workers per Occupation Groups

```{r Share of AI Workers by Occupation, message=FALSE, warning=FALSE}
# 2. Plot the comparison trends
ggplot(ai_exposure_by_occ_grp, aes(
  x = YEAR, 
  y = pct_workers, 
  color = demographic,
  group = demographic
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Facet by occupation to give each job its own comparison panel
  facet_wrap(~ ai_vul_grp, ncol = 4, scales = "free_y") +
  
  # Distinct colors for the demographics
  scale_color_manual(values = c(
    "Puerto Rican Workers" = "#d95f02", # Orange for emphasis
    "Total US Workers" = "#7570b3"      # Cooler purple/blue for baseline
  )) +
  
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(), 
    legend.text = element_text(size = 10, face = "bold"),
    
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  ) +
  
  # Format Y-axis as percentages
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  
  labs(
    title = "Concentration of AI-Vulnerable Occupations",
    subtitle = "Comparing the distribution of exposed Puerto Rican workers vs. the Total US baseline",
    x = "Year",
    y = "Share of AI-Vulnerable Workforce (%)"
  )
```

## Share of workers per AI Vulnerable Occupations - 3-yr Rolling Average

```{r Share of AI Workers by Occupation Rolling Avg, message=FALSE, warning=FALSE}

# 1. Calculate the 3-year rolling average
smoothed_data <- ai_exposure_by_occ_grp %>%
  # Crucial: Ensure the data is sorted by year before calculating the average
  arrange(demographic, ai_vul_grp, YEAR) %>%
  
  # Group by both demographic and your facet category so the math doesn't bleed across groups
  group_by(demographic, ai_vul_grp) %>%
  
  mutate(
    # k = 3 sets the 3-year window
    # fill = NA ensures the first and last years are handled correctly without shifting the data
    # align = "center" places the calculated average exactly on the middle year
    pct_workers_smoothed = zoo::rollmean(pct_workers, k = 3, fill = NA, align = "center")
  ) %>%
  ungroup()

# 2. Plot the smoothed data
ggplot(smoothed_data, aes(x = YEAR, color = demographic)) +
  
  # The Background: Raw data plotted as thin, transparent lines
  geom_line(aes(y = pct_workers), alpha = 0.3, linewidth = 0.5) +
  
  # The Foreground: Smoothed rolling average plotted as thick, bold lines
  geom_line(aes(y = pct_workers_smoothed), linewidth = 1.2) +
  
  # Facet by your occupation categories (matching your image)
  facet_wrap(~ai_vul_grp, scales = "free_y") +
  
  # Aesthetics and styling
  scale_color_manual(values = c("Puerto Rican Workers" = "#d95f02", "Total US Workers" = "#7570b3")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "gray90", color = NA) # Styles the facet headers
  ) +
  
  # Labels
  labs(
    title = "Concentration of AI-Vulnerable Occupations",
    subtitle = "Comparing a 3-Year Rolling Average (raw annual data shown as faded lines)",
    x = "Year",
    y = "Share of AI-Vulnerable Workforce (%)",
    color = NULL
  )
```

## AI Vulnerable Occupations by Occupation Group

```{r Calculate Puerto Ricans in AI Vulnerable Occ, message=FALSE,warning=FALSE}

ai_occ_exposure_pr <- mapped_workers_db %>%
  filter(is_pr == TRUE, ai_vuln == "AI Vulnerable") %>%
  group_by(YEAR, ai_vul_grp, OCC2010) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  group_by(YEAR) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100, demographic = "Puerto Rican Workers") %>%
  ungroup()

ai_occ_exposure_us <- mapped_workers_db %>%
  filter(ai_vuln == "AI Vulnerable") %>%
  group_by(YEAR, ai_vul_grp, OCC2010) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  group_by(YEAR) %>%
  mutate(pct_workers = (workers / sum(workers)) * 100, demographic = "Total US Workers") %>%
  ungroup()

ai_exposure_by_occ <- bind_rows(ai_occ_exposure_pr, ai_occ_exposure_us)

```

## Occupations vulnerable to AI by Occupation Groups

```{r Calculate Occupations by Occ Group, message=FALSE,warning=FALSE}
# 1. Get a list of all unique AI vulnerable groups
ai_groups <- unique(ai_exposure_by_occ$ai_vul_grp)

# 2. Loop through each group
for (group_name in ai_groups) {
  
  plot_data <- ai_occ_exposure_pr %>% 
    filter(ai_vul_grp == group_name)
  
  label_data <- plot_data %>% 
    filter(YEAR == max(YEAR))
  
  # 3. Build the plot (Notice 'color = lbl' added here)
  p <- ggplot(plot_data, aes(x = YEAR, y = pct_workers, group = lbl, color = lbl)) +
    
    # Removed the hardcoded color so it uses the distinct colors generated by ggplot
    geom_line(linewidth = 0.8, alpha = 0.8) +
    
    # Add labels at the end. They will automatically inherit the line's color!
    geom_text(
      data = label_data,
      aes(label = str_wrap(lbl, 25)), 
      hjust = 0,                      
      nudge_x = 0.5,                  
      size = 3,
      lineheight = 0.8,
      show.legend = FALSE # Prevents 'a' from showing up in legends if you re-enable them
    ) +
    
    facet_wrap(~ demographic) +
    
    scale_x_continuous(
      breaks = seq(2001, 2024, by = 5),
      expand = expansion(mult = c(0.05, 0.6)) 
    ) +
    
    scale_y_continuous(labels = function(x) paste0(x, "%")) +
    
    theme_bw() +
    theme(
      legend.position = "none", # Keeps the side legend hidden since we have direct labels
      strip.background = element_rect(fill = "gray90", color = NA),
      strip.text = element_text(face = "bold"),
      plot.title = element_text(face = "bold")
    ) +
    
    labs(
      title = paste("Occupation-Level Exposure:", group_name),
      x = "Year",
      y = "Share of AI-Vulnerable Workforce (%)"
    )
  
  # 4. Print the plot
  print(p)
}
```

## Puerto Rican concentration by specific occupation

```{r Top 10 occupations for PR, message = FALSE, warning = FALSE}
top_10_pr <- mapped_workers_db %>%
  filter(is_pr == TRUE, YEAR == 2024, ai_vuln == "AI Vulnerable") %>%
  group_by(OCC2010) %>%
  summarise(total_workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  mutate(
    pct_workers = (total_workers / sum(total_workers)) * 100, 
    demographic = "Puerto Rican Workers"
  ) %>%
  slice_max(order_by = total_workers, n = 10) %>%
  left_join(occ_labels, by = c("OCC2010" = "val"))

# 2. Get the Total US Baseline for ONLY those 10 occupations
top_10_us <- mapped_workers_db %>%
  filter(YEAR == 2024, ai_vuln == "AI Vulnerable") %>%
  group_by(OCC2010) %>%
  summarise(total_workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  mutate(
    pct_workers = (total_workers / sum(total_workers)) * 100, 
    demographic = "Total US Workers"
  ) %>%
  # NEW: Filter using the exact occupation codes from the PR dataset
  filter(OCC2010 %in% top_10_pr$OCC2010) %>%
  left_join(occ_labels, by = c("OCC2010" = "val"))

# 3. Combine the data and lock in the sorting order
# We want the bars sorted by the Puerto Rican highest values, not alphabetically
ordered_labels <- top_10_pr %>% 
  arrange(pct_workers) %>% 
  pull(lbl)

plot_data <- bind_rows(top_10_pr, top_10_us) %>%
  mutate(lbl = factor(lbl, levels = ordered_labels))

# 4. Plot the horizontal bar chart
p_compare <- ggplot(plot_data, aes(x = pct_workers, y = lbl, fill = demographic)) +
  
  # position = "dodge" places the PR and US bars side-by-side rather than stacking them
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  
  # Add text labels to the end of the bars
  geom_text(
    aes(label = paste0(round(pct_workers, 1), "%")), 
    position = position_dodge(width = 0.8), 
    hjust = -0.15,
    size = 3.5
  ) +
  
  # Colors to match your existing presentation theme
  scale_fill_manual(values = c("Puerto Rican Workers" = "#d95f02", "Total US Workers" = "#7570b3")) +
  
  # Format X-axis and expand it so labels aren't cut off
  scale_x_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.15))
  ) +
  
  my_theme +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.major.y = element_blank(), # Removes horizontal lines for a cleaner look
    axis.text.y = element_text(size = 10)
  ) +
  
  labs(
    title = "Top 10 AI-Vulnerable Occupations for Puerto Rican Workers",
    subtitle = "Comparing the share of AI-vulnerable workers in 2024 against the Total US baseline",
    x = "Share of AI-Vulnerable Workforce (%)",
    y = NULL,
    caption = "Source: IPUMS USA, ACS (2024)"
  )

print(p_compare)

```

## Demographic variations in AI vulnerable occupations

```{r Demographic difference calculation and visualization, message=FALSE, warning=FALSE}
# Puerto Rican Data
ai_demographics_pr <- mapped_workers_db %>%
  filter(is_pr == TRUE, ai_vuln != "Unclassified", SEX != 9) %>%
  mutate(sex_label = ifelse(SEX == 1, "Male", "Female")) %>%
  group_by(YEAR, sex_label, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  group_by(YEAR, sex_label) %>%
  mutate(
    pct_workers = (workers / sum(workers)) * 100,
    demographic = "Puerto Rican Workers" # Added identifier
  ) %>%
  ungroup()

# Total US Data
ai_demographics_us <- mapped_workers_db %>%
  filter(ai_vuln != "Unclassified", SEX != 9) %>%
  mutate(sex_label = ifelse(SEX == 1, "Male", "Female")) %>%
  group_by(YEAR, sex_label, ai_vuln) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  group_by(YEAR, sex_label) %>%
  mutate(
    pct_workers = (workers / sum(workers)) * 100,
    demographic = "Total US Workers" # Added identifier
  ) %>%
  ungroup()

# Combine datasets
ai_demographics <- bind_rows(ai_demographics_pr, ai_demographics_us)

```

### Visualize trends among men and women

```{r}
p <- ggplot(
  ai_demographics %>% filter(ai_vuln == "AI Vulnerable"), 
  aes(
    x = YEAR, 
    y = pct_workers, 
    color = demographic, 
    linetype = sex_label, 
    shape = sex_label # Makes the points distinct too!
  )
) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  
  my_theme +
  
  # Assign colors to the demographics
  scale_color_manual(values = c(
    "Puerto Rican Workers" = "#d95f02", 
    "Total US Workers" = "#7570b3"
  )) +
  
  # Format axes
  scale_y_continuous(
    limits = c(0, 35),
    breaks = seq(0,100, by = 5),
    labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(1950, 2024, by = 10)) +
  
  labs(
    title = "AI Vulnerability by Sex: Puerto Rican vs. Total US Workers",
    subtitle = "Percentage of each sex's workforce in AI-vulnerable occupations",
    x = NULL, 
    y = "Exposure Rate",
    color = "Demographic", # Titles for the legends
    linetype = "Sex",
    shape = "Sex",
    caption = "Source: IPUMS USA, ACS (2001-2024)"
  )

print(p)

# ggsave(
#   filename = "figures/ai_exposure_gender_presentation.png",
#   plot = p,
#   width = 10, # Width in inches (standard for 16:9 slides)
#   height = 5.625, # Height in inches (10 / (16/9) = 5.625)
#   dpi = 300, # 300 DPI ensures it won't look pixelated on a projector
#   bg = "white" # Ensures the background is solid white, not transparent
# )
# 
```
## Top 10 industry by Sex

```{r}
# 1. Filter and calculate the gender distribution for the Top 10 PR occupations
pr_gender_dist <- mapped_workers_db %>%
  # Filter for PR workers in 2024, exposed to AI, and remove unknown gender (if any)
  filter(
    is_pr == TRUE, 
    YEAR == 2024, 
    ai_vuln == "AI Vulnerable",
    SEX != 9, 
    OCC2010 %in% top_10_pr$OCC2010 # Limits data to ONLY the top 10 jobs from the previous step
  ) %>%
  
  # Create a readable label for gender
  mutate(sex_label = ifelse(SEX == 1, "Male", "Female")) %>%
  
  # Group by the occupation and gender to get the worker counts
  group_by(OCC2010, sex_label) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  
  # Join with occupation labels
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  
  # Calculate the percentage breakdown WITHIN each occupation
  group_by(lbl) %>%
  mutate(pct_gender = (workers / sum(workers)) * 100) %>%
  ungroup() %>%
  
  # Lock in the same sorting order used in your previous graph
  mutate(lbl = factor(lbl, levels = ordered_labels))

# 2. Plot the stacked bar chart
p_gender <- ggplot(pr_gender_dist, aes(x = pct_gender, y = lbl, fill = sex_label)) +
  
  # Use position = "stack" to make each bar add up to 100%
  geom_col(position = "stack", width = 0.7) +
  
  # Add percentage text labels inside the bars
  geom_text(
    aes(label = paste0(round(pct_gender, 1), "%")), 
    position = position_stack(vjust = 0.5), # Centers the text inside each colored block
    color = "white", 
    size = 3.5,
    fontface = "bold"
  ) +
  
  # Distinct colors for Male and Female
  scale_fill_manual(values = c("Male" = "#1b9e77", "Female" = "#e7298a")) +
  
  # Format X-axis
  scale_x_continuous(
    labels = function(x) paste0(x, "%"),
    expand = c(0, 0) # Removes the blank space at the edges of the 100% bars
  ) +
  
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.major.y = element_blank(), 
    axis.text.y = element_text(size = 10)
  ) +
  
  labs(
    title = "Gender Distribution in Top 10 AI-Vulnerable Occupations",
    subtitle = "Male vs. Female share within each occupation for Puerto Rican Workers (2024)",
    x = "Share of Workers within Occupation (%)",
    y = NULL,
    caption = "Source: IPUMS USA, ACS (2024)"
  )

print(p_gender)
```


## Comparing Occupations by Industry

```{r Industry by Occupations, message=FALSE, warning=FALSE}
#We want to know in what industries are vulnerable occupations associated with.
ai_occ_by_industry_pr <- mapped_workers_db %>%
  # 1. Filter for PR workers and explicitly isolate AI Vulnerable occupations
  filter(is_pr == TRUE, ai_vuln == "AI Vulnerable") %>%
  
  # 2. Group by Year, Industry, and specific Occupation Code
  group_by(YEAR, IND1990, OCC2010) %>%
  summarise(workers = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
  collect() %>%
  
  # 3. Join with your label dataset so you can read the actual job titles
  left_join(occ_labels, by = c("OCC2010" = "val")) %>%
  
  # 4. Group by Year AND Industry to calculate the percentage
  group_by(YEAR, IND1990) %>%
  mutate(
    # This now means: "What % of this industry's vulnerable workforce does this job make up?"
    pct_workers = (workers / sum(workers)) * 100, 
    demographic = "Puerto Rican Workers"
  ) %>%
  ungroup()
```


## Share of Workers per Occupation 2001 - 2024

```{r Share of AI Workers by Occupation, message=FALSE, warning=FALSE}
# 2. Plot the comparison trends
ggplot(ai_exposure_by_occ_grp, aes(
  x = YEAR, 
  y = pct_workers, 
  color = demographic,
  group = demographic
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  # Facet by occupation to give each job its own comparison panel
  facet_wrap(~ ai_vul_grp, ncol = 4, scales = "free_y") +
  
  # Distinct colors for the demographics
  scale_color_manual(values = c(
    "Puerto Rican Workers" = "#d95f02", # Orange for emphasis
    "Total US Workers" = "#7570b3"      # Cooler purple/blue for baseline
  )) +
  
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(), 
    legend.text = element_text(size = 10, face = "bold"),
    
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  ) +
  
  # Format Y-axis as percentages
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  
  labs(
    title = "Concentration of AI-Vulnerable Occupations",
    subtitle = "Comparing the distribution of exposed Puerto Rican workers vs. the Total US baseline",
    x = "Year",
    y = "Share of AI-Vulnerable Workforce (%)"
  )
```


